---
title: "CART From Scratch: Decision Trees That Think in If-Then Rules"
subtitle: "The most interpretable model in machine learning"
engine: knitr
---

## 1. The Problem: Explain Your Decision

You've learned models that are powerful but hard to explain:

- SVM: "the hyperplane with maximum margin decided..." (glazed eyes)
- KNN: "the 5 nearest neighbors voted..." (what neighbors?)
- Regression: "the weighted sum of 12 coefficients..." (too many numbers)

Sometimes you need a model that a **non-technical person can read and follow**.
A loan officer. A doctor. A manager who asks "why did the model say no?"

CART gives you **if-then rules** that anyone can understand:

```
IF income > $50K AND credit_score > 700
  → APPROVE
ELSE IF income > $50K AND credit_score ≤ 700
  → REVIEW
ELSE
  → DENY
```

That's a decision tree.


## 2. How CART Works: Splitting the Data

CART builds a tree by **repeatedly splitting** the data on the feature that
best separates the groups.

```{r}
#| label: fig-tree-visual
#| fig-cap: "CART splits the space into rectangles using if-then rules"

library(ggplot2)
library(rpart)
library(rpart.plot)

set.seed(42)
n <- 200

income <- runif(n, 20, 100)
credit <- runif(n, 400, 850)

# Decision rule: approve if income > 50 AND credit > 650 (with noise)
prob_approve <- plogis(-5 + 0.06 * income + 0.005 * credit)
approved <- rbinom(n, 1, prob_approve)

df <- data.frame(
  income = income,
  credit = credit,
  decision = factor(approved, labels = c("Deny", "Approve"))
)

# Fit a tree
tree <- rpart(decision ~ income + credit, data = df, method = "class",
              control = rpart.control(maxdepth = 3, cp = 0.02))

rpart.plot(tree, type = 4, extra = 101, under = TRUE,
  main = "Decision Tree: Loan Approval",
  box.palette = c("coral", "steelblue"))
```

```{r}
#| label: fig-tree-regions
#| fig-cap: "Each split creates a rectangular region — the tree carves up the feature space"

# Show the rectangular regions
grid <- expand.grid(
  income = seq(20, 100, length.out = 200),
  credit = seq(400, 850, length.out = 200)
)
grid$pred <- predict(tree, grid, type = "class")

ggplot() +
  geom_tile(data = grid, aes(income, credit, fill = pred), alpha = 0.3) +
  geom_point(data = df, aes(income, credit, color = decision), size = 1.5) +
  scale_fill_manual(values = c("coral", "steelblue")) +
  scale_color_manual(values = c("coral", "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(x = "Income ($K)", y = "Credit Score",
       fill = "Tree Region", color = "Actual",
       title = "CART splits the space into rectangular regions")
```

Compare this to SVM (which draws a single line/curve) or KNN (which draws
irregular boundaries). CART always makes **axis-aligned rectangular splits**.


## 3. The Splitting Rule: How CART Chooses Where to Cut

At each step, CART asks: "which feature and which split point best separates
the classes?"

### For Classification: Gini Impurity

$$\text{Gini} = 1 - \sum_{k=1}^{K} p_k^2$$

where $p_k$ is the proportion of class $k$ in the node.

| Gini Value | Meaning |
|:----------:|:--------|
| 0 | **Pure** — all points belong to one class |
| 0.5 | **Maximum impurity** — 50/50 split (for 2 classes) |

CART picks the split that creates the **lowest weighted average Gini** across
the two child nodes.

```{r}
#| label: fig-gini
#| fig-cap: "Gini impurity: 0 = pure node, 0.5 = maximum mixed"

p <- seq(0, 1, length.out = 100)
gini <- 2 * p * (1 - p)  # simplified for 2 classes: 1 - p² - (1-p)²

plot(p, gini, type = "l", lwd = 2, col = "steelblue",
  main = "Gini Impurity for Binary Classification",
  xlab = "Proportion of Class A (p)", ylab = "Gini Impurity")
abline(h = 0.5, lty = 2, col = "coral")
text(0.15, 0.45, "Maximum impurity\n(50/50 mix)", col = "coral", cex = 0.8)
points(0, 0, pch = 19, cex = 2, col = "forestgreen")
points(1, 0, pch = 19, cex = 2, col = "forestgreen")
text(0.08, 0.05, "Pure\n(all B)", col = "forestgreen", cex = 0.8)
text(0.92, 0.05, "Pure\n(all A)", col = "forestgreen", cex = 0.8)
```

```{r}
#| label: fig-split-demo
#| fig-cap: "CART evaluates every possible split and picks the one with lowest total Gini"

# Simple example: split on income at different points
split_points <- seq(30, 90, by = 5)
gini_scores <- sapply(split_points, function(sp) {
  left <- df$decision[df$income <= sp]
  right <- df$decision[df$income > sp]

  gini_left <- 1 - sum((table(left) / length(left))^2)
  gini_right <- 1 - sum((table(right) / length(right))^2)

  # Weighted average
  n_left <- length(left)
  n_right <- length(right)
  (n_left * gini_left + n_right * gini_right) / (n_left + n_right)
})

best_split <- split_points[which.min(gini_scores)]

plot(split_points, gini_scores, type = "b", pch = 19, col = "steelblue", lwd = 2,
  main = "Finding the Best Split on Income",
  xlab = "Split Point (income)", ylab = "Weighted Gini (lower = better)")
abline(v = best_split, col = "coral", lty = 2, lwd = 2)
text(best_split + 5, max(gini_scores),
  paste0("Best split: income = ", best_split, "K"),
  col = "coral", cex = 0.9)
```

### For Regression Trees: Sum of Squared Errors

When predicting a **number** instead of a class, CART minimizes SSE at each
split — same objective as linear regression, but applied locally in each
rectangle.

$$\text{SSE} = \sum_{i \in \text{left}} (y_i - \bar{y}_{\text{left}})^2 + \sum_{i \in \text{right}} (y_i - \bar{y}_{\text{right}})^2$$


## 4. The Tree Grows: Recursive Splitting

CART doesn't stop at one split. It takes each child node and splits it again,
then splits those children, and so on — **recursively**.

```{r}
#| label: fig-tree-growth
#| fig-cap: "Deeper trees create more splits — from simple to complex"

par(mfrow = c(1, 3), mar = c(2, 2, 3, 1))

for (depth in c(1, 3, 6)) {
  tree_d <- rpart(decision ~ income + credit, data = df, method = "class",
    control = rpart.control(maxdepth = depth, cp = 0.001, minsplit = 5))

  grid$pred <- predict(tree_d, grid, type = "class")

  plot(df$income, df$credit,
    col = ifelse(df$decision == "Approve", "steelblue", "coral"),
    pch = 19, cex = 0.8,
    main = paste0("Depth = ", depth, " (", nrow(tree_d$frame), " nodes)"),
    xlab = "Income", ylab = "Credit")

  points(grid$income, grid$credit,
    col = ifelse(grid$pred == "Approve",
      adjustcolor("steelblue", 0.15),
      adjustcolor("coral", 0.15)),
    pch = 15, cex = 0.3)
}
par(mfrow = c(1, 1))
```

A deeper tree creates more rectangles — fitting the training data more
closely. But there's a problem...


## 5. The Overfitting Trap: Deep Trees Memorize

::: {.callout-warning}
### Deep Trees Memorize
A tree with enough depth can perfectly classify every training point. But
it's just **memorizing** — it won't work on new data.
:::

```{r}
#| label: fig-overfit-tree
#| fig-cap: "Deep tree: perfect on training data, terrible on new data"

set.seed(42)

# Split into train/test
train_idx <- sample(n, round(0.7 * n))
train <- df[train_idx, ]
test <- df[-train_idx, ]

# Shallow tree
tree_shallow <- rpart(decision ~ income + credit, data = train, method = "class",
  control = rpart.control(maxdepth = 2, cp = 0.01))

# Deep tree (overfit)
tree_deep <- rpart(decision ~ income + credit, data = train, method = "class",
  control = rpart.control(maxdepth = 10, cp = 0.001, minsplit = 2))

# Accuracies
train_acc_shallow <- mean(predict(tree_shallow, train, type = "class") == train$decision)
test_acc_shallow <- mean(predict(tree_shallow, test, type = "class") == test$decision)
train_acc_deep <- mean(predict(tree_deep, train, type = "class") == train$decision)
test_acc_deep <- mean(predict(tree_deep, test, type = "class") == test$decision)

results <- data.frame(
  Model = rep(c("Shallow (depth=2)", "Deep (depth=10)"), each = 2),
  Set = rep(c("Train", "Test"), 2),
  Accuracy = c(train_acc_shallow, test_acc_shallow, train_acc_deep, test_acc_deep)
)

ggplot(results, aes(Model, Accuracy * 100, fill = Set)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("steelblue", "coral")) +
  geom_text(aes(label = paste0(round(Accuracy * 100, 1), "%")),
            position = position_dodge(0.6), vjust = -0.5, size = 3.5) +
  ylim(0, 105) +
  theme_minimal(base_size = 14) +
  labs(y = "Accuracy (%)", title = "Deep tree: high training, low test = OVERFIT")
```

**Signs of an overfit tree:**

- High training accuracy, low test accuracy
- Many levels / many leaves
- Some leaves have very few data points
- Complex rules that don't make intuitive sense


## 6. The Fix: Pruning

**Pruning** cuts back branches that don't improve generalization. There are
two strategies:

| Strategy | How it works |
|:---------|:-------------|
| **Pre-pruning** | Stop growing early (limit depth, minimum samples per leaf) |
| **Post-pruning** | Grow the full tree, then cut back branches that don't help on validation data |

```{r}
#| label: fig-pruning
#| fig-cap: "Pruning: grow a full tree, then trim branches that don't help"

# Grow a full tree
tree_full <- rpart(decision ~ income + credit, data = train, method = "class",
  control = rpart.control(cp = 0.001, minsplit = 2))

# Find best CP via cross-validation
printcp(tree_full)

# Plot CP table
plotcp(tree_full, main = "Cross-Validation Error vs Complexity (cp)")
```

```{r}
#| label: fig-pruned-tree
#| fig-cap: "The pruned tree: simpler, more generalizable"

# Prune to optimal cp
best_cp <- tree_full$cptable[which.min(tree_full$cptable[, "xerror"]), "CP"]
tree_pruned <- prune(tree_full, cp = best_cp)

par(mfrow = c(1, 2), mar = c(2, 2, 3, 1))

rpart.plot(tree_full, type = 4, extra = 101, under = TRUE,
  main = paste0("Full Tree (", sum(tree_full$frame$var == "<leaf>"), " leaves)"),
  box.palette = c("coral", "steelblue"), cex = 0.6)

rpart.plot(tree_pruned, type = 4, extra = 101, under = TRUE,
  main = paste0("Pruned Tree (", sum(tree_pruned$frame$var == "<leaf>"), " leaves)"),
  box.palette = c("coral", "steelblue"), cex = 0.7)

par(mfrow = c(1, 1))

# Compare test accuracy
acc_full <- mean(predict(tree_full, test, type = "class") == test$decision)
acc_pruned <- mean(predict(tree_pruned, test, type = "class") == test$decision)
cat(sprintf("Full tree test accuracy:   %.1f%%\n", acc_full * 100))
cat(sprintf("Pruned tree test accuracy: %.1f%%\n", acc_pruned * 100))
cat("Simpler tree often performs just as well (or better) on new data!")
```


## 7. CART for Regression

CART isn't just for classification. It can predict **continuous values** too.
Instead of majority vote in each leaf, it predicts the **mean** of the
training points in that leaf.

```{r}
#| label: fig-regression-tree
#| fig-cap: "Regression tree: predicts the average value in each rectangular region"

set.seed(42)
x_reg <- runif(150, 0, 10)
y_reg <- sin(x_reg) * 3 + rnorm(150, 0, 0.5)

df_reg <- data.frame(x = x_reg, y = y_reg)

# Fit regression trees of different depths
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

for (depth in c(2, 4, 8)) {
  tree_reg <- rpart(y ~ x, data = df_reg,
    control = rpart.control(maxdepth = depth, cp = 0.001))

  x_grid <- data.frame(x = seq(0, 10, length.out = 500))
  x_grid$pred <- predict(tree_reg, x_grid)

  plot(df_reg$x, df_reg$y, pch = 19, cex = 0.6, col = "gray40",
    main = paste("Depth =", depth),
    xlab = "x", ylab = "y")
  lines(x_grid$x, x_grid$pred, col = "steelblue", lwd = 2)
  lines(x_grid$x, sin(x_grid$x) * 3, col = "coral", lty = 2)
}
par(mfrow = c(1, 1))
```

Notice the **staircase** shape — regression trees make piecewise-constant
predictions (flat steps). They approximate curves by breaking them into
enough small rectangles.


## 8. CART vs Other Models

| | CART | SVM | KNN | Logistic Reg |
|:--|:-----|:----|:----|:-------------|
| **Interpretable?** | Yes — if-then rules | Moderate | No | Moderate (coefficients) |
| **Handles non-linear?** | Yes (inherently) | Yes (kernels) | Yes (inherently) | Only with transforms |
| **Handles mixed types?** | Yes (numeric + categorical) | Numeric only | Numeric only | Numeric only |
| **Needs scaling?** | **No** | Yes | Yes | Recommended |
| **Overfitting risk** | High (deep trees) | Moderate | High (small k) | Low |
| **Feature selection?** | Built in (top splits = most important) | No | No | Via p-values |

**CART's unique advantages:**

- **No scaling needed** — splits are based on thresholds, not distances
- **Handles mixed data** — categorical and numeric features together
- **Built-in feature importance** — features used in top splits matter most
- **Human-readable** — the only model a non-technical stakeholder can follow

```{r}
#| label: fig-variable-importance
#| fig-cap: "CART naturally tells you which features are most important"

# Fit a tree with more features
set.seed(42)
df_imp <- data.frame(
  income = income,
  credit = credit,
  age = runif(n, 22, 70),
  noise1 = rnorm(n),
  noise2 = runif(n),
  decision = df$decision
)

tree_imp <- rpart(decision ~ ., data = df_imp, method = "class",
  control = rpart.control(maxdepth = 4, cp = 0.01))

importance <- tree_imp$variable.importance
barplot(sort(importance, decreasing = TRUE),
  col = "steelblue", border = NA, las = 2,
  main = "Variable Importance (from CART)",
  ylab = "Importance Score")
```


## 9. When to Use CART (Practice Scenarios)

| Scenario | Use CART? | Why |
|:---------|:---------|:----|
| "Need interpretable rules for a loan officer" | **Yes** | CART = human-readable if-then rules |
| "Explain to management why the model rejected an applicant" | **Yes** | Trace the tree path |
| "Classify with maximum accuracy on large dataset" | Maybe not | SVM often more accurate |
| "Features are a mix of numbers and categories" | **Yes** | CART handles both natively |
| "Predict a probability" | Logistic regression | CART outputs class, not calibrated probability |
| "Non-linear relationships, need simple explanation" | **Yes** | Splits naturally capture non-linearity |

**Connection to advanced models:**

- **Random Forest** = many CART trees averaged together (reduces overfitting)
- **Gradient Boosting** = sequential CART trees correcting each other's errors
- These are the most powerful ML models in practice, and they're all built from CART


## 10. Cheat Sheet: The Whole Story on One Page

```
CART (Decision Tree) RECIPE
=============================

1. ALGORITHM:
   - Pick the feature + split point with lowest Gini impurity (classification)
     or lowest SSE (regression)
   - Split the data into two groups
   - Repeat recursively on each group
   - Stop when a stopping criterion is met (depth, min samples, etc.)

2. SPLITTING CRITERIA:
   Classification: Gini = 1 - Σpₖ²    (0 = pure, 0.5 = max impurity)
   Regression:     SSE = Σ(yᵢ - ȳ)²   (minimize within each split)

3. OVERFITTING:
   Deep trees memorize → prune!
   Pre-pruning: limit depth, min samples per leaf
   Post-pruning: grow full tree, cut back branches (use cp / cross-validation)

4. UNIQUE ADVANTAGES:
   - Human-readable (if-then rules)
   - No scaling required
   - Handles mixed data (numeric + categorical)
   - Built-in feature importance
   - Non-linear without transforms

5. WEAKNESSES:
   - High overfitting risk without pruning
   - Axis-aligned splits only (rectangular boundaries)
   - Unstable: small data changes → very different tree

6. PRACTICE TIPS:
   "Interpretable model" → CART
   "Explain the decision" → CART
   "Perfect training accuracy" → overfit, prune it
   Building block for Random Forest and Boosting
```


## 11. Check Your Understanding

::: {.callout-note}
### Test Yourself

Before moving on, try to answer these without scrolling up:

1. Why is CART called the most "interpretable" model?
2. What is Gini impurity? What value means a pure node? A maximally mixed node?
3. How does CART decide where to split?
4. A decision tree gets 100% accuracy on training data. Is this good?
   What should you do?
5. What is pruning? Why is it necessary?
6. Name two advantages CART has over SVM.
7. Name two advantages SVM has over CART.
8. A manager says "I need to explain to the board why we denied this loan."
   Which model would you recommend? Why?
9. How does a regression tree differ from a classification tree?
10. What are Random Forests and Gradient Boosting, and how do they relate
    to CART?
:::
