---
title: "Confusion Matrix Cost Drill"
subtitle: "Compute actual costs - don't guess | Binary choice practice"
engine: knitr
---

## Instructions

A practice prompt gives you confusion matrices and a cost structure, then asks
which model or threshold is cheaper. Plug in the actual numbers and compute
before you choose.

**Format:** IS / IS NOT fill-in-blank. Click the collapsed callout to reveal
each answer after you've committed.

**Key formula:** Total Cost = (# False Positives × Cost_FP) + (# False Negatives × Cost_FN)

---

## Scenario 1: Medical Screening

A hospital builds a classifier to detect a rare disease in blood samples.

**Confusion Matrix (Model A, threshold = 0.5):**

|  | Predicted Positive | Predicted Negative |
|--|-------------------|-------------------|
| Actually Positive | 42 (TP) | 8 (FN) |
| Actually Negative | 35 (FP) | 115 (TN) |

**Confusion Matrix (Model B, threshold = 0.3):**

|  | Predicted Positive | Predicted Negative |
|--|-------------------|-------------------|
| Actually Positive | 48 (TP) | 2 (FN) |
| Actually Negative | 62 (FP) | 88 (TN) |

**Costs:** Missing a disease case (FN) costs \$50,000 in delayed treatment. A false alarm (FP) costs \$500 in unnecessary follow-up testing.

### Q1

Model A's total cost \_\_\_\_\_ (**IS / IS NOT**) lower than Model B's total cost.

::: {.callout-note collapse="true" title="Answer"}
**IS NOT.**

```{r}
#| code-fold: true
#| code-summary: "Show computation"
cost_FP <- 500
cost_FN <- 50000

# Model A
A_FP <- 35; A_FN <- 8
cost_A <- A_FP * cost_FP + A_FN * cost_FN
cat("Model A:", A_FP, "FP ×", cost_FP, "+", A_FN, "FN ×", cost_FN, "=", cost_A, "\n")

# Model B
B_FP <- 62; B_FN <- 2
cost_B <- B_FP * cost_FP + B_FN * cost_FN
cat("Model B:", B_FP, "FP ×", cost_FP, "+", B_FN, "FN ×", cost_FN, "=", cost_B, "\n")

cat("\nModel A costs $", format(cost_A, big.mark=","),
    "\nModel B costs $", format(cost_B, big.mark=","),
    "\nModel B is cheaper by $", format(cost_A - cost_B, big.mark=","))
```

Model A: $417,500. Model B: $131,000. Model B is far cheaper despite having nearly twice as many false positives, because each missed disease case is 100× more expensive than a false alarm.
:::

---

### Q2

Model A \_\_\_\_\_ (**DOES / DOES NOT**) have higher overall accuracy than Model B.

::: {.callout-note collapse="true" title="Answer"}
**DOES.**

```{r}
#| code-fold: true
#| code-summary: "Show computation"
acc_A <- (42 + 115) / 200
acc_B <- (48 + 88) / 200
cat("Model A accuracy:", acc_A, "\nModel B accuracy:", acc_B)
```

Model A: 78.5% accuracy. Model B: 68.0% accuracy. Model A is more accurate but MORE EXPENSIVE. This is the key lesson: accuracy ≠ cost-effectiveness when costs are asymmetric.
:::

---

### Q3

If both error types cost the same (\$500 each), the cheaper model \_\_\_\_\_ (**WOULD / WOULD NOT**) change.

::: {.callout-note collapse="true" title="Answer"}
**WOULD.**

```{r}
#| code-fold: true
#| code-summary: "Show computation"
cost_sym <- 500
cost_A_sym <- 35 * cost_sym + 8 * cost_sym
cost_B_sym <- 62 * cost_sym + 2 * cost_sym
cat("Symmetric costs:\n")
cat("Model A:", 35, "FP +", 8, "FN =", 43, "errors × $500 = $", format(cost_A_sym, big.mark=","), "\n")
cat("Model B:", 62, "FP +", 2, "FN =", 64, "errors × $500 = $", format(cost_B_sym, big.mark=","), "\n")
cat("\nWith symmetric costs, Model A is cheaper (fewer total errors)")
```

With symmetric costs, Model A ($21,500) beats Model B ($32,000). The cost structure determines the optimal model.
:::

---

## Scenario 2: Three Threshold Comparison

A retailer predicts which products will sell out. Three probability thresholds:

**p = 0.3:**

|  | Predicted: Restock | Predicted: Skip |
|--|--------------------|----------------|
| Actually sold out | 88 | 12 |
| Actually didn't | 55 | 45 |

**p = 0.5:**

|  | Predicted: Restock | Predicted: Skip |
|--|--------------------|----------------|
| Actually sold out | 71 | 29 |
| Actually didn't | 22 | 78 |

**p = 0.7:**

|  | Predicted: Restock | Predicted: Skip |
|--|--------------------|----------------|
| Actually sold out | 45 | 55 |
| Actually didn't | 5 | 95 |

**Costs:** Unnecessary restock (FP) = \$100. Missed sellout (FN) = \$400 in lost revenue.

### Q4

The threshold with the lowest total cost \_\_\_\_\_ (**IS / IS NOT**) $p = 0.5$.

::: {.callout-note collapse="true" title="Answer"}
**IS NOT.** Compute the actual costs:

```{r}
#| code-fold: true
#| code-summary: "Show computation"
cost_FP <- 100; cost_FN <- 400

scenarios <- data.frame(
  threshold = c(0.3, 0.5, 0.7),
  FP = c(55, 22, 5),
  FN = c(12, 29, 55)
)
scenarios$total_cost <- scenarios$FP * cost_FP + scenarios$FN * cost_FN
scenarios$cost_label <- paste0("$", format(scenarios$total_cost, big.mark=","))

print(scenarios)
cat("\np = 0.3 has the lowest total cost at $", format(scenarios$total_cost[1], big.mark=","))
```

- p=0.3: 55×\$100 + 12×\$400 = \$5,500 + \$4,800 = **\$10,300**
- p=0.5: 22×\$100 + 29×\$400 = \$2,200 + \$11,600 = **\$13,800**
- p=0.7: 5×\$100 + 55×\$400 = \$500 + \$22,000 = **\$22,500**

$p = 0.3$ is cheapest at \$10,300 — not p=0.5. The lower threshold has more unnecessary restocks (55 vs 22), but each missed sellout costs 4× as much (\$400 vs \$100), so catching sellouts aggressively wins. Don't assume the middle threshold is always best — compute.
:::

---

### Q5

If the cost of a missed sellout DOUBLED to \$800, the optimal threshold \_\_\_\_\_ (**WOULD / WOULD NOT**) likely shift toward a LOWER threshold.

::: {.callout-note collapse="true" title="Answer"}
**WOULD.**

```{r}
#| code-fold: true
#| code-summary: "Show computation"
cost_FP <- 100; cost_FN <- 800

cat("With FN cost = $800:\n")
cat("p=0.3:", 55*cost_FP + 12*cost_FN, "\n")
cat("p=0.5:", 22*cost_FP + 29*cost_FN, "\n")
cat("p=0.7:", 5*cost_FP + 55*cost_FN, "\n")
```

Higher FN cost → want to avoid missing sellouts → predict "restock" more aggressively → lower threshold. p=0.3 becomes even more dominant.
:::

---

## Scenario 3: Build Your Own

Given any confusion matrix and cost structure, the computation is always:

$$\text{Total Cost} = (\text{FP} \times C_{FP}) + (\text{FN} \times C_{FN})$$

**Practice habit:** When you see a confusion matrix, immediately:

1. Identify FP and FN counts
2. Multiply each by its cost
3. Sum for total cost
4. Compare across options

Never skip the computation. The habit is simple: compute first, then choose.
