---
title: "Cross-Validation From Scratch: Why Testing on Training Data Is Cheating"
subtitle: "The one technique every model depends on"
engine: knitr
---

## 1. The Problem: How Good Is Your Model, Really?

You build a model. It predicts your training data perfectly. Is it good?

**No.** It might have just **memorized** the data. A model that memorizes is
useless on new data — like a student who memorizes answers but can't solve
new problems.

```{r}
#| label: fig-overfitting-demo
#| fig-cap: "The wiggly model fits training data perfectly but is clearly wrong"

library(ggplot2)
set.seed(42)

# True relationship: simple curve
x <- sort(runif(20, 0, 10))
y <- 2 * sin(x) + rnorm(20, 0, 0.8)
df <- data.frame(x = x, y = y)

# Overfit model: high-degree polynomial
fit_over <- lm(y ~ poly(x, 15), data = df)
# Good model: low-degree polynomial
fit_good <- lm(y ~ poly(x, 3), data = df)

x_grid <- seq(0, 10, length.out = 200)
pred_over <- predict(fit_over, newdata = data.frame(x = x_grid))
pred_good <- predict(fit_good, newdata = data.frame(x = x_grid))

ggplot(df, aes(x, y)) +
  geom_point(size = 3) +
  geom_line(data = data.frame(x = x_grid, y = pred_over),
            aes(x, y), color = "coral", linewidth = 1) +
  geom_line(data = data.frame(x = x_grid, y = pred_good),
            aes(x, y), color = "steelblue", linewidth = 1) +
  geom_line(data = data.frame(x = x_grid, y = 2 * sin(x_grid)),
            aes(x, y), color = "gray50", linetype = "dashed") +
  annotate("text", x = 8.5, y = 4, label = "Overfit (degree 15)",
           color = "coral", size = 4) +
  annotate("text", x = 8.5, y = -1, label = "Good fit (degree 3)",
           color = "steelblue", size = 4) +
  annotate("text", x = 1.5, y = -2.5, label = "True function",
           color = "gray50", size = 3.5) +
  theme_minimal(base_size = 14) +
  labs(x = "x", y = "y", title = "Training accuracy is misleading")
```

The red line hits every point — **100% training accuracy**. But it's wildly
wrong between the points. The blue line misses some points but captures the
real pattern.

**How do we tell the difference?** We need data the model has **never seen**.


## 2. The Simplest Fix: Train/Test Split

Split your data into two pieces:

- **Training set** (~70-80%): build the model
- **Test set** (~20-30%): evaluate the model

The model never sees the test data during training. Its performance on the
test set tells you how it will do on genuinely new data.

```{r}
#| label: fig-train-test-split
#| fig-cap: "Split data into training (build the model) and test (evaluate it)"

set.seed(7)
n <- 40
x_all <- sort(runif(n, 0, 10))
y_all <- 2 * sin(x_all) + rnorm(n, 0, 0.8)

# Random 70/30 split
train_idx <- sample(n, round(0.7 * n))
test_idx <- setdiff(1:n, train_idx)

df_all <- data.frame(
  x = x_all, y = y_all,
  set = ifelse(1:n %in% train_idx, "Train", "Test")
)

ggplot(df_all, aes(x, y, color = set, shape = set)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("coral", "steelblue")) +
  scale_shape_manual(values = c(17, 19)) +
  theme_minimal(base_size = 14) +
  labs(x = "x", y = "y",
    title = paste0("70/30 split: ", length(train_idx), " train, ",
                   length(test_idx), " test"))
```

```{r}
#| label: fig-train-test-eval
#| fig-cap: "Overfit model: great on train, terrible on test. Good model: consistent on both."

df_train <- df_all[df_all$set == "Train", ]
df_test <- df_all[df_all$set == "Test", ]

# Overfit
fit_over <- lm(y ~ poly(x, 15), data = df_train)
# Good
fit_good <- lm(y ~ poly(x, 3), data = df_train)

# Errors
results <- data.frame(
  Model = rep(c("Overfit (degree 15)", "Good (degree 3)"), each = 2),
  Set = rep(c("Train", "Test"), 2),
  RMSE = c(
    sqrt(mean(residuals(fit_over)^2)),
    sqrt(mean((df_test$y - predict(fit_over, df_test))^2)),
    sqrt(mean(residuals(fit_good)^2)),
    sqrt(mean((df_test$y - predict(fit_good, df_test))^2))
  )
)

ggplot(results, aes(x = Model, y = RMSE, fill = Set)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("coral", "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(y = "RMSE (lower = better)", title = "The reveal: test performance exposes overfitting") +
  coord_flip()
```

The overfit model looks amazing on training data but falls apart on test data.
The good model performs **consistently** on both.

### The Problem with a Single Split

But there's a catch: **which 30% you hold out matters**. A different random
split gives a different score. Your evaluation depends on luck.

```{r}
#| label: fig-split-variance
#| fig-cap: "Different random splits give different test scores — which one is 'right'?"

set.seed(1)
scores <- sapply(1:20, function(i) {
  idx <- sample(n, round(0.7 * n))
  train <- data.frame(x = x_all[idx], y = y_all[idx])
  test <- data.frame(x = x_all[-idx], y = y_all[-idx])
  fit <- lm(y ~ poly(x, 3), data = train)
  sqrt(mean((test$y - predict(fit, test))^2))
})

ggplot(data.frame(trial = 1:20, rmse = scores), aes(trial, rmse)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = mean(scores), color = "coral", linewidth = 1, linetype = "dashed") +
  annotate("text", x = 17, y = mean(scores) + 0.1,
           label = paste0("Average: ", round(mean(scores), 2)),
           color = "coral", size = 4) +
  theme_minimal(base_size = 14) +
  labs(x = "Random Split #", y = "Test RMSE",
       title = "Same model, 20 different splits → 20 different scores")
```

**Solution:** Don't do one split. Do **many** splits and **average**. That's
cross-validation.


## 3. Cross-Validation: The Core Idea

### k-Fold Cross-Validation

1. Divide data into $k$ equal **folds** (groups)
2. Use fold 1 as test, train on folds 2-5 → get score 1
3. Use fold 2 as test, train on folds 1,3-5 → get score 2
4. ... repeat for all folds
5. **Average** all $k$ scores

Every point gets to be in the test set **exactly once**.

```{r}
#| label: fig-kfold-diagram
#| fig-cap: "5-fold cross-validation: each fold takes a turn as the test set"

# Visual diagram of k-fold CV
k <- 5
fold_colors <- c("coral", "steelblue")

par(mar = c(2, 6, 3, 2))
plot(NULL, xlim = c(0, k), ylim = c(0.5, k + 0.5),
     xlab = "", ylab = "", xaxt = "n", yaxt = "n",
     main = "5-Fold Cross-Validation")

for (round in 1:k) {
  for (fold in 1:k) {
    col <- ifelse(fold == round,
      adjustcolor("coral", 0.8),
      adjustcolor("steelblue", 0.4))
    rect(fold - 1, k - round + 0.6, fold, k - round + 1.4, col = col, border = "white", lwd = 2)
    label <- ifelse(fold == round, "TEST", "train")
    text(fold - 0.5, k - round + 1, label, cex = 0.9,
         col = ifelse(fold == round, "white", "white"),
         font = ifelse(fold == round, 2, 1))
  }
  text(-0.3, k - round + 1, paste("Round", round), cex = 0.8, adj = 1)
}

text(k / 2, 0.3, "→ Average all 5 scores = cross-validation estimate", cex = 1, font = 2)
```

### The Formula

The cross-validation estimate of model performance is just the average error
across all folds:

$$\text{CV}(k) = \frac{1}{k} \sum_{i=1}^{k} \text{Error}_i$$

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $k$ | Number of folds | How many splits you make |
| $\text{Error}_i$ | Error on fold $i$ when it's the test set | How badly the model did on that fold |
| $\text{CV}(k)$ | Cross-validation score | The averaged, reliable performance estimate |

For classification, "Error" is usually **misclassification rate** (% wrong).
For regression, it's usually **RMSE** or **MSE**.


## 4. Choosing k (the Number of Folds)

Confusing but important: **this $k$ is different from KNN's $k$.**
This is the number of folds, not the number of neighbors.

| Folds | Name | Train size | Test size | Pros | Cons |
|:-----:|:-----|:-----------|:----------|:-----|:-----|
| 5 | 5-fold | 80% | 20% | Fast, good balance | Moderate variance |
| 10 | 10-fold | 90% | 10% | Better estimate | Slower |
| $n$ | Leave-one-out (LOO) | $n-1$ | 1 | Uses max training data | Very slow, high variance |

```{r}
#| label: fig-fold-comparison
#| fig-cap: "More folds = more stable estimate, but diminishing returns past 10"

set.seed(42)
n <- 50
x <- sort(runif(n, 0, 10))
y <- 2 * sin(x) + rnorm(n, 0, 0.8)
df_cv <- data.frame(x = x, y = y)

# Run k-fold CV for different k values, repeat to see variance
fold_vals <- c(3, 5, 10, 25, n)
fold_labels <- c("3", "5", "10", "25", paste0(n, "\n(LOO)"))

results_list <- list()
for (fi in seq_along(fold_vals)) {
  kf <- fold_vals[fi]
  scores <- sapply(1:30, function(rep) {
    folds <- sample(rep(1:kf, length.out = n))
    errs <- sapply(1:kf, function(f) {
      train <- df_cv[folds != f, ]
      test <- df_cv[folds == f, ]
      if (nrow(train) < 4 || nrow(test) < 1) return(NA)
      fit <- lm(y ~ poly(x, 3), data = train)
      mean((test$y - predict(fit, test))^2)
    })
    mean(errs, na.rm = TRUE)
  })
  results_list[[fi]] <- data.frame(
    folds = fold_labels[fi],
    score = scores
  )
}

results_all <- do.call(rbind, results_list)
results_all$folds <- factor(results_all$folds, levels = fold_labels)

ggplot(results_all, aes(folds, score)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.size = 1) +
  theme_minimal(base_size = 14) +
  labs(x = "Number of Folds", y = "CV Score (MSE)",
       title = "More folds = less variance in the estimate")
```

**Rule of thumb for this guide:** Use **5-fold or 10-fold**. These are the
standard choices and almost always sufficient.


## 5. What Cross-Validation Is Actually Used For

Cross-validation answers two questions:

### Question 1: How good is this model?

"What accuracy will my SVM get on new data?" → Train with CV, report the
average score.

### Question 2: Which parameter value is best?

This is the **primary use in this guide**: choosing hyperparameters.

- What $k$ should I use in KNN?
- What $C$ should I use in SVM?
- How many components in PCA?
- Which variables in regression?

```{r}
#| label: fig-tune-k-knn
#| fig-cap: "Cross-validation finds the best k for KNN by testing each value"

library(class)
set.seed(42)

# Generate classification data
df_class <- data.frame(
  x1 = c(rnorm(50, 2, 1), rnorm(50, 4.5, 1)),
  x2 = c(rnorm(50, 2, 1), rnorm(50, 4.5, 1)),
  group = factor(rep(c("A", "B"), each = 50))
)

# 5-fold CV for different k values
k_values <- c(1, 3, 5, 7, 9, 11, 15, 21, 31, 45)
n_obs <- nrow(df_class)
folds <- sample(rep(1:5, length.out = n_obs))

cv_results <- sapply(k_values, function(k_nn) {
  fold_acc <- sapply(1:5, function(f) {
    train <- df_class[folds != f, ]
    test <- df_class[folds == f, ]
    pred <- knn(train[, 1:2], test[, 1:2], train$group, k = k_nn)
    mean(pred == test$group)
  })
  mean(fold_acc)
})

best_k <- k_values[which.max(cv_results)]

ggplot(data.frame(k = k_values, accuracy = cv_results), aes(k, accuracy)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = best_k, linetype = "dashed", color = "coral") +
  annotate("text", x = best_k + 3, y = min(cv_results) + 0.01,
           label = paste0("Best k = ", best_k), color = "coral", size = 4) +
  scale_x_continuous(breaks = k_values) +
  theme_minimal(base_size = 14) +
  labs(x = "k (number of neighbors)", y = "CV Accuracy (5-fold)",
       title = "Using cross-validation to choose k in KNN")
```

```{r}
#| label: fig-tune-c-svm
#| fig-cap: "Same idea for SVM: cross-validate over different C values"

library(e1071)

c_values <- c(0.001, 0.01, 0.1, 1, 10, 100, 1000)

cv_svm <- sapply(c_values, function(C) {
  fold_acc <- sapply(1:5, function(f) {
    train <- df_class[folds != f, ]
    test <- df_class[folds == f, ]
    fit <- svm(group ~ x1 + x2, data = train, kernel = "linear", cost = C)
    pred <- predict(fit, test)
    mean(pred == test$group)
  })
  mean(fold_acc)
})

best_c <- c_values[which.max(cv_svm)]

ggplot(data.frame(C = c_values, accuracy = cv_svm), aes(C, accuracy)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = best_c, linetype = "dashed", color = "coral") +
  annotate("text", x = best_c * 5, y = min(cv_svm) + 0.005,
           label = paste0("Best C = ", best_c), color = "coral", size = 4) +
  scale_x_log10() +
  theme_minimal(base_size = 14) +
  labs(x = "C (cost parameter, log scale)", y = "CV Accuracy (5-fold)",
       title = "Using cross-validation to choose C in SVM")
```


## 6. The Train / Validate / Test Split

When you use CV to **tune** parameters, the CV folds become your validation
set. You still need a **separate test set** that was never part of any tuning.

```
┌──────────────────────────────────────────────────────────┐
│  ALL DATA                                                │
│                                                          │
│  ┌────────────────────────────────┐  ┌────────────────┐  │
│  │  Training + Validation (80%)   │  │  Test Set (20%) │  │
│  │                                │  │                 │  │
│  │  ← cross-validation happens    │  │  ← touched ONCE │  │
│  │    here to tune parameters     │  │    at the very  │  │
│  │                                │  │    end          │  │
│  └────────────────────────────────┘  └────────────────┘  │
└──────────────────────────────────────────────────────────┘
```

**The workflow:**

1. Set aside 20% as the **test set** (lock it away)
2. Use the remaining 80% for **k-fold CV** to choose parameters
3. Retrain the final model on all 80% with the best parameters
4. Evaluate **once** on the test set → this is your reported performance

**Why?** If you tune on test data, you're indirectly fitting to it. The test
set must be truly unseen.

```{r}
#| label: fig-full-workflow
#| fig-cap: "The complete model-building workflow with cross-validation"

# Simulate the full workflow
set.seed(42)

# Step 1: Hold out test set
test_idx <- sample(n_obs, round(0.2 * n_obs))
train_val <- df_class[-test_idx, ]
test_final <- df_class[test_idx, ]

# Step 2: CV on train_val to find best k
folds_tv <- sample(rep(1:5, length.out = nrow(train_val)))

cv_k <- sapply(k_values, function(k_nn) {
  fold_acc <- sapply(1:5, function(f) {
    tr <- train_val[folds_tv != f, ]
    va <- train_val[folds_tv == f, ]
    pred <- knn(tr[, 1:2], va[, 1:2], tr$group, k = k_nn)
    mean(pred == va$group)
  })
  mean(fold_acc)
})

best_k_final <- k_values[which.max(cv_k)]

# Step 3: Evaluate on test set with best k
test_pred <- knn(train_val[, 1:2], test_final[, 1:2], train_val$group, k = best_k_final)
test_acc <- mean(test_pred == test_final$group)

cat(sprintf("Best k from CV: %d\n", best_k_final))
cat(sprintf("CV accuracy (validation): %.1f%%\n", max(cv_k) * 100))
cat(sprintf("Final test accuracy: %.1f%%\n", test_acc * 100))
cat("\nIf these two numbers are close → model generalizes well")
cat("\nIf test << CV → you overfit during tuning (rare with proper CV)")
```


## 7. Common Mistakes (Common Pitfalls)

::: {.callout-warning}
### Trap 1: Testing on Training Data

> "Our model achieves 98% accuracy!"
>
> "On what data?"
>
> "The training data."
>
> **Meaningless.** A model that memorizes gets 100% on training data.
:::

::: {.callout-warning}
### Trap 2: Confusing Training Accuracy with Real Performance

High training accuracy + low validation accuracy = **overfitting**.

The **validation score is the one that matters.**
:::

```{r}
#| label: fig-overfit-gap
#| fig-cap: "The gap between training and validation accuracy reveals overfitting"

# Show train vs validation accuracy across model complexity
complexities <- 1:15
train_acc <- numeric(15)
val_acc <- numeric(15)

set.seed(42)
folds_gap <- sample(rep(1:5, length.out = nrow(df_cv)))

for (d in complexities) {
  # Training accuracy (full data)
  fit <- lm(y ~ poly(x, d), data = df_cv)
  train_acc[d] <- 1 - mean(residuals(fit)^2) / var(df_cv$y)

  # CV accuracy
  fold_r2 <- sapply(1:5, function(f) {
    tr <- df_cv[folds_gap != f, ]
    te <- df_cv[folds_gap == f, ]
    if (nrow(tr) < d + 1) return(NA)
    fit_cv <- lm(y ~ poly(x, min(d, nrow(tr) - 2)), data = tr)
    pred <- predict(fit_cv, te)
    1 - mean((te$y - pred)^2) / var(te$y)
  })
  val_acc[d] <- mean(fold_r2, na.rm = TRUE)
}

gap_df <- data.frame(
  complexity = rep(complexities, 2),
  r_squared = c(train_acc, val_acc),
  set = rep(c("Training", "Validation (CV)"), each = 15)
)

ggplot(gap_df, aes(complexity, r_squared, color = set)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("steelblue", "coral")) +
  geom_vline(xintercept = which.max(val_acc), linetype = "dashed", color = "gray40") +
  annotate("text", x = which.max(val_acc) + 1.5, y = 0.3,
           label = paste0("Best complexity = ", which.max(val_acc)),
           color = "gray40", size = 3.5) +
  annotate("text", x = 12, y = 0.85, label = "OVERFIT ZONE",
           color = "coral", size = 4, fontface = "bold") +
  theme_minimal(base_size = 14) +
  labs(x = "Model Complexity (polynomial degree)", y = "R²",
       color = "Evaluated on", title = "Training R² always goes up — validation reveals the truth")
```

::: {.callout-warning}
### Trap 3: Using Test Data to Tune Parameters

If you try 20 different values of $k$ on the test set and pick the best one,
you've **fit to the test set**. It's no longer a fair evaluation.

**Rule:** Test set = used **once**, at the very end.
:::

::: {.callout-warning}
### Trap 4: Not Enough Data in Each Fold

If $n = 20$ and $k = 10$, each test fold has only 2 points. That's not enough
for a reliable error estimate. Use fewer folds with small datasets.
:::


## 8. Special Case: Leave-One-Out Cross-Validation (LOOCV)

When $k = n$ (number of folds = number of data points):

- Each round: train on $n-1$ points, test on 1 point
- Repeat $n$ times
- Average all $n$ errors

```{r}
#| label: fig-loocv
#| fig-cap: "LOOCV: each point gets a turn as the sole test point"

# Visual: highlight one point at a time
par(mfrow = c(2, 3), mar = c(3, 3, 2, 1))

set.seed(42)
small_df <- data.frame(
  x = c(1, 2, 3, 5, 6, 8),
  y = c(2.1, 3.8, 4.5, 7.2, 8.1, 10.5)
)

for (i in 1:6) {
  train <- small_df[-i, ]
  test <- small_df[i, ]
  fit <- lm(y ~ x, data = train)

  plot(small_df$x, small_df$y, pch = 19, cex = 1.5,
    col = ifelse(1:6 == i, "coral", "steelblue"),
    main = paste("Round", i, "— test point", i),
    xlab = "x", ylab = "y", xlim = c(0, 9), ylim = c(0, 12))
  abline(fit, col = "gray50")
  pred <- predict(fit, test)
  segments(test$x, test$y, test$x, pred, col = "coral", lwd = 2, lty = 2)
  text(test$x + 0.5, (test$y + pred) / 2,
       paste0("err=", round(abs(test$y - pred), 2)),
       col = "coral", cex = 0.8)
}
par(mfrow = c(1, 1))
```

**LOOCV pros:** Uses maximum training data (n-1 points each round).
**LOOCV cons:** Runs $n$ times (slow for large data), high variance.

**When to use:** Small datasets where every point matters.


## 9. Cheat Sheet: The Whole Story on One Page

```
CROSS-VALIDATION RECIPE
========================

1. PURPOSE: Estimate how well a model generalizes to unseen data

2. k-FOLD CV:
   - Split data into k equal folds
   - Each fold takes a turn as test set
   - Average the k scores
   - Standard choice: k = 5 or k = 10

3. THE FORMULA:
   CV(k) = (1/k) × Σ Errorᵢ

4. PRIMARY USE: Tune hyperparameters
   - KNN: which k (neighbors)?        → CV over k = 1,3,5,7,...
   - SVM: which C (cost)?             → CV over C = 0.01, 0.1, 1, 10,...
   - Regression: which variables?     → CV with different subsets
   - PCA: how many components?        → CV with 1, 2, 3, ... components

5. THREE-WAY SPLIT:
   Training (fit model) → Validation (tune params via CV) → Test (final eval)
   Test set: touched ONCE, at the very end

6. OVERFITTING DETECTION:
   Training accuracy >> Validation accuracy = OVERFIT
   Training accuracy ≈ Validation accuracy = GOOD

7. COMMON PITFALLS:
   - Never evaluate on training data
   - Never tune on test data
   - High training accuracy alone means NOTHING
   - The validation/CV score is what matters
```


## 10. Check Your Understanding

::: {.callout-note}
### Test Yourself

Before moving on, try to answer these without scrolling up:

1. Why can't you evaluate a model on its training data?
2. What does k-fold cross-validation do, step by step?
3. What's the difference between the validation set and the test set?
4. How do you use cross-validation to choose $k$ in KNN?
5. You build a model that gets 95% training accuracy and 62% validation
   accuracy. What's happening? What would you do?
6. Why is leave-one-out CV sometimes worse than 10-fold, despite using
   more training data per round?
7. A classmate says "I tried 50 different models on the test set and picked
   the best one." What's wrong with this?
:::
