---
title: "CUSUM From Scratch: Catching the Moment Everything Changed"
subtitle: "Change detection — when did the process shift?"
engine: knitr
---

## 1. The Problem: Did Something Change?

A factory makes bolts that should be exactly 10 cm long. Every hour you
measure a bolt. The measurements bounce around a little — that's normal
random variation. But at some point, the machine drifts and starts making
bolts that are slightly too long.

**When did the change happen?**

You can't just look at one measurement — any single value might be high by
chance. You need a method that **accumulates evidence** over time and shouts
"something changed!" only when it's confident.

That's CUSUM.

```{r}
#| label: fig-the-problem
#| fig-cap: "Can you spot where the process shifted? It's hard to tell from individual points."

library(ggplot2)
set.seed(42)

# Normal process, then a shift at t = 60
n <- 100
mu <- 10       # expected mean
shift <- 0.5   # how much the mean shifts

x <- c(rnorm(60, mu, 0.3), rnorm(40, mu + shift, 0.3))
time <- 1:n

df <- data.frame(time = time, value = x)

ggplot(df, aes(time, value)) +
  geom_point(color = "gray40", size = 1.5) +
  geom_hline(yintercept = mu, color = "steelblue", linewidth = 1, linetype = "dashed") +
  annotate("text", x = 10, y = mu + 0.15, label = "Expected mean (μ = 10)",
           color = "steelblue", size = 3.5) +
  geom_vline(xintercept = 60, color = "coral", linetype = "dotted", linewidth = 0.8) +
  annotate("text", x = 63, y = 9.2, label = "True shift\n(unknown to us)",
           color = "coral", size = 3.5) +
  theme_minimal(base_size = 14) +
  labs(x = "Time", y = "Measurement (cm)",
       title = "Individual measurements — the shift is buried in noise")
```

Each point looks reasonable on its own. The shift is only 0.5 cm in a process
with 0.3 cm of natural variation. CUSUM will find it.


## 2. The Idea: Keep a Running Score

CUSUM stands for **CU**mulative **SUM**. The idea:

1. Start with a score of **zero**
2. Each time step, check: is this observation **above** what we expect?
3. If yes, **add** to the score (evidence is accumulating)
4. If no, **subtract** from the score (but never go below zero)
5. When the score crosses a **threshold** → declare a change

It's like a detective building a case. Each above-average observation is a
small piece of evidence. One piece isn't convincing. But when evidence keeps
piling up, eventually you're confident something changed.

```{r}
#| label: fig-accumulation
#| fig-cap: "CUSUM accumulates positive deviations. When they pile up, a change is detected."

# Simple visual: show deviations accumulating
deviations <- x - mu

colors <- ifelse(deviations > 0, "coral", "steelblue")

par(mfrow = c(2, 1), mar = c(3, 4, 3, 1))

# Top: individual deviations
barplot(deviations, col = adjustcolor(colors, 0.6), border = NA,
  main = "Individual deviations from μ (each observation - 10)",
  ylab = "Deviation", names.arg = rep("", n))
abline(h = 0)
mtext("After the shift, more bars are positive (red)", side = 1, line = 1, cex = 0.8)

# Bottom: cumulative sum
cumdev <- cumsum(deviations)
plot(time, cumdev, type = "l", lwd = 2, col = "steelblue",
  main = "Naive cumulative sum of deviations",
  ylab = "Cumulative Sum", xlab = "Time")
abline(h = 0, lty = 2)
abline(v = 60, col = "coral", lty = 3)
text(65, max(cumdev) * 0.3, "Shift here →", col = "coral", cex = 0.9)

par(mfrow = c(1, 1))
```

The naive cumulative sum shows the trend, but it's noisy before the shift
because random below-average values create a downward drift. CUSUM fixes this
with two clever additions: the **allowance** (C) and the **floor at zero**.


## 3. The CUSUM Formula

$$S_t = \max(0, \; S_{t-1} + (x_t - \mu) - C)$$

**Declare a change when** $S_t > T$

That's the whole formula. Let's unpack every piece:

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $S_t$ | CUSUM statistic at time $t$ | The running score (accumulated evidence) |
| $S_{t-1}$ | Previous score | What we had so far |
| $x_t$ | Current observation | The new measurement |
| $\mu$ | Expected mean | What the process *should* be producing |
| $x_t - \mu$ | Deviation | How far this observation is from expected |
| $C$ | Allowance (critical value) | Filters out normal noise — only counts "big enough" deviations |
| $\max(0, \dots)$ | Floor at zero | Prevents negative accumulation — past good behavior doesn't mask future problems |
| $T$ | Threshold | How much evidence you need before declaring a change |

### Step by Step

At each time step:

1. Take the new observation $x_t$
2. Compute the deviation: $x_t - \mu$ (how far from expected?)
3. Subtract the allowance $C$ (ignore small, normal fluctuations)
4. Add to previous score $S_{t-1}$
5. If the result is negative, reset to 0 (fresh start)
6. If $S_t > T$, **sound the alarm**

```{r}
#| label: fig-cusum-formula
#| fig-cap: "CUSUM in action: the score accumulates after the shift and crosses the threshold"

# Compute CUSUM
C_val <- 0.2  # allowance
T_val <- 4    # threshold

S <- numeric(n)
S[1] <- max(0, (x[1] - mu) - C_val)

for (t in 2:n) {
  S[t] <- max(0, S[t - 1] + (x[t] - mu) - C_val)
}

# When does it cross T?
alarm_time <- which(S > T_val)[1]

df_cusum <- data.frame(time = time, S = S)

ggplot(df_cusum, aes(time, S)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_hline(yintercept = T_val, color = "coral", linewidth = 1, linetype = "dashed") +
  annotate("text", x = 15, y = T_val + 0.5,
           label = paste0("Threshold T = ", T_val), color = "coral", size = 4) +
  geom_vline(xintercept = 60, color = "gray50", linetype = "dotted") +
  annotate("text", x = 55, y = max(S) * 0.8, label = "True shift", color = "gray50", size = 3.5) +
  { if (!is.na(alarm_time))
    list(
      geom_vline(xintercept = alarm_time, color = "darkred", linewidth = 1),
      annotate("text", x = alarm_time + 3, y = T_val + 1.5,
               label = paste0("ALARM at t = ", alarm_time),
               color = "darkred", fontface = "bold", size = 4)
    )
  } +
  theme_minimal(base_size = 14) +
  labs(x = "Time", y = expression(S[t] ~ "(CUSUM statistic)"),
       title = "CUSUM detects the shift")
```


## 4. What C and T Do (The Two Knobs)

### C (Allowance): How Much Noise to Ignore

$C$ is a filter. It subtracts from each deviation so that normal random
fluctuations don't accumulate.

- **Large C**: Only big deviations count → slower detection, fewer false alarms
- **Small C**: Small deviations count → faster detection, more false alarms
- **C = 0**: Every above-average observation adds to the score (very sensitive)

```{r}
#| label: fig-c-effect
#| fig-cap: "Larger C = slower detection but fewer false alarms"

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

for (C_test in c(0.05, 0.2, 0.5)) {
  S_test <- numeric(n)
  S_test[1] <- max(0, (x[1] - mu) - C_test)
  for (t in 2:n) {
    S_test[t] <- max(0, S_test[t - 1] + (x[t] - mu) - C_test)
  }

  alarm <- which(S_test > T_val)[1]

  plot(time, S_test, type = "l", lwd = 2, col = "steelblue",
    main = paste0("C = ", C_test),
    xlab = "Time", ylab = expression(S[t]),
    ylim = c(0, max(S_test, T_val + 1)))
  abline(h = T_val, col = "coral", lty = 2, lwd = 1.5)
  abline(v = 60, col = "gray50", lty = 3)

  if (!is.na(alarm)) {
    abline(v = alarm, col = "darkred", lwd = 1.5)
    legend("topleft", bty = "n", cex = 0.8,
      legend = c(
        paste0("Alarm at t = ", alarm),
        paste0("Delay: ", alarm - 60, " steps")
      ))
  } else {
    legend("topleft", bty = "n", cex = 0.8, legend = "No alarm triggered")
  }
}
par(mfrow = c(1, 1))
```

### T (Threshold): How Much Evidence Before Alarm

$T$ controls how **certain** you need to be before declaring a change.

- **Large T**: Need lots of evidence → fewer false alarms, but slower detection
- **Small T**: Trigger on less evidence → faster detection, more false alarms

```{r}
#| label: fig-t-effect
#| fig-cap: "Larger T = more evidence needed before triggering an alarm"

# Fix C, vary T
C_fixed <- 0.2

S_fixed <- numeric(n)
S_fixed[1] <- max(0, (x[1] - mu) - C_fixed)
for (t in 2:n) {
  S_fixed[t] <- max(0, S_fixed[t - 1] + (x[t] - mu) - C_fixed)
}

T_vals <- c(2, 4, 8)
alarm_times <- sapply(T_vals, function(Tv) which(S_fixed > Tv)[1])

df_t <- data.frame(time = time, S = S_fixed)

ggplot(df_t, aes(time, S)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_hline(yintercept = T_vals[1], color = "coral", linetype = "dashed") +
  geom_hline(yintercept = T_vals[2], color = "darkorange", linetype = "dashed") +
  geom_hline(yintercept = T_vals[3], color = "darkred", linetype = "dashed") +
  annotate("text", x = 10, y = T_vals[1] + 0.4,
           label = paste0("T=2 → alarm at t=", alarm_times[1]), color = "coral", size = 3.5) +
  annotate("text", x = 10, y = T_vals[2] + 0.4,
           label = paste0("T=4 → alarm at t=", alarm_times[2]), color = "darkorange", size = 3.5) +
  annotate("text", x = 10, y = T_vals[3] + 0.4,
           label = paste0("T=8 → alarm at t=",
                          ifelse(is.na(alarm_times[3]), "never", alarm_times[3])),
           color = "darkred", size = 3.5) +
  geom_vline(xintercept = 60, color = "gray50", linetype = "dotted") +
  theme_minimal(base_size = 14) +
  labs(x = "Time", y = expression(S[t]),
       title = "Same data, different thresholds")
```


## 5. The Speed vs. False Alarm Tradeoff

This is the **central tradeoff** to understand. There's no free lunch.

```
     Fast detection                    Slow detection
     More false alarms                 Fewer false alarms
     ◄─────────────────────────────────────────────────►
     Small C, Small T                  Large C, Large T
```

**The right C and T depend on the costs:**

| Scenario | Want | Set |
|:---------|:-----|:----|
| Nuclear reactor monitoring | Fast detection (safety!) | Small C, small T |
| Marketing campaign tracking | Few false alarms (budget!) | Large C, large T |
| Manufacturing quality | Balance both | Moderate C and T |

::: {.callout-warning}
### Common Pitfall: T and C Parameters

"What happens if we increase T?" → Fewer false alarms, slower
detection. "What happens if we increase C?" → Same effect. There's no
universal best value — it depends on the application.
:::


## 6. Why max(0, ...) Matters

The floor at zero is crucial. Without it, a long stretch of below-average
values would build up a large negative score. Then when a real shift happens,
the score has to climb back from that hole before triggering an alarm.

```{r}
#| label: fig-max-zero
#| fig-cap: "Without max(0,...), past good behavior delays detection of real changes"

C_demo <- 0.2

# With max(0, ...)
S_with <- numeric(n)
S_with[1] <- max(0, (x[1] - mu) - C_demo)
for (t in 2:n) {
  S_with[t] <- max(0, S_with[t - 1] + (x[t] - mu) - C_demo)
}

# Without max(0, ...) — naive version
S_without <- numeric(n)
S_without[1] <- (x[1] - mu) - C_demo
for (t in 2:n) {
  S_without[t] <- S_without[t - 1] + (x[t] - mu) - C_demo
}

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(time, S_with, type = "l", lwd = 2, col = "steelblue",
  main = "With max(0,...)",
  xlab = "Time", ylab = expression(S[t]))
abline(h = T_val, col = "coral", lty = 2)
abline(v = 60, col = "gray50", lty = 3)
alarm1 <- which(S_with > T_val)[1]
if (!is.na(alarm1)) {
  abline(v = alarm1, col = "darkred")
  text(alarm1 + 3, T_val + 1, paste("t =", alarm1), col = "darkred", cex = 0.9)
}

plot(time, S_without, type = "l", lwd = 2, col = "steelblue",
  main = "Without max(0,...)",
  xlab = "Time", ylab = expression(S[t]))
abline(h = T_val, col = "coral", lty = 2)
abline(v = 60, col = "gray50", lty = 3)
alarm2 <- which(S_without > T_val)[1]
if (!is.na(alarm2)) {
  abline(v = alarm2, col = "darkred")
  text(alarm2 + 3, T_val + 1, paste("t =", alarm2), col = "darkred", cex = 0.9)
}

par(mfrow = c(1, 1))
```

The version without `max(0,...)` digs into a negative hole during the normal
period, then takes longer to climb back up after the shift. The proper CUSUM
resets to zero whenever it would go negative — every moment is a potential
fresh start for detecting a change.


## 7. Detecting Decreases (Two-Sided CUSUM)

The formula above only detects **increases** (upward shifts). To detect
**decreases**, run a second CUSUM in the other direction:

$$S_t^+ = \max(0, \; S_{t-1}^+ + (x_t - \mu) - C) \quad \text{(detects increase)}$$
$$S_t^- = \max(0, \; S_{t-1}^- - (x_t - \mu) - C) \quad \text{(detects decrease)}$$

Alarm when either $S_t^+ > T$ or $S_t^- > T$.

```{r}
#| label: fig-two-sided
#| fig-cap: "Two-sided CUSUM catches both upward and downward shifts"

set.seed(99)

# Process with a DECREASE at t=50
x2 <- c(rnorm(50, 10, 0.3), rnorm(50, 9.4, 0.3))

S_up <- numeric(n)
S_down <- numeric(n)
C2 <- 0.2

for (t in 2:n) {
  S_up[t] <- max(0, S_up[t - 1] + (x2[t] - mu) - C2)
  S_down[t] <- max(0, S_down[t - 1] - (x2[t] - mu) - C2)
}

par(mfrow = c(2, 1), mar = c(3, 4, 3, 1))

plot(time, x2, pch = 19, cex = 0.8, col = "gray40",
  main = "Data with a DECREASE at t = 50",
  ylab = "Value", xlab = "")
abline(h = mu, col = "steelblue", lty = 2)
abline(v = 50, col = "gray50", lty = 3)

plot(time, S_up, type = "l", lwd = 2, col = "coral",
  ylim = c(0, max(c(S_up, S_down, T_val + 1))),
  main = "Two-sided CUSUM", ylab = expression(S[t]), xlab = "Time")
lines(time, S_down, lwd = 2, col = "steelblue")
abline(h = T_val, lty = 2, col = "gray40")
legend("topleft", bty = "n",
  legend = c("S⁺ (detects increase)", "S⁻ (detects decrease)", "Threshold"),
  col = c("coral", "steelblue", "gray40"), lty = c(1, 1, 2), lwd = 2, cex = 0.8)

alarm_down <- which(S_down > T_val)[1]
if (!is.na(alarm_down)) {
  abline(v = alarm_down, col = "darkred")
  text(alarm_down + 3, T_val + 0.5, paste("Alarm t =", alarm_down),
       col = "darkred", cex = 0.9)
}

par(mfrow = c(1, 1))
```


## 8. CUSUM Cannot Tell You WHY

::: {.callout-warning}
### Common Pitfall: CUSUM Cannot Tell You WHY

CUSUM tells you **when** something changed. It does
**not** tell you **what** changed or **why**.

> "CUSUM detected a shift in manufacturing quality at day 60."
>
> **Wrong conclusion:** "The machine is broken."
>
> **Right conclusion:** "Something changed around day 60. We need to
> investigate what happened."
:::

CUSUM is a **detection** tool, not a **diagnosis** tool. The investigation
happens after the alarm.


## 9. Real-World Applications

| Application | What $x_t$ measures | What "change" means |
|:------------|:-------------------|:-------------------|
| Manufacturing | Part dimension | Machine needs recalibration |
| Healthcare | Patient vitals | Condition is deteriorating |
| Finance | Transaction amounts | Potential fraud pattern |
| Website | Page load times | Server performance degraded |
| Retail | Daily sales | Market shift or competitor action |

```{r}
#| label: fig-real-world
#| fig-cap: "CUSUM applied to website response times — detects server degradation"

set.seed(42)

# Website response time (ms)
days <- 1:90
resp_time <- c(
  rnorm(50, 200, 20),   # normal
  rnorm(40, 230, 25)    # degraded after day 50
)

mu_web <- 200
C_web <- 5
T_web <- 80

S_web <- numeric(90)
for (t in 2:90) {
  S_web[t] <- max(0, S_web[t - 1] + (resp_time[t] - mu_web) - C_web)
}

alarm_web <- which(S_web > T_web)[1]

par(mfrow = c(2, 1), mar = c(3, 4, 3, 1))

plot(days, resp_time, pch = 19, cex = 0.8, col = "gray40",
  main = "Website Response Time (ms)",
  ylab = "Response Time", xlab = "")
abline(h = mu_web, col = "steelblue", lty = 2)

plot(days, S_web, type = "l", lwd = 2, col = "steelblue",
  main = "CUSUM Monitoring", ylab = expression(S[t]), xlab = "Day")
abline(h = T_web, col = "coral", lty = 2)
if (!is.na(alarm_web)) {
  abline(v = alarm_web, col = "darkred")
  text(alarm_web + 3, T_web + 10,
    paste0("Alert: investigate server\n(day ", alarm_web, ")"),
    col = "darkred", cex = 0.8)
}

par(mfrow = c(1, 1))
```


## 10. CUSUM vs Other Methods

| Method | Detects | When to use |
|:-------|:--------|:------------|
| **CUSUM** | Shift in process mean | Ongoing monitoring, small persistent shifts |
| **Control chart** | Individual outlier points | Quick spikes, obvious violations |
| **t-test** | Difference between two groups | Comparing before/after with known breakpoint |
| **Exponential smoothing** | Trends and patterns | Forecasting future values |

CUSUM's superpower is detecting **small, gradual shifts** that individual
observations can't catch. A single point might look normal, but CUSUM
accumulates the evidence until the pattern is undeniable.


## 11. Cheat Sheet: The Whole Story on One Page

```
THE CUSUM RECIPE
=================

1. PURPOSE: Detect when a process mean has shifted

2. THE FORMULA:
   Sₜ = max(0, Sₜ₋₁ + (xₜ - μ) - C)

   Alarm when Sₜ > T

3. SYMBOLS:
   Sₜ  = accumulated evidence (running score)
   xₜ  = current observation
   μ   = expected baseline mean
   C   = allowance (noise filter)
   T   = threshold (alarm trigger)

4. KEY PARAMETERS:
   C → how much noise to ignore
   T → how much evidence needed for alarm
   Both: ↑ value = slower detection, fewer false alarms
         ↓ value = faster detection, more false alarms

5. WHY max(0,...)?
   Prevents past good behavior from masking future changes.
   Every moment is a fresh start for detection.

6. TWO-SIDED:
   S⁺ detects increases, S⁻ detects decreases

7. CUSUM CANNOT:
   - Tell you WHY something changed
   - Tell you WHAT changed
   - Only tells you WHEN

8. COMMON PITFALLS:
   - No universal best C and T (depends on costs)
   - "Increase T" → fewer false alarms, slower detection
   - CUSUM detects, it doesn't diagnose
```


## 12. Check Your Understanding

::: {.callout-note}
### Test Yourself

Before moving on, try to answer these without scrolling up:

1. What does CUSUM stand for and what does it detect?
2. Walk through the formula $S_t = \max(0, S_{t-1} + (x_t - \mu) - C)$
   in plain English.
3. What does $C$ (allowance) control? What happens if you increase it?
4. What does $T$ (threshold) control? What happens if you decrease it?
5. Why is the $\max(0, \dots)$ necessary? What would happen without it?
6. CUSUM triggers an alarm. Can you conclude the machine is broken? Why
   or why not?
7. A nuclear power plant wants fast detection at all costs. A marketing
   team wants to avoid false alarms. How should each set C and T?
8. How is CUSUM different from just looking at individual measurements?
:::
