---
title: "Design of Experiments: Collecting Data With Purpose"
subtitle: "A/B testing, factorial designs, and multi-armed bandits"
engine: knitr
---

## The Problem: You Don't Have Data Yet

Most of this guide assumes data already exists. DOE flips that assumption: **you need to design the data collection itself.** When getting the full dataset isn't possible or would take too long, you must decide *what* to test, *how many* tests to run, and *how* to structure the tests to get useful answers efficiently.

**When DOE matters:**

- Choosing between banner ad designs (which gets more clicks?)
- An online retailer deciding what products to suggest
- Political polls — surveying only 600 people but needing representative coverage across demographics
- Comparing medical treatments, agricultural methods, startup strategies

---

## Comparison, Control, and Blocking

Before any specific method, two foundational concepts:

### Comparison and Control
To determine whether red cars sell for more than blue cars, you must **control for other factors**. Comparing two-year-old red Porsches to ten-year-old blue family cars conflates color with age and car type. You need groups that are comparable in all dimensions except the one you're testing.

### Blocking Factors
A **blocking factor** is something that could create variation in your results. Car type (sports car vs. family car) is a blocking factor — sports cars are more likely to be red. By analyzing red sports cars separately from red family cars, you **reduce variance** in your estimates because you've accounted for one source of variation.

---

## A/B Testing

The simplest DOE method: compare **two alternatives** by randomly assigning test subjects and measuring outcomes.

### How It Works

1. Randomly show Ad A or Ad B to the first 2,000 visitors
2. Track clicks (binomial data)
3. Run a hypothesis test

```{r}
#| label: ab-test-demo
#| fig-cap: "A/B test: is the difference in click rates statistically significant?"

# Banner ad A/B test results
clicks_A <- 46;  shown_A <- 1003
clicks_B <- 97;  shown_B <- 997

# Two-proportion z-test
result <- prop.test(c(clicks_A, clicks_B), c(shown_A, shown_B))

cat("Ad A click rate:", round(clicks_A / shown_A * 100, 1), "%\n")
cat("Ad B click rate:", round(clicks_B / shown_B * 100, 1), "%\n")
cat("p-value:", format.pval(result$p.value, digits = 3), "\n")
cat("Conclusion:", ifelse(result$p.value < 0.05,
    "Statistically significant difference — use Ad B",
    "No significant difference — either ad is fine"), "\n")
```

### Three Requirements

| Requirement | Why It Matters |
|------------|----------------|
| **Enough data collected quickly** | The answer must arrive in time to be useful |
| **Representative sample** | Results from one population may not apply to another |
| **Test is small vs. total population** | If you expect 2,500 views and test 2,000, only 500 remain to benefit |

> **Key nuance:** "No significant difference" is also a useful answer — it means you can choose either alternative based on other criteria (cost, aesthetics, etc.).

---

## Factorial Designs

When you have **multiple factors** to test, not just two alternatives.

### Full Factorial
Test **every combination** of factor levels:

| Factor | Choices |
|--------|---------|
| Font | Arial, Roboto |
| Wording | "MS in Analytics", "Analytics Masters" |
| Background | Gold, White |

$2 \times 2 \times 2 = 8$ combinations. Run all 8, then use **ANOVA** (Analysis of Variance) to decompose how much each factor contributes.

### The Scaling Problem
Seven factors with three choices each: $3^7 = 2{,}187$ combinations. That's almost always too many to test.

### Fractional Factorial
Test a **carefully chosen subset** of combinations. The subset must be **balanced**:

- Each choice appears the same number of times
- Each *pair* of choices appears the same number of times

This preserves the ability to estimate main effects without testing every combination.

### Regression Approach
If factors are believed to be **independent**, use regression with categorical variables to estimate each factor's effect from the subset. But watch out for interactions:

::: {.callout-warning}
### The Interaction Trap
A white font and a white background might each test well individually — but a white font *on* a white background would be unreadable. When factors interact, you need **interaction terms** in your regression, and you may need to test those specific combinations.
:::

---

## Multi-Armed Bandits

### The Problem with Fixed Testing
With 10 alternatives tested 1,000 times each, you show the best ad only 1,000 times but show *suboptimal* ads 9,000 times. Every trial spent on a bad alternative is **lost value**.

### Exploration vs. Exploitation

| Strategy | Focus | Risk |
|----------|-------|------|
| **Exploration** | Getting more information (which is truly best?) | Wastes value on bad options |
| **Exploitation** | Using the apparent best option now | Might miss a better option |

The multi-armed bandit algorithm **balances both simultaneously**.

### The Algorithm

```{r}
#| label: bandit-simulation
#| fig-cap: "Multi-armed bandit simulation: the algorithm quickly shifts probability toward the best-performing alternative."
#| fig-width: 8
#| fig-height: 4

set.seed(42)

# True click probabilities (unknown to the algorithm)
true_probs <- c(0.04, 0.08, 0.06)  # Ad A, B, C
K <- length(true_probs)
n_rounds <- 20
batch_size <- 50

# Track results
successes <- rep(0, K)
trials <- rep(0, K)
prob_history <- matrix(NA, nrow = n_rounds, ncol = K)

for (round in 1:n_rounds) {
  # Calculate selection probabilities (Thompson sampling approximation)
  if (round == 1) {
    probs <- rep(1/K, K)
  } else {
    # Estimate probability of each being best (simplified)
    rates <- (successes + 1) / (trials + 2)  # Laplace smoothing
    probs <- rates / sum(rates)
  }
  prob_history[round, ] <- probs

  # Assign batch according to probabilities
  assignments <- sample(1:K, batch_size, replace = TRUE, prob = probs)

  # Simulate outcomes
  for (k in 1:K) {
    n_assigned <- sum(assignments == k)
    if (n_assigned > 0) {
      trials[k] <- trials[k] + n_assigned
      successes[k] <- successes[k] + rbinom(1, n_assigned, true_probs[k])
    }
  }
}

# Plot probability evolution
cols <- c("#d62828", "#2d6a4f", "#457b9d")
matplot(1:n_rounds, prob_history, type = "l", lwd = 2, lty = 1, col = cols,
        xlab = "Round", ylab = "Selection Probability",
        main = "Bandit Learns to Favor the Best Ad Over Time",
        ylim = c(0, 0.7))
legend("topright", legend = paste0("Ad ", LETTERS[1:K],
       " (true rate: ", true_probs * 100, "%)"),
       col = cols, lwd = 2, cex = 0.85, bg = "white")
abline(h = 1/K, col = "gray70", lty = 2)
text(n_rounds * 0.8, 1/K + 0.03, "Equal probability", col = "gray50", cex = 0.8)
```

### Name Origin
"One-armed bandit" = a slot machine (one lever arm, calibrated so you lose money on average). "Multi-armed bandit" = choosing among multiple slot machines with unknown payouts.

### Parameters to Tune

| Parameter | Options |
|-----------|---------|
| Batch size | How many tests between recalculations |
| Probability updates | Bayesian updates vs. observed distributions |
| Assignment rule | Based on probability of being best, or on expected value |

---

## Connecting the Methods

| Method | When to Use | Number of Alternatives | Adaptive? |
|--------|-------------|----------------------|-----------|
| **A/B Testing** | Simple comparison, two options | 2 | No (fixed test) |
| **Factorial Design** | Understand factor contributions | Many combinations | No (fixed design) |
| **Multi-Armed Bandit** | Choose best while minimizing waste | Multiple | Yes (learns as it goes) |

All three address the same core question: *how do I collect data efficiently to make a good decision?* They differ in complexity, the number of alternatives, and whether the algorithm adapts during testing.

---

## Common Pitfalls

::: {.callout-warning}
### Common Misconceptions

1. **A/B testing requires a REPRESENTATIVE sample** — results from college students may not apply to retirees. Sample must match the target population.
2. **A/B testing also requires enough REMAINING population** — if your test uses most of the population, there's nobody left to benefit from the answer.
3. **Full factorial tests ALL combinations** — don't confuse with fractional factorial, which tests a balanced subset.
4. **Fractional factorial must be BALANCED** — not any random subset of combinations works. Each choice and pair of choices must appear equally often.
5. **Multi-armed bandits beat fixed testing** — they learn faster AND earn more value during the learning process.
6. **DOE happens BEFORE data collection** — it's about designing what data to collect, not analyzing data you already have.
7. **Interaction terms can invalidate independence assumptions** — the regression approach in factorial designs assumes factors don't interact unless interaction terms are included.
:::
