---
title: "k-Means Clustering From Scratch: Finding Groups Nobody Told You About"
subtitle: "Unsupervised learning — no labels, no teacher, just patterns"
engine: knitr
---

## 1. The Problem: No Labels

Everything so far — SVM, KNN — assumed you had **labeled** data. You knew
which group each point belonged to, and the goal was to learn the rule.

But what if you have **no labels**? Nobody told you the groups. You just have
a pile of data and a question: **are there natural groupings hiding in here?**

```{r}
#| label: fig-unlabeled
#| fig-cap: "You see the data. Can you spot the groups? k-means can."

library(ggplot2)
set.seed(42)

# Three hidden clusters
true_centers <- data.frame(x1 = c(2, 6, 4), x2 = c(2, 2, 6))
n_per <- 60

df <- data.frame(
  x1 = c(rnorm(n_per, 2, 0.7), rnorm(n_per, 6, 0.7), rnorm(n_per, 4, 0.7)),
  x2 = c(rnorm(n_per, 2, 0.7), rnorm(n_per, 2, 0.7), rnorm(n_per, 6, 0.7)),
  true_group = factor(rep(c("A", "B", "C"), each = n_per))
)

# Show without labels
ggplot(df, aes(x1, x2)) +
  geom_point(size = 2, color = "gray30") +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2",
       title = "All you see: unlabeled points")
```

This is **unsupervised learning**. No response variable, no "right answer."
The algorithm has to discover structure on its own.


## 2. The k-Means Algorithm: 4 Steps on Repeat

k-means is embarrassingly simple:

1. **Pick** $k$ random starting centers (centroids)
2. **Assign** each point to the nearest center
3. **Move** each center to the average of its assigned points
4. **Repeat** steps 2-3 until nothing changes

That's the whole algorithm. Let's watch it happen.

```{r}
#| label: fig-algorithm-steps
#| fig-cap: "k-means in action: assign → move centers → assign → move → converge"

set.seed(15)

# Manual k-means to show each step
k <- 3
centers <- df[sample(nrow(df), k), c("x1", "x2")]
colors <- c("coral", "steelblue", "forestgreen")

par(mfrow = c(2, 3), mar = c(3, 3, 3, 1))

for (iter in 1:6) {
  # Step 2: Assign to nearest center
  dists <- sapply(1:k, function(j) {
    sqrt((df$x1 - centers$x1[j])^2 + (df$x2 - centers$x2[j])^2)
  })
  assignments <- apply(dists, 1, which.min)

  plot(df$x1, df$x2,
    col = adjustcolor(colors[assignments], 0.5),
    pch = 19, cex = 1,
    main = paste("Iteration", iter),
    xlab = "x1", ylab = "x2")

  # Show centers
  points(centers$x1, centers$x2, pch = 4, cex = 3, lwd = 3,
         col = colors[1:k])

  # Draw lines from centers to their points (first iteration only)
  if (iter == 1) {
    legend("topright", bty = "n", cex = 0.7,
      legend = "✕ = cluster centers")
  }

  # Step 3: Move centers
  for (j in 1:k) {
    members <- df[assignments == j, ]
    centers$x1[j] <- mean(members$x1)
    centers$x2[j] <- mean(members$x2)
  }
}
par(mfrow = c(1, 1))
```

After a few iterations, the centers stop moving — the algorithm has
**converged**. Each point belongs to the cluster whose center is closest.


## 3. The Formula: What k-Means Minimizes

k-means minimizes the **total within-cluster sum of squares (WCSS)**:

$$\min \sum_{j=1}^{k} \sum_{\mathbf{x}_i \in C_j} \|\mathbf{x}_i - \boldsymbol{\mu}_j\|^2$$

Let's unpack every piece:

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $k$ | Number of clusters | How many groups you want |
| $C_j$ | Cluster $j$ | The set of points assigned to group $j$ |
| $\mathbf{x}_i$ | A data point | One observation |
| $\boldsymbol{\mu}_j$ | Center (centroid) of cluster $j$ | The average of all points in group $j$ |
| $\|\mathbf{x}_i - \boldsymbol{\mu}_j\|^2$ | Squared distance | How far point $i$ is from its cluster center |

**In plain English:** make each point as close as possible to the center of
its assigned cluster. Tight clusters = low WCSS = good.

```{r}
#| label: fig-wcss-visual
#| fig-cap: "WCSS = sum of all these distances (squared). k-means tries to make this small."

# Final k-means result
km <- kmeans(df[, c("x1", "x2")], centers = 3, nstart = 20)

df$cluster <- factor(km$cluster)
centers_final <- as.data.frame(km$centers)
centers_final$cluster <- factor(1:3)

ggplot(df, aes(x1, x2, color = cluster)) +
  geom_segment(aes(xend = km$centers[as.numeric(cluster), 1],
                    yend = km$centers[as.numeric(cluster), 2]),
               alpha = 0.15, linewidth = 0.3) +
  geom_point(size = 2) +
  geom_point(data = centers_final, aes(x1, x2),
             shape = 4, size = 5, stroke = 3, color = "black") +
  scale_color_manual(values = c("coral", "steelblue", "forestgreen")) +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2",
       title = paste("Total WCSS =", round(km$tot.withinss, 1)),
       subtitle = "Each line = distance from point to its center") +
  theme(legend.position = "none")
```

### The Centroid Formula

The center of cluster $j$ is simply the **mean** of all points assigned to it:

$$\boldsymbol{\mu}_j = \frac{1}{|C_j|} \sum_{\mathbf{x}_i \in C_j} \mathbf{x}_i$$

That's why it's called k-**means** — the centers are the means of their
clusters.


## 4. The Assignment Rule: Nearest Center

Each point goes to the cluster whose center is closest (Euclidean distance):

$$\text{cluster}(\mathbf{x}_i) = \arg\min_j \|\mathbf{x}_i - \boldsymbol{\mu}_j\|^2$$

This is the same distance formula from KNN. The difference: KNN looks at
nearest **data points**. k-means looks at nearest **cluster centers**.

```{r}
#| label: fig-voronoi
#| fig-cap: "Each region contains all points closest to that center (Voronoi diagram)"

# Show decision regions
grid <- expand.grid(
  x1 = seq(min(df$x1) - 1, max(df$x1) + 1, length.out = 200),
  x2 = seq(min(df$x2) - 1, max(df$x2) + 1, length.out = 200)
)

dists_grid <- sapply(1:3, function(j) {
  sqrt((grid$x1 - km$centers[j, 1])^2 + (grid$x2 - km$centers[j, 2])^2)
})
grid$cluster <- factor(apply(dists_grid, 1, which.min))

ggplot() +
  geom_tile(data = grid, aes(x1, x2, fill = cluster), alpha = 0.15) +
  geom_point(data = df, aes(x1, x2, color = cluster), size = 2) +
  geom_point(data = centers_final, aes(x1, x2),
             shape = 4, size = 5, stroke = 3, color = "black") +
  scale_fill_manual(values = c("coral", "steelblue", "forestgreen")) +
  scale_color_manual(values = c("coral", "steelblue", "forestgreen")) +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2",
       title = "Cluster territories: every location maps to its nearest center") +
  theme(legend.position = "none")
```


## 5. Choosing k: The Elbow Method

The hardest part of k-means: **how many clusters?** There's no formula.
You use the **elbow method**:

1. Run k-means for $k = 1, 2, 3, \dots$
2. Record the total WCSS for each $k$
3. Plot $k$ vs WCSS
4. Look for the **elbow** — where the curve bends and adding more clusters
   stops helping much

```{r}
#| label: fig-elbow
#| fig-cap: "The elbow method: WCSS drops sharply, then flattens — the bend is your k"

k_range <- 1:10
wcss <- sapply(k_range, function(k) {
  km_fit <- kmeans(df[, c("x1", "x2")], centers = k, nstart = 20)
  km_fit$tot.withinss
})

elbow_df <- data.frame(k = k_range, wcss = wcss)

ggplot(elbow_df, aes(k, wcss)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_point(data = elbow_df[3, ], aes(k, wcss),
             size = 6, shape = 1, color = "coral", stroke = 2) +
  annotate("text", x = 4.5, y = wcss[3],
           label = "Elbow at k = 3", color = "coral", size = 4.5, fontface = "bold") +
  annotate("segment", x = 4.2, xend = 3.2, y = wcss[3], yend = wcss[3],
           arrow = arrow(length = unit(0.3, "cm")), color = "coral") +
  scale_x_continuous(breaks = k_range) +
  theme_minimal(base_size = 14) +
  labs(x = "Number of Clusters (k)", y = "Total Within-Cluster Sum of Squares",
       title = "Elbow Method: Where does adding clusters stop helping?")
```

**Why $k = 3$ here?** Going from 2 to 3 clusters gives a big drop in WCSS.
Going from 3 to 4 gives almost nothing. The "elbow" bends at 3.

**Important:** The elbow is a **judgment call**, not an exact calculation.
Sometimes the elbow is obvious; sometimes it's ambiguous. In this guide,
the setup will make it clear.


## 6. The Random Initialization Problem

k-means starts with **random** centers. Different starting positions can lead
to **different results** — sometimes bad ones.

```{r}
#| label: fig-bad-init
#| fig-cap: "Same data, same k=3, but different random starts → different results"

par(mfrow = c(1, 3), mar = c(3, 3, 3, 1))

colors <- c("coral", "steelblue", "forestgreen")

for (trial in c(1, 2, 3)) {
  set.seed(trial * 7)
  km_trial <- kmeans(df[, c("x1", "x2")], centers = 3, nstart = 1)

  plot(df$x1, df$x2,
    col = adjustcolor(colors[km_trial$cluster], 0.6),
    pch = 19, cex = 1,
    main = paste0("Start #", trial, " (WCSS = ", round(km_trial$tot.withinss, 0), ")"),
    xlab = "x1", ylab = "x2")
  points(km_trial$centers[, 1], km_trial$centers[, 2],
         pch = 4, cex = 3, lwd = 3)
}
par(mfrow = c(1, 1))
```

**The fix:** Run k-means **many times** with different random starts and keep
the result with the **lowest WCSS**.

In R: `kmeans(data, centers = k, nstart = 20)` runs it 20 times automatically.

```{r}
#| label: fig-nstart
#| fig-cap: "More random starts → more likely to find the best clustering"

nstart_vals <- c(1, 5, 10, 25, 50)
best_wcss <- sapply(nstart_vals, function(ns) {
  results <- replicate(20, {
    km_fit <- kmeans(df[, c("x1", "x2")], centers = 3, nstart = ns)
    km_fit$tot.withinss
  })
  mean(results)
})

ggplot(data.frame(nstart = nstart_vals, wcss = best_wcss), aes(nstart, wcss)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  theme_minimal(base_size = 14) +
  labs(x = "nstart (number of random starts)", y = "Average WCSS found",
       title = "More starts = better solution (diminishing returns past ~20)")
```


## 7. Why Scaling Is Mandatory (Again)

Same story as KNN and SVM. k-means uses **distance**. Unscaled features with
large ranges dominate.

```{r}
#| label: fig-scaling-kmeans
#| fig-cap: "Without scaling, the high-range feature dominates cluster assignments"

set.seed(42)

# Customer data: income (30k-90k) and age (18-70)
customers <- data.frame(
  income = c(rnorm(40, 40000, 5000), rnorm(40, 75000, 5000)),
  age = c(rnorm(40, 25, 4), rnorm(40, 55, 4))
)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Unscaled
km_raw <- kmeans(customers, 2, nstart = 20)
plot(customers$income, customers$age,
  col = c("coral", "steelblue")[km_raw$cluster],
  pch = 19, main = "Unscaled: income dominates",
  xlab = "Income ($)", ylab = "Age")

# Scaled
cust_scaled <- scale(customers)
km_scaled <- kmeans(cust_scaled, 2, nstart = 20)
plot(customers$income, customers$age,
  col = c("coral", "steelblue")[km_scaled$cluster],
  pch = 19, main = "Scaled: both features contribute",
  xlab = "Income ($)", ylab = "Age")

par(mfrow = c(1, 1))
```


## 8. Supervised vs Unsupervised: The Big Picture

This is an **high-value distinction**. Know the difference cold.

| | Supervised (SVM, KNN) | Unsupervised (k-means) |
|:--|:----------------------|:-----------------------|
| **Labels?** | Yes — you know the groups | No — groups are unknown |
| **Goal** | Learn the rule to predict labels | Discover hidden structure |
| **Evaluation** | Accuracy, cross-validation | WCSS, silhouette, elbow method |
| **k means...** | (KNN) number of neighbors | (k-means) number of clusters |
| **Example** | "Is this email spam?" | "What types of customers do we have?" |

```{r}
#| label: fig-supervised-vs-unsupervised
#| fig-cap: "Left: supervised (labels known, learn boundary). Right: unsupervised (no labels, find groups)."

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Supervised: labeled data with boundary
plot(df$x1, df$x2,
  col = c("coral", "steelblue", "forestgreen")[as.numeric(df$true_group)],
  pch = 19, cex = 1.2,
  main = "Supervised: labels given",
  xlab = "Feature 1", ylab = "Feature 2")
legend("topleft", legend = c("A", "B", "C"), bty = "n",
  col = c("coral", "steelblue", "forestgreen"), pch = 19, cex = 0.8,
  title = "Known labels")

# Unsupervised: same data, no labels, discover clusters
plot(df$x1, df$x2,
  col = c("coral", "steelblue", "forestgreen")[km$cluster],
  pch = 19, cex = 1.2,
  main = "Unsupervised: clusters discovered",
  xlab = "Feature 1", ylab = "Feature 2")
points(km$centers[, 1], km$centers[, 2], pch = 4, cex = 3, lwd = 3)
legend("topleft", legend = paste("Cluster", 1:3), bty = "n",
  col = c("coral", "steelblue", "forestgreen"), pch = 19, cex = 0.8,
  title = "Discovered groups")

par(mfrow = c(1, 1))
```


## 9. k-Means Limitations

### It assumes spherical clusters

k-means draws boundaries equidistant from centers — producing circular
(spherical) regions. Data with elongated, curved, or irregular shapes won't
cluster well.

```{r}
#| label: fig-limitations
#| fig-cap: "k-means struggles with non-spherical shapes"

set.seed(42)

# Concentric rings (k-means will fail)
n <- 200
angle <- runif(n, 0, 2 * pi)
r_inner <- rnorm(n / 2, 1.5, 0.2)
r_outer <- rnorm(n / 2, 4, 0.3)

rings <- data.frame(
  x1 = c(r_inner * cos(angle[1:(n/2)]), r_outer * cos(angle[(n/2+1):n])),
  x2 = c(r_inner * sin(angle[1:(n/2)]), r_outer * sin(angle[(n/2+1):n])),
  true = factor(rep(c("Inner", "Outer"), each = n / 2))
)

km_rings <- kmeans(rings[, 1:2], 2, nstart = 20)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(rings$x1, rings$x2,
  col = c("coral", "steelblue")[as.numeric(rings$true)],
  pch = 19, main = "True groups (rings)",
  xlab = "x1", ylab = "x2", asp = 1)

plot(rings$x1, rings$x2,
  col = c("coral", "steelblue")[km_rings$cluster],
  pch = 19, main = "k-means result (wrong!)",
  xlab = "x1", ylab = "x2", asp = 1)
points(km_rings$centers[, 1], km_rings$centers[, 2],
       pch = 4, cex = 3, lwd = 3)

par(mfrow = c(1, 1))
```

### Other limitations

- **You must choose k in advance** — the algorithm can't decide for you
- **Sensitive to outliers** — a single extreme point can pull a center away
- **Random starts** — can get stuck in bad solutions (fix: use `nstart`)
- **Only finds convex clusters** — can't wrap around shapes


## 10. Practical Use: Customer Segmentation

This is the classic practice scenario. You have customer data, no labels, and the
business wants to know: **what types of customers do we have?**

```{r}
#| label: fig-segmentation
#| fig-cap: "Customer segmentation: k-means discovers 3 natural groups from spending data"

set.seed(42)

# Simulated customer data
customers2 <- data.frame(
  annual_spend = c(rnorm(50, 200, 40), rnorm(50, 600, 80), rnorm(50, 1200, 150)),
  visit_frequency = c(rnorm(50, 4, 1), rnorm(50, 12, 2), rnorm(50, 25, 3))
)

# Scale then cluster
cust_s <- scale(customers2)
km_cust <- kmeans(cust_s, 3, nstart = 20)

customers2$segment <- factor(km_cust$cluster)

# Label segments meaningfully
seg_means <- aggregate(cbind(annual_spend, visit_frequency) ~ segment, data = customers2, mean)
seg_means <- seg_means[order(seg_means$annual_spend), ]
seg_labels <- c("Budget Shoppers", "Regular Customers", "VIP / High-Value")

ggplot(customers2, aes(annual_spend, visit_frequency, color = segment)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("coral", "steelblue", "forestgreen"),
                     labels = seg_labels) +
  theme_minimal(base_size = 14) +
  labs(x = "Annual Spend ($)", y = "Visit Frequency (per month)",
       color = "Segment",
       title = "k-means segmentation: 3 customer groups discovered")
```

**The workflow:**

1. Collect customer features (spend, visits, demographics, etc.)
2. Scale the features
3. Use elbow method to choose $k$
4. Run k-means with `nstart = 20`
5. Interpret the clusters → give them business-meaningful names
6. Build different strategies per segment


## 11. Cheat Sheet: The Whole Story on One Page

```
THE k-MEANS RECIPE
===================

1. ALGORITHM (repeat until convergence):
   - Assign each point to nearest center
   - Move each center to mean of its points

2. THE FORMULA:

   min  Σⱼ Σᵢ∈Cⱼ ||xᵢ - μⱼ||²
        -------------------------
        "total distance from points
         to their cluster centers"

   Center of cluster j:  μⱼ = (1/|Cⱼ|) × Σ xᵢ

3. KEY PARAMETER:
   k → number of clusters (choose via elbow method)

4. ELBOW METHOD:
   Plot k vs WCSS → pick the "bend" point

5. GOTCHAS:
   - Random initialization → use nstart = 20
   - ALWAYS scale features first
   - Assumes spherical clusters
   - You must choose k in advance

6. SUPERVISED vs UNSUPERVISED:
   Labels → supervised (SVM, KNN)
   No labels → unsupervised (k-means)

7. PRACTICE TIP:
   "Segment customers" / "find groups" / "no labels" → k-means
   "Classify" / "predict category" / "labeled data" → SVM, KNN
```


## 12. Check Your Understanding

::: {.callout-note}
### Test Yourself

Before moving on, try to answer these without scrolling up:

1. What makes k-means **unsupervised**? How is this different from KNN?
2. Describe the k-means algorithm in 4 steps.
3. What does k-means minimize? Write the formula in words.
4. Why is the centroid the **mean** of the cluster points?
5. How does the elbow method work? What are you plotting?
6. Why do you need to run k-means with multiple random starts?
7. Give a scenario where k-means would fail. Why does it fail?
8. A question says "we have customer purchase data and want to identify
   different customer types." Which model? Why not SVM?
:::
