---
title: "KNN From Scratch: Unpacking the Formula"
subtitle: "The simplest idea in machine learning — and why it works"
engine: knitr
---

## 1. The Problem: Classify a New Point

You have labeled data — points that already belong to known groups. A new,
unlabeled point shows up. Which group does it belong to?

```{r}
#| label: fig-mystery-point
#| fig-cap: "A new point (black diamond) arrives. Which group does it belong to?"

library(ggplot2)
set.seed(42)

df <- data.frame(
  x1 = c(rnorm(30, 2, 0.7), rnorm(30, 5, 0.7)),
  x2 = c(rnorm(30, 2, 0.7), rnorm(30, 5, 0.7)),
  group = factor(rep(c("Red", "Blue"), each = 30))
)

new_point <- data.frame(x1 = 3.8, x2 = 3.5)

ggplot(df, aes(x1, x2, color = group)) +
  geom_point(size = 3) +
  geom_point(data = new_point, aes(x1, x2), color = "black",
             shape = 18, size = 6, inherit.aes = FALSE) +
  scale_color_manual(values = c("steelblue", "coral")) +
  annotate("text", x = 4.1, y = 3.2, label = "???", fontface = "bold", size = 5) +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2")
```

KNN's answer is absurdly simple: **look at the closest points and go with the
majority vote.**

That's it. That's the whole algorithm.


## 2. The Algorithm in Plain English

1. Pick a number $k$ (e.g., 5)
2. Measure the distance from the new point to **every** training point
3. Find the $k$ closest neighbors
4. Count how many belong to each class
5. The new point gets the class with the most votes

```{r}
#| label: fig-knn-step-by-step
#| fig-cap: "k=5: find 5 nearest neighbors, take majority vote"

# Calculate distances from new point to all training points
df$dist <- sqrt((df$x1 - new_point$x1)^2 + (df$x2 - new_point$x2)^2)
df$rank <- rank(df$dist)
df$is_neighbor <- df$rank <= 5

# Count votes
neighbors <- df[df$is_neighbor, ]
vote_red <- sum(neighbors$group == "Red")
vote_blue <- sum(neighbors$group == "Blue")

ggplot(df, aes(x1, x2, color = group)) +
  geom_point(size = 3, alpha = 0.3) +
  # Highlight neighbors
  geom_point(data = neighbors, aes(x1, x2, color = group), size = 4) +
  # Draw lines from new point to neighbors
  geom_segment(data = neighbors,
    aes(x = new_point$x1, y = new_point$x2, xend = x1, yend = x2),
    color = "gray50", linetype = "dashed", inherit.aes = FALSE) +
  # New point
  geom_point(data = new_point, aes(x1, x2), color = "black",
             shape = 18, size = 6, inherit.aes = FALSE) +
  # Circle around neighborhood
  annotate("path",
    x = new_point$x1 + max(neighbors$dist) * cos(seq(0, 2*pi, length.out = 100)),
    y = new_point$x2 + max(neighbors$dist) * sin(seq(0, 2*pi, length.out = 100)),
    linetype = "dotted", color = "gray40") +
  scale_color_manual(values = c("steelblue", "coral")) +
  annotate("label", x = 5.5, y = 1.5, size = 4,
    label = paste0("Vote: ", vote_red, " Red, ", vote_blue, " Blue\n",
                   "Winner: ", ifelse(vote_red > vote_blue, "Red", "Blue"))) +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2")
```


## 3. The Formula: Distance

The only math KNN needs is a way to measure **how far apart two points are**.

### Euclidean Distance (the default)

$$d(\mathbf{x}, \mathbf{x}') = \sqrt{\sum_{j=1}^{p} (x_j - x_j')^2}$$

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $\mathbf{x}$ | The new point | The thing we're classifying |
| $\mathbf{x}'$ | A training point | One of the known, labeled examples |
| $p$ | Number of features | How many measurements each point has |
| $x_j$ | Feature $j$ of the new point | e.g., income |
| $x_j'$ | Feature $j$ of the training point | e.g., that person's income |

In 2D, this is just the Pythagorean theorem:

$$d = \sqrt{(x_1 - x_1')^2 + (x_2 - x_2')^2}$$

```{r}
#| label: fig-distance-visual
#| fig-cap: "Euclidean distance = straight-line distance between two points"

p1 <- data.frame(x1 = 2, x2 = 1)
p2 <- data.frame(x1 = 5, x2 = 5)
d <- sqrt((p2$x1 - p1$x1)^2 + (p2$x2 - p1$x2)^2)

ggplot() +
  # The two points
  geom_point(data = p1, aes(x1, x2), size = 5, color = "coral") +
  geom_point(data = p2, aes(x1, x2), size = 5, color = "steelblue") +
  # Direct line (hypotenuse)
  geom_segment(aes(x = p1$x1, y = p1$x2, xend = p2$x1, yend = p2$x2),
               linewidth = 1.2) +
  # Horizontal leg
  geom_segment(aes(x = p1$x1, y = p1$x2, xend = p2$x1, yend = p1$x2),
               linetype = "dashed", color = "gray50") +
  # Vertical leg
  geom_segment(aes(x = p2$x1, y = p1$x2, xend = p2$x1, yend = p2$x2),
               linetype = "dashed", color = "gray50") +
  # Labels
  annotate("text", x = 3.5, y = 0.6, label = paste0("Δx₁ = ", p2$x1 - p1$x1), size = 4) +
  annotate("text", x = 5.5, y = 3, label = paste0("Δx₂ = ", p2$x2 - p1$x2), size = 4) +
  annotate("text", x = 3, y = 3.8,
    label = paste0("d = √(3² + 4²) = ", round(d, 1)), size = 4.5, fontface = "bold") +
  coord_equal() +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2")
```

### Other Distance Metrics

| Metric | Formula | When to use |
|:------:|:--------|:------------|
| **Euclidean** | $\sqrt{\sum(x_j - x_j')^2}$ | Continuous features (default) |
| **Manhattan** | $\sum |x_j - x_j'|$ | Grid-like data, robust to outliers |
| **Hamming** | Count of differing features | Categorical/binary features |

For this guide, Euclidean is the default unless told otherwise.


## 4. The Classification Rule

Once you have the $k$ nearest neighbors, the predicted class is:

$$\hat{y} = \text{mode}(\{y_i : \mathbf{x}_i \in N_k(\mathbf{x})\})$$

In English: **the most common class among the k neighbors wins.**

| Symbol | What it is |
|:------:|:-----------|
| $\hat{y}$ | Predicted class for the new point |
| $y_i$ | Known class of training point $i$ |
| $N_k(\mathbf{x})$ | The set of $k$ nearest neighbors of $\mathbf{x}$ |
| mode | The most frequent value |

What about ties? If $k$ is even, ties are possible. Common fixes:

- Use odd $k$
- Weight votes by distance (closer neighbors count more)
- Break ties randomly

```{r}
#| label: fig-voting
#| fig-cap: "Different k values lead to different votes"

library(class)

# Show votes for k = 1, 3, 7
k_vals <- c(1, 3, 7)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

for (k in k_vals) {
  # Find neighbors
  nn <- df[order(df$dist)[1:k], ]
  vote_r <- sum(nn$group == "Red")
  vote_b <- sum(nn$group == "Blue")
  winner <- ifelse(vote_r > vote_b, "Red", "Blue")

  plot(df$x1, df$x2,
    col = ifelse(df$group == "Red", "coral", "steelblue"),
    pch = 19, cex = 1, main = paste0("k = ", k),
    xlab = "Feature 1", ylab = "Feature 2")

  # Circle neighbors
  points(nn$x1, nn$x2,
    col = ifelse(nn$group == "Red", "coral", "steelblue"),
    pch = 19, cex = 2)
  points(nn$x1, nn$x2, pch = 1, cex = 2.5, lwd = 2)

  # New point
  points(new_point$x1, new_point$x2, pch = 18, cex = 3, col = "black")

  # Draw radius
  radius <- max(nn$dist)
  theta <- seq(0, 2 * pi, length.out = 100)
  lines(new_point$x1 + radius * cos(theta),
        new_point$x2 + radius * sin(theta),
        lty = 3, col = "gray40")

  legend("topleft", bty = "n", cex = 0.9,
    legend = c(
      paste0("Red: ", vote_r, "  Blue: ", vote_b),
      paste0("→ ", winner)
    ))
}
par(mfrow = c(1, 1))
```


## 5. The Key Parameter: Choosing k

$k$ is the **only** parameter in KNN. It controls the bias-variance tradeoff,
just like C does in SVM.

| Small k (e.g., 1) | Large k (e.g., 50) |
|:-------------------|:--------------------|
| Jagged, complex boundary | Smooth, simple boundary |
| Memorizes training data | Averages over many points |
| Low bias, high variance | High bias, low variance |
| **Overfitting risk** | **Underfitting risk** |

```{r}
#| label: fig-k-boundaries
#| fig-cap: "Small k = jagged boundary (overfit). Large k = smooth boundary (underfit)."

# Full decision boundary comparison
grid <- expand.grid(
  x1 = seq(min(df$x1) - 0.5, max(df$x1) + 0.5, length.out = 150),
  x2 = seq(min(df$x2) - 0.5, max(df$x2) + 0.5, length.out = 150)
)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

for (k in c(1, 5, 25)) {
  pred <- knn(
    train = df[, c("x1", "x2")],
    test = grid[, c("x1", "x2")],
    cl = df$group,
    k = k
  )

  plot(df$x1, df$x2,
    col = ifelse(df$group == "Red", "coral", "steelblue"),
    pch = 19, cex = 1.2,
    main = paste("k =", k),
    xlab = "Feature 1", ylab = "Feature 2")

  points(grid$x1, grid$x2,
    col = ifelse(pred == "Red",
      adjustcolor("coral", 0.15),
      adjustcolor("steelblue", 0.15)),
    pch = 15, cex = 0.3)

  legend("topleft", bty = "n", cex = 0.9,
    legend = ifelse(k == 1, "Memorizes data (overfit)",
      ifelse(k >= 25, "Too smooth (underfit)", "Balanced")))
}
par(mfrow = c(1, 1))
```

### How to Choose k

You **cannot** calculate the best $k$ from a formula. You must **try
different values and see which performs best** on held-out data.

This is **cross-validation** (Module 3) — the standard way to tune any
hyperparameter.

**Rules of thumb:**

- Start with $k = \sqrt{n}$ where $n$ is the number of training points
- Use odd $k$ for binary classification (avoids ties)
- Typical range: 3 to 20


## 6. Why Scaling Is Mandatory

KNN is entirely based on distance. If one feature has a range of 0–100,000
and another has a range of 0–1, the large-range feature **completely
dominates** the distance calculation.

```{r}
#| label: fig-scaling
#| fig-cap: "Without scaling, income dominates. With scaling, both features contribute equally."

set.seed(7)

# Income: 30k-90k range. Credit score: 300-850 range.
df_raw <- data.frame(
  income = c(rnorm(40, 50000, 8000), rnorm(40, 70000, 8000)),
  credit = c(rnorm(40, 580, 40), rnorm(40, 720, 40)),
  group = factor(rep(c("Deny", "Approve"), each = 40))
)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Unscaled
plot(df_raw$income, df_raw$credit,
  col = ifelse(df_raw$group == "Deny", "coral", "steelblue"),
  pch = 19, main = "Unscaled",
  xlab = "Income ($)", ylab = "Credit Score")
mtext("Income range: 60,000\nCredit range: 500\nIncome dominates!",
      side = 1, line = -3, cex = 0.8, col = "red")

# Scaled
df_scaled <- df_raw
df_scaled$income <- scale(df_scaled$income)
df_scaled$credit <- scale(df_scaled$credit)

plot(df_scaled$income, df_scaled$credit,
  col = ifelse(df_scaled$group == "Deny", "coral", "steelblue"),
  pch = 19, main = "Scaled (standardized)",
  xlab = "Income (z-score)", ylab = "Credit Score (z-score)")
mtext("Both features: mean=0, sd=1\nEqual contribution to distance",
      side = 1, line = -3, cex = 0.8, col = "darkgreen")

par(mfrow = c(1, 1))
```

**Standardization** (z-score scaling):

$$x_j^{\text{scaled}} = \frac{x_j - \bar{x}_j}{s_j}$$

Each feature gets mean = 0 and standard deviation = 1. Now all features
contribute equally to distance.

```{r}
#| label: fig-scaling-accuracy
#| fig-cap: "Scaling dramatically improves KNN accuracy when features have different ranges"

# Compare accuracy with and without scaling
acc_unscaled <- mean(
  knn(df_raw[, c("income", "credit")], df_raw[, c("income", "credit")],
      df_raw$group, k = 5) == df_raw$group
)

acc_scaled <- mean(
  knn(df_scaled[, c("income", "credit")], df_scaled[, c("income", "credit")],
      df_scaled$group, k = 5) == df_scaled$group
)

barplot(
  c(Unscaled = acc_unscaled, Scaled = acc_scaled) * 100,
  col = c("coral", "steelblue"),
  ylab = "Accuracy (%)", ylim = c(0, 100),
  main = "KNN (k=5) Accuracy: Scaled vs Unscaled"
)
abline(h = 50, lty = 2, col = "gray50")
text(0.7, 55, "Random guessing", col = "gray50", cex = 0.8)
```


## 7. KNN for Regression (Not Just Classification)

KNN can also predict **continuous values**, not just classes.

Instead of majority vote, take the **average** of the neighbors' values:

$$\hat{y} = \frac{1}{k} \sum_{i \in N_k(\mathbf{x})} y_i$$

```{r}
#| label: fig-knn-regression
#| fig-cap: "KNN regression: predict by averaging the k nearest neighbors' values"

set.seed(12)
x <- sort(runif(50, 0, 10))
y <- sin(x) + rnorm(50, 0, 0.3)
df_reg <- data.frame(x = x, y = y)

# Predict for a grid of x values
x_grid <- seq(0, 10, length.out = 200)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

for (k in c(1, 5, 15)) {
  y_pred <- sapply(x_grid, function(xnew) {
    dists <- abs(df_reg$x - xnew)
    nn_idx <- order(dists)[1:k]
    mean(df_reg$y[nn_idx])
  })

  plot(df_reg$x, df_reg$y, pch = 19, col = "gray40",
    main = paste("KNN Regression: k =", k),
    xlab = "x", ylab = "y")
  lines(x_grid, y_pred, col = "steelblue", lwd = 2)
  lines(x_grid, sin(x_grid), col = "coral", lty = 2, lwd = 1.5)
  legend("topright", bty = "n", cex = 0.8,
    legend = c("KNN prediction", "True function"),
    col = c("steelblue", "coral"), lty = c(1, 2), lwd = 2)
}
par(mfrow = c(1, 1))
```

Same tradeoff: small $k$ is noisy (overfits), large $k$ is smooth (underfits).


## 8. KNN vs SVM: When to Use Which

| | KNN | SVM |
|:--|:----|:----|
| **Training** | None — just stores data | Solves optimization to find boundary |
| **Prediction** | Slow — computes distance to every point | Fast — just evaluate $\mathbf{w} \cdot \mathbf{x} + b$ |
| **Memory** | Stores entire training set | Stores only support vectors |
| **Multi-class** | Natural (majority vote) | Needs workarounds (one-vs-one) |
| **Interpretability** | Low (no model, just neighbors) | Moderate (weight vector shows feature importance) |
| **Non-linear** | Inherently non-linear | Needs kernel trick |
| **Best for** | Small datasets, quick prototyping | Large datasets, clear margin between classes |

**Practice tip:** If the scenario mentions "no training phase" or "lazy learner"
→ KNN. If it mentions "maximum margin" or "support vectors" → SVM.

```{r}
#| label: fig-svm-vs-knn
#| fig-cap: "Same data, different approaches: SVM draws a boundary, KNN uses local voting"

library(e1071)

# Slightly overlapping data
set.seed(55)
df_cmp <- data.frame(
  x1 = c(rnorm(50, 2, 1), rnorm(50, 4.5, 1)),
  x2 = c(rnorm(50, 2, 1), rnorm(50, 4.5, 1)),
  group = factor(rep(c("A", "B"), each = 50))
)

grid_cmp <- expand.grid(
  x1 = seq(-1, 8, length.out = 150),
  x2 = seq(-1, 8, length.out = 150)
)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# SVM
fit_svm <- svm(group ~ x1 + x2, data = df_cmp, kernel = "linear", cost = 1)
grid_cmp$svm_pred <- predict(fit_svm, grid_cmp)

plot(df_cmp$x1, df_cmp$x2,
  col = ifelse(df_cmp$group == "A", "steelblue", "coral"),
  pch = 19, cex = 1.2, main = "SVM (linear)",
  xlab = "Feature 1", ylab = "Feature 2")
points(grid_cmp$x1, grid_cmp$x2,
  col = ifelse(grid_cmp$svm_pred == "A",
    adjustcolor("steelblue", 0.15), adjustcolor("coral", 0.15)),
  pch = 15, cex = 0.3)
points(df_cmp$x1[fit_svm$index], df_cmp$x2[fit_svm$index],
  pch = 1, cex = 2.5, lwd = 2)
legend("topleft", bty = "n", legend = "○ = support vectors", cex = 0.8)

# KNN
knn_pred <- knn(
  train = df_cmp[, c("x1", "x2")],
  test = grid_cmp[, c("x1", "x2")],
  cl = df_cmp$group, k = 5
)

plot(df_cmp$x1, df_cmp$x2,
  col = ifelse(df_cmp$group == "A", "steelblue", "coral"),
  pch = 19, cex = 1.2, main = "KNN (k = 5)",
  xlab = "Feature 1", ylab = "Feature 2")
points(grid_cmp$x1, grid_cmp$x2,
  col = ifelse(knn_pred == "A",
    adjustcolor("steelblue", 0.15), adjustcolor("coral", 0.15)),
  pch = 15, cex = 0.3)
legend("topleft", bty = "n", legend = "No explicit boundary", cex = 0.8)

par(mfrow = c(1, 1))
```


## 9. Strengths and Weaknesses

### Strengths
- **Dead simple** — no training, no assumptions about data shape
- **Non-linear for free** — the boundary naturally curves around data
- **Multi-class for free** — just count votes
- **No training time** — just store the data

### Weaknesses (The Curse of Dimensionality)

As the number of features grows, distances become **meaningless**. In high
dimensions, all points are roughly the same distance from each other.

```{r}
#| label: fig-curse
#| fig-cap: "In high dimensions, the gap between nearest and farthest neighbor shrinks"

set.seed(1)
dims <- c(2, 5, 10, 50, 100, 500)
ratios <- sapply(dims, function(d) {
  data <- matrix(rnorm(1000 * d), ncol = d)
  query <- rnorm(d)
  dists <- sqrt(rowSums((data - matrix(query, nrow = 1000, ncol = d, byrow = TRUE))^2))
  max(dists) / min(dists)
})

barplot(ratios, names.arg = dims,
  col = "steelblue", border = NA,
  xlab = "Number of Features",
  ylab = "Max Distance / Min Distance",
  main = "Curse of Dimensionality")
abline(h = 1, lty = 2, col = "red")
text(6, 1.3, "ratio → 1 means all distances are similar\n(KNN can't distinguish neighbors)",
     col = "red", cex = 0.8)
```

**Fix:** Use PCA (Module 9) to reduce dimensions before applying KNN.


## 10. Cheat Sheet: The Whole Story on One Page

```
THE KNN RECIPE
==============

1. ALGORITHM:
   - Compute distance from new point to all training points
   - Find k closest neighbors
   - Classification: majority vote
   - Regression: average of neighbors' values

2. THE FORMULA:

   Euclidean distance:  d = √(Σ(xⱼ - xⱼ')²)
   Prediction:          ŷ = mode of k nearest labels

3. KEY PARAMETER:
   k → number of neighbors  (tune via cross-validation)
     small k = overfit (jagged boundary, memorizes noise)
     large k = underfit (smooth boundary, misses patterns)
     Rule of thumb: start with k = √n

4. MANDATORY: SCALE YOUR FEATURES
   Same reason as SVM — distance-based method

5. STRENGTHS:
   - No training needed (lazy learner)
   - Handles non-linear boundaries naturally
   - Multi-class without modification

6. WEAKNESSES:
   - Slow prediction (computes all distances)
   - Stores entire training set
   - Curse of dimensionality (fails with many features)
   - No model to interpret
```


## 11. Check Your Understanding

::: {.callout-note}
### Test Yourself

Before moving on, try to answer these without scrolling up:

1. In one sentence, what does KNN do?
2. Why is KNN called a "lazy learner"?
3. What happens to the decision boundary when you increase $k$? Decrease it?
4. Why is scaling mandatory for KNN?
5. What is the curse of dimensionality and how does it affect KNN?
6. When would you pick KNN over SVM? When would you pick SVM over KNN?
7. How do you choose the best $k$? (Hint: not a formula)
:::
