---
title: "Optimization: Variables, Constraints, and the Art of Modeling"
subtitle: "Linear programming, binary variables, and why the modeling matters more than the solver"
engine: knitr
---

## The Prescriptive Turn

Most of this guide so far has answered *what will happen?* Optimization answers *what should we do?* Given a goal, some constraints, and a set of decisions we control, what combination of decisions is best?

Two uses of optimization to remember:

1. **Top-level prescriptive analytics** — what should we invest in, how should we schedule workers, which route should the delivery truck take?
2. **Underneath almost every model in this guide** — linear regression, LASSO, ridge, logistic regression, SVM, exponential smoothing, ARIMA, GARCH, k-means — all are optimization problems being solved to pick parameters.

::: {.callout-tip}
**The big insight:** software can *solve* optimization problems, but software cannot *formulate* them for you. The modeling is the hard part — and that makes it a competitive skill.
:::

---

## The Three Pieces of Every Optimization Model

| Piece | What it represents |
|---|---|
| **Variables** | The decisions the solver picks values for |
| **Constraints** | The rules those decisions must obey |
| **Objective function** | The single number we're maximizing or minimizing |

A **solution** = a value for every variable. A **feasible solution** satisfies every constraint. The **optimal solution** is the feasible solution with the best objective value.

::: {.callout-warning}
**A useful language habit:** never say "more optimal" or "most optimal." It's like saying "more best." A solution is either optimal or it isn't.
:::

---

## A Tiny LP: The Diet Problem

The original problem that kicked off linear programming in the 1940s: feed soldiers at minimum cost subject to nutritional constraints. Here's a two-food, two-nutrient version.

**Decision**: how much of each food to serve.
- $x_1$ = servings of bread
- $x_2$ = servings of cheese

**Constraints**: need at least 8 g of protein and 10 g of carbs per meal.
- Bread: 2 g protein, 4 g carbs per serving
- Cheese: 5 g protein, 1 g carbs per serving

**Objective**: minimize cost.
- Bread: $0.20 / serving
- Cheese: $0.50 / serving

$$
\begin{aligned}
\text{minimize} \quad & 0.20 x_1 + 0.50 x_2 \\
\text{subject to} \quad & 2 x_1 + 5 x_2 \ge 8 \quad (\text{protein}) \\
& 4 x_1 + 1 x_2 \ge 10 \quad (\text{carbs}) \\
& x_1, x_2 \ge 0
\end{aligned}
$$

```{r}
#| label: lp-solve
#| echo: true

library(lpSolve)

# Objective: minimize 0.20 x1 + 0.50 x2
objective <- c(0.20, 0.50)

# Constraints: LHS coefficients, directions, RHS
constraints <- matrix(c(
  2, 5,   # protein
  4, 1    # carbs
), nrow = 2, byrow = TRUE)

directions <- c(">=", ">=")
rhs        <- c(8, 10)

sol <- lp(direction    = "min",
          objective.in = objective,
          const.mat    = constraints,
          const.dir    = directions,
          const.rhs    = rhs)

cat("Optimal cost:       $", round(sol$objval, 3), "\n")
cat("Bread servings x1:  ", round(sol$solution[1], 3), "\n")
cat("Cheese servings x2: ", round(sol$solution[2], 3), "\n")
```

### The Feasible Region — Seeing Why the Optimum Is at a Corner

For two-variable LPs, we can draw the feasible region.

```{r}
#| label: fig-feasible
#| fig-cap: "Feasible region for the diet LP. Each constraint cuts off a half-plane; the feasible set is their intersection. The optimum of a linear objective lives at a vertex of this region."
#| fig-width: 7
#| fig-height: 6
#| echo: false

par(mar = c(4, 4, 3, 1))
plot(NULL, xlim = c(0, 6), ylim = c(0, 6),
     xlab = expression(x[1] * " (bread)"),
     ylab = expression(x[2] * " (cheese)"),
     main = "Feasible region for the diet LP")
abline(h = 0, v = 0, col = "gray85")
grid()

# Protein: 2 x1 + 5 x2 = 8  ->  x2 = (8 - 2 x1) / 5
x1_seq <- seq(0, 6, length.out = 200)
lines(x1_seq, (8 - 2 * x1_seq) / 5, col = "#d62828", lwd = 2)
text(5.3, (8 - 2 * 5.3) / 5 - 0.25, "protein", col = "#d62828", cex = 0.9)

# Carbs:   4 x1 + 1 x2 = 10 ->  x2 = 10 - 4 x1
lines(x1_seq, 10 - 4 * x1_seq, col = "#1f4e79", lwd = 2)
text(1.1, 10 - 4 * 1.1 + 0.3, "carbs", col = "#1f4e79", cex = 0.9)

# Shade feasible region (above both lines and in the positive quadrant)
grid_points <- expand.grid(x1 = seq(0, 6, 0.05), x2 = seq(0, 6, 0.05))
feasible <- with(grid_points,
                 2 * x1 + 5 * x2 >= 8 &
                 4 * x1 + 1 * x2 >= 10 &
                 x1 >= 0 & x2 >= 0)
points(grid_points[feasible, ], col = adjustcolor("#2d6a4f", 0.15),
       pch = 15, cex = 0.4)

# Mark the optimal solution
points(sol$solution[1], sol$solution[2], pch = 19, col = "#6c00a3", cex = 1.8)
text(sol$solution[1] + 0.2, sol$solution[2] + 0.3,
     sprintf("optimum\n(%.2f, %.2f)", sol$solution[1], sol$solution[2]),
     col = "#6c00a3", cex = 0.85, adj = 0)

# Show one iso-cost line at the optimum for the direction
iso_level <- sol$objval
lines(x1_seq, (iso_level - 0.20 * x1_seq) / 0.50,
      col = "gray40", lty = 3)
text(4.8, (iso_level - 0.20 * 4.8) / 0.50 + 0.2,
     "iso-cost line", col = "gray40", cex = 0.8)

legend("topright",
       legend = c("protein constraint", "carbs constraint",
                  "feasible region", "optimum"),
       col = c("#d62828", "#1f4e79",
               adjustcolor("#2d6a4f", 0.4), "#6c00a3"),
       lty = c(1, 1, NA, NA), pch = c(NA, NA, 15, 19),
       lwd = c(2, 2, NA, NA), bty = "n", cex = 0.85)
```

::: {.callout-note}
**Check your understanding:** Why does the optimum of an LP always sit at a vertex of the feasible region?

*Answer: because the objective is linear. Moving along any edge changes the objective monotonically — you can always improve by going to a vertex. That's the entire idea behind the simplex algorithm, which walks from vertex to vertex until no neighbor improves the objective.*
:::

---

## The Variable-Choice Trap: Call Center Scheduling

Picking the right variables matters more than the algebra. The canonical scheduling warning is a call center that needs $d_i$ workers on day $i$ of the week and wants everyone to work 5 consecutive days before 2 off.

**"Obvious" variables** — $x_i$ = number of workers on day $i$:
- Objective ($\sum x_i$) and demand ($x_i \ge d_i$) are trivial to write.
- The 5-consecutive-days rule is *nearly impossible* to express.

**Correct variables** — $x_i$ = number of workers who *start* their 5-day stretch on day $i$:
- Objective becomes $5 \sum x_i$.
- Tuesday demand is $x_{\text{Fri}} + x_{\text{Sat}} + x_{\text{Sun}} + x_{\text{Mon}} + x_{\text{Tue}} \ge d_{\text{Tue}}$.
- The 5-in-a-row rule is *built into the variable definition* — zero extra constraints needed.

Let's solve it.

```{r}
#| label: call-center
#| echo: true

days <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

# Demand forecast
demand <- c(Mon = 17, Tue = 13, Wed = 15, Thu = 19, Fri = 14, Sat = 16, Sun = 11)

# Variables: x[i] = workers starting their 5-day stretch on day i
# Each x[i] covers 5 consecutive days starting at i (wrapping around the week)

# Build the coverage matrix: rows = days of demand, cols = start days
n_days <- 7
coverage <- matrix(0, nrow = n_days, ncol = n_days)
for (start in 1:n_days) {
  covered <- ((start - 1) + 0:4) %% n_days + 1
  coverage[covered, start] <- 1
}
dimnames(coverage) <- list(day_of_demand = days, start_day = days)
coverage

# Solve: minimize sum of workers hired, subject to coverage >= demand
lp_sched <- lp(
  direction    = "min",
  objective.in = rep(1, n_days),
  const.mat    = coverage,
  const.dir    = rep(">=", n_days),
  const.rhs    = demand,
  all.int      = TRUE
)

cat("Total workers needed:", lp_sched$objval, "\n\n")
cat("Workers starting each day:\n")
schedule <- setNames(lp_sched$solution, days)
print(schedule)

cat("\nCoverage per day (must be >= demand):\n")
coverage_actual <- as.vector(coverage %*% lp_sched$solution)
print(setNames(coverage_actual, days))
```

::: {.callout-warning}
**Common pitfall:** optimization is as much about *problem formulation* as about solving. The same problem, written with the wrong variables, can become intractable. Whenever a constraint feels awkward to write, consider whether a different variable definition would make it disappear.
:::

---

## Binary Variables: Fixed Charges and Logical Rules

Introducing $y_i \in \{0, 1\}$ (1 if we invest in stock $i$, 0 otherwise) lets us model *real* constraints that are impossible with continuous variables alone.

### Fixed charges (transaction fees)

If buying any amount of a stock costs a flat transaction fee $t$, add $\sum_i t \cdot y_i$ to the objective, and link: $x_i \le B \cdot y_i$ (if not buying, must be zero; if buying, up to budget $B$).

### Minimum investment size

$x_i \ge m_i \cdot y_i$ — can't put $3 into Google because one share is hundreds.

### Logical rules

| Rule | Constraint |
|---|---|
| Must invest in Tesla | $y_{\text{TSLA}} = 1$ |
| Invest in at least one of {A, G, AP} | $y_A + y_G + y_{AP} \ge 1$ |
| Exactly one of them | $y_A + y_G + y_{AP} = 1$ |
| FedEx and UPS together or neither | $y_{\text{FDX}} = y_{\text{UPS}}$ |
| Coke and Pepsi opposite decisions | $y_{\text{KO}} = 1 - y_{\text{PEP}}$ |

Let's solve a small portfolio problem that uses some of these.

```{r}
#| label: portfolio-binary
#| echo: true

stocks <- c("TSLA", "AMZN", "GOOG", "AAPL", "FDX", "UPS")
returns <- c(0.18, 0.15, 0.12, 0.14, 0.09, 0.08)  # expected return (decimal)
budget <- 100000
min_invest <- 5000      # can't invest less than $5k per stock
txn_fee <- 150           # fixed fee per stock we pick

# Decision variables: x1..x6 (dollars in each) then y1..y6 (binary pick)
# Maximize: sum(returns * x) - sum(txn_fee * y)
# Equivalently: minimize -(sum(returns * x) - sum(txn_fee * y))
n <- length(stocks)
objective <- c(-returns, rep(txn_fee, n))  # minimize

# Constraints:
# 1. Budget:              sum(x_i) <= budget
# 2. Linking (upper):     x_i - budget * y_i <= 0    for each i
# 3. Linking (lower min): x_i - min_invest * y_i >= 0 for each i
# 4. At least one of {AMZN, GOOG, AAPL}: y_AMZN + y_GOOG + y_AAPL >= 1
# 5. FedEx/UPS both or neither: y_FDX - y_UPS = 0

# Budget row
A <- rbind(c(rep(1, n), rep(0, n)))
dir <- c("<=")
rhs <- c(budget)

# Linking constraints
for (i in 1:n) {
  row_upper <- rep(0, 2 * n)
  row_upper[i] <- 1
  row_upper[n + i] <- -budget
  A <- rbind(A, row_upper); dir <- c(dir, "<="); rhs <- c(rhs, 0)

  row_lower <- rep(0, 2 * n)
  row_lower[i] <- 1
  row_lower[n + i] <- -min_invest
  A <- rbind(A, row_lower); dir <- c(dir, ">="); rhs <- c(rhs, 0)
}

# At least one of AMZN (2), GOOG (3), AAPL (4)
one_of <- rep(0, 2 * n)
one_of[n + 2] <- 1; one_of[n + 3] <- 1; one_of[n + 4] <- 1
A <- rbind(A, one_of); dir <- c(dir, ">="); rhs <- c(rhs, 1)

# FDX (5) == UPS (6)
both_or_neither <- rep(0, 2 * n)
both_or_neither[n + 5] <- 1
both_or_neither[n + 6] <- -1
A <- rbind(A, both_or_neither); dir <- c(dir, "="); rhs <- c(rhs, 0)

sol_port <- lp(
  direction    = "min",
  objective.in = objective,
  const.mat    = A,
  const.dir    = dir,
  const.rhs    = rhs,
  int.vec      = (n + 1):(2 * n),          # y_i are integer
  binary.vec   = (n + 1):(2 * n)           # and binary
)

x_vals <- sol_port$solution[1:n]
y_vals <- sol_port$solution[(n + 1):(2 * n)]

portfolio <- data.frame(
  stock    = stocks,
  picked   = as.logical(y_vals),
  invested = round(x_vals, 0),
  expected_return = round(x_vals * returns, 0)
)
portfolio

cat("\nTotal invested: $", sum(x_vals), "\n")
cat("Expected return net of fees: $",
    round(sum(x_vals * returns) - sum(y_vals * txn_fee), 0), "\n")
```

The solver picks the best set, respects the minimum-investment rule, includes at least one of {AMZN, GOOG, AAPL}, and treats FedEx/UPS as a matched pair.

---

## The Problem-Type Hierarchy (Easiest → Hardest)

| Type | Notes |
|---|---|
| **Linear Program (LP)** | Millions of variables, fast. What we've been doing. |
| **Convex Quadratic Program (QP)** | Fast. Portfolio optimization, ridge regression. |
| **Convex Optimization (general)** | Still "easy" in theory, can be slow. |
| **Integer Program (IP)** | Hard. Some small IPs don't have known optimal solutions even after days of compute. |
| **General Non-Convex** | Brutal. Even tiny problems may be intractable. |
| **Network Optimization** | *Faster* than LP. Shortest path, assignment, max flow. |

::: {.callout-tip}
**Recognizing the type matters.** A problem written as an IP may be intractable, but the same problem re-formulated as a network-flow problem can solve in milliseconds. This is why modeling ability matters more than knowing how to call a solver.
:::

---

## Handling Uncertainty — Four Approaches

When input data is uncertain (demand forecasts, returns, transition rates), the deterministic optimum can be catastrophically bad. Four approaches:

| Approach | Idea |
|---|---|
| **Conservative constraints** | Pad the RHS: require demand + $\theta$ extra buffer. |
| **Chance constraints** | $P(\text{demand met}) \ge p$. Needs a distribution estimate. |
| **Scenario optimization** | Define multiple scenarios, solve for a *robust* solution (feasible in all) or minimize expected cost across them. |
| **Stochastic dynamic programming / Markov Decision Processes** | Model the system as states + decisions + transition probabilities and find the best policy via Bellman's equation. |

Scenario optimization in particular is very flexible — you can represent a messy probabilistic world as a discrete set of "what-ifs" and let the solver pick a policy that works across them.

---

## Regression as Optimization — The Terminology Collision

Every regression, LASSO, ridge, logistic regression, SVM, exponential-smoothing, ARIMA, GARCH, and k-means model discussed in this guide is really an optimization problem inside. The confusing bit is terminology: statisticians call the data $x_{ij}$ "variables" (they vary across rows); optimizers call the coefficients $a_0, a_1, \ldots$ "variables" (they're what the solver picks).

Let's solve linear regression explicitly as optimization to drive the point home.

```{r}
#| label: regression-as-optim
#| echo: true

set.seed(42)
n <- 60
x <- runif(n, 0, 10)
y <- 2 + 1.5 * x + rnorm(n, 0, 1)

# Standard lm
fit_lm <- lm(y ~ x)
cat("lm() intercept =",    round(coef(fit_lm)[1], 4),
    "   slope =",          round(coef(fit_lm)[2], 4), "\n")

# As an optimization problem: minimize sum((y - a0 - a1*x)^2) over (a0, a1)
sse <- function(params) {
  a0 <- params[1]; a1 <- params[2]
  sum((y - a0 - a1 * x)^2)
}

fit_optim <- optim(par = c(0, 0), fn = sse, method = "BFGS")
cat("optim() intercept =", round(fit_optim$par[1], 4),
    "   slope =",          round(fit_optim$par[2], 4), "\n")
```

Same answer. The `optim()` version is doing what `lm()` does under the hood — it just exposes the optimization nature. And because it's an optimization problem, we can *customize* it: force the intercept to zero, constrain coefficients to be monotone, use $|y - \hat{y}|^{1.5}$ instead of squared error, etc. Standard regression packages can't do that; optimization solvers can.

---

## Cheat Sheet

::: {.callout-warning}
**Common pitfalls recap:**

1. **"More optimal" is not a thing.** A solution is optimal or it isn't.
2. **Problem formulation > solver choice.** The call-center example is the canonical illustration.
3. **Link the variables.** Solvers only see math, not your English intent. $\sum_d v_{id} = y_i$ may be invisible to the solver if you don't write it.
4. **LP optima live at vertices.** That's why simplex works.
5. **Binary variables make IPs hard.** Don't default to "$y \in \{0, 1\}$" without considering whether an LP or network formulation would do.
6. **Regression is optimization in disguise.** So are LASSO, SVM, exponential smoothing, and k-means. Knowing this lets you customize them.
7. **Statisticians and optimizers mean different things by "variable."** Always check which hat the person is wearing.
:::
