---
title: "PCA & Box-Cox From Scratch: Fixing Your Data Before Modeling"
subtitle: "Dimensionality reduction and variance stabilization"
engine: knitr
---

## 1. The Problem: Too Many Correlated Predictors

You're building a regression model. You have 50 predictors. Many of them
measure similar things — they're **correlated**. This causes problems:

- **Multicollinearity**: correlated predictors make coefficients unstable
- **Overfitting**: too many predictors relative to data points
- **Noise**: some predictors are mostly noise, not signal

PCA solves all three by compressing your predictors into a smaller set of
**uncorrelated** components that capture the important patterns.

```{r}
#| label: fig-correlated-data
#| fig-cap: "These two features are highly correlated — they carry redundant information"

library(ggplot2)
set.seed(42)

n <- 100

# Two correlated features
x1 <- rnorm(n, 50, 10)
x2 <- 0.8 * x1 + rnorm(n, 10, 5)  # correlated with x1

df <- data.frame(x1 = x1, x2 = x2)

ggplot(df, aes(x1, x2)) +
  geom_point(size = 2, color = "gray40") +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1 (e.g., assessment score)",
       y = "Feature 2 (e.g., homework score)",
       title = paste0("Correlation = ", round(cor(x1, x2), 2),
                       " — these carry overlapping information"))
```

If both features carry almost the same information, do you really need both?
PCA finds the **directions** in the data that capture the most variation, and
lets you keep just those.


## 2. PCA: What It Does Geometrically

PCA **rotates** your coordinate system so that:

1. The **first axis** (PC1) points in the direction of maximum spread
2. The **second axis** (PC2) points perpendicular to PC1, capturing the next most spread
3. And so on...

```{r}
#| label: fig-pca-rotation
#| fig-cap: "PCA rotates the axes to align with the directions of maximum variance"

# Standardize
df_s <- scale(df)

# PCA
pca <- prcomp(df_s)

# Plot original data with PCA directions
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Original space with PCA arrows
plot(df_s[, 1], df_s[, 2], pch = 19, col = "gray40", asp = 1,
  main = "Original Space + PCA Directions",
  xlab = "Feature 1 (standardized)", ylab = "Feature 2 (standardized)")

# PC1 direction
arrows(0, 0, pca$rotation[1, 1] * 2, pca$rotation[2, 1] * 2,
  col = "coral", lwd = 3, length = 0.15)
text(pca$rotation[1, 1] * 2.3, pca$rotation[2, 1] * 2.3,
  "PC1", col = "coral", font = 2, cex = 1.2)

# PC2 direction
arrows(0, 0, pca$rotation[1, 2] * 1.5, pca$rotation[2, 2] * 1.5,
  col = "steelblue", lwd = 3, length = 0.15)
text(pca$rotation[1, 2] * 1.8, pca$rotation[2, 2] * 1.8,
  "PC2", col = "steelblue", font = 2, cex = 1.2)

# Rotated space (PCA scores)
scores <- pca$x
plot(scores[, 1], scores[, 2], pch = 19, col = "gray40", asp = 1,
  main = "PCA Space (Rotated)",
  xlab = "PC1 (most variance)", ylab = "PC2 (remaining variance)")
abline(h = 0, v = 0, lty = 2, col = "gray70")

par(mfrow = c(1, 1))
```

Notice in the PCA space:

- **PC1** (horizontal) captures the main spread — most of the information
- **PC2** (vertical) captures what's left — much less spread
- The two components are **uncorrelated** (perpendicular)


## 3. The PCA Algorithm: Step by Step

1. **Standardize** the data (mean = 0, SD = 1 for each feature)
2. Compute the **covariance matrix** ($\mathbf{X}^T \mathbf{X}$)
3. Find the **eigenvalues** and **eigenvectors**
4. Sort by eigenvalue (largest first)
5. First eigenvector = PC1, second = PC2, etc.

### What Are Eigenvalues and Eigenvectors?

| Concept | What it is | Plain English |
|:--------|:-----------|:-------------|
| **Eigenvector** | A direction in feature space | Which way does the data spread? |
| **Eigenvalue** | A number paired with each eigenvector | How much variance is in that direction? |

The eigenvector with the **largest** eigenvalue points in the direction of
**maximum variance** — that's PC1.

```{r}
#| label: fig-eigenvalues
#| fig-cap: "Eigenvalues tell you how much variance each component captures"

# Show eigenvalues
eigenvalues <- pca$sdev^2
var_explained <- eigenvalues / sum(eigenvalues) * 100

cat("Eigenvalues:\n")
cat(sprintf("  PC1: %.2f (%.1f%% of variance)\n", eigenvalues[1], var_explained[1]))
cat(sprintf("  PC2: %.2f (%.1f%% of variance)\n", eigenvalues[2], var_explained[2]))
cat(sprintf("\nPC1 alone captures %.1f%% of the information!\n", var_explained[1]))
cat("With correlated data, you can often drop PC2 and lose very little.")
```


## 4. Higher Dimensions: Where PCA Really Shines

Two features is easy to visualize. But the real power of PCA is compressing
**many** features into a few components.

```{r}
#| label: fig-high-dim
#| fig-cap: "10 correlated features compressed into components — most variance in first 2-3"

set.seed(42)

# Generate 10 correlated features (all driven by 2 hidden factors)
hidden1 <- rnorm(n)
hidden2 <- rnorm(n)

features <- matrix(nrow = n, ncol = 10)
for (j in 1:10) {
  w1 <- runif(1, 0.5, 1.5) * sample(c(-1, 1), 1)
  w2 <- runif(1, 0.2, 0.8) * sample(c(-1, 1), 1)
  features[, j] <- w1 * hidden1 + w2 * hidden2 + rnorm(n, 0, 0.5)
}
colnames(features) <- paste0("X", 1:10)

# PCA
pca10 <- prcomp(scale(features))
var_exp10 <- pca10$sdev^2 / sum(pca10$sdev^2) * 100
cum_var <- cumsum(var_exp10)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Scree plot
barplot(var_exp10, names.arg = paste0("PC", 1:10),
  col = ifelse(cum_var <= 90, "steelblue", "gray80"),
  main = "Scree Plot: Variance per Component",
  ylab = "% Variance Explained", las = 2)

# Cumulative variance
plot(1:10, cum_var, type = "b", pch = 19, col = "steelblue", lwd = 2,
  main = "Cumulative Variance Explained",
  xlab = "Number of Components", ylab = "% Total Variance",
  ylim = c(0, 100))
abline(h = 90, lty = 2, col = "coral")
text(7, 85, "90% threshold", col = "coral", cex = 0.9)

# Mark how many needed for 90%
n_for_90 <- which(cum_var >= 90)[1]
points(n_for_90, cum_var[n_for_90], pch = 19, cex = 2, col = "coral")
text(n_for_90 + 1, cum_var[n_for_90] - 5,
  paste0(n_for_90, " components\nfor 90%"), col = "coral", cex = 0.9)

par(mfrow = c(1, 1))
```

**10 features → `r n_for_90` components** that capture 90%+ of the
information. The remaining components are mostly noise.


## 5. Choosing How Many Components: The Scree Plot / Elbow

Same idea as the elbow method for k-means:

1. Plot eigenvalues (or % variance) for each component
2. Look for the **elbow** — where the curve flattens
3. Keep components before the dropoff

**Alternative rule:** Keep components until you reach **80-90%** cumulative
variance explained.

```{r}
#| label: fig-scree-elbow
#| fig-cap: "The scree plot elbow suggests 2-3 components are sufficient"

plot(1:10, pca10$sdev^2, type = "b", pch = 19, col = "steelblue", lwd = 2,
  main = "Scree Plot: Look for the Elbow",
  xlab = "Component Number", ylab = "Eigenvalue")

# Highlight the elbow
points(3, pca10$sdev[3]^2, pch = 1, cex = 3, col = "coral", lwd = 2)
text(4.5, pca10$sdev[3]^2, "Elbow here →\nkeep 2-3 components",
  col = "coral", cex = 0.9)
```


## 6. The Critical PCA Trap (High-Value Distinction)

::: {.callout-warning}
### PCA Ranks by Variance, NOT Predictive Power

**PCA ranks components by variance in X, NOT by predictive power for Y.**

A high-variance component might be **irrelevant noise**. A low-variance
component might contain the **actual signal** you need for prediction.
:::

```{r}
#| label: fig-pca-trap
#| fig-cap: "PC1 has the most variance but PC3 actually predicts the response"

set.seed(42)

# Create data where the signal is in a low-variance direction
n_trap <- 150
noise_big <- rnorm(n_trap, 0, 5)    # big variance, no signal
noise_med <- rnorm(n_trap, 0, 3)    # medium variance, no signal
signal <- rnorm(n_trap, 0, 1)       # small variance, IS the signal

X_trap <- cbind(
  noise_big + 0.1 * signal,
  noise_med + 0.05 * signal,
  0.5 * noise_big + 0.8 * signal    # signal mixed in
)

y_trap <- 3 * signal + rnorm(n_trap, 0, 0.5)  # y depends on the hidden signal

pca_trap <- prcomp(scale(X_trap))

# Correlation of each PC with y
pc_cor <- sapply(1:3, function(j) cor(pca_trap$x[, j], y_trap))

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

barplot(pca_trap$sdev^2, names.arg = paste0("PC", 1:3),
  col = "steelblue", main = "Variance (what PCA ranks by)",
  ylab = "Eigenvalue")

barplot(abs(pc_cor), names.arg = paste0("PC", 1:3),
  col = c("gray70", "gray70", "coral"),
  main = "Correlation with Y (what matters!)",
  ylab = "|Correlation with response|")

par(mfrow = c(1, 1))

cat("PC1 has the MOST variance but the LEAST predictive power.\n")
cat("PC3 has the LEAST variance but the MOST predictive power.\n")
cat("\nThis is why you MUST validate PCA-based models on actual prediction.\n")
cat("Don't blindly drop low-variance components!")
```

::: {.callout-warning}
### Practical Rule
PCA is a **preprocessing** step. You must still validate the
full model (PCA + regression/SVM/etc.) via cross-validation.
:::


## 7. PCA Requirements

| Requirement | Why | What happens if violated |
|:------------|:----|:------------------------|
| **Standardize first** | Features with larger ranges dominate | PCA reflects scale, not importance |
| **Continuous features** | PCA assumes linear relationships | Binary/categorical features need other methods |
| **Some correlation** | If features are uncorrelated, PCA can't compress | You get back the same features |

```{r}
#| label: fig-scaling-pca
#| fig-cap: "Without scaling, the high-range feature dominates PC1"

set.seed(42)

# Income: 30k-90k, Age: 20-65
data_unscaled <- data.frame(
  income = rnorm(80, 60000, 15000),
  age = rnorm(80, 40, 10)
)

pca_unscaled <- prcomp(data_unscaled, scale. = FALSE)
pca_scaled <- prcomp(data_unscaled, scale. = TRUE)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Loadings comparison
barplot(abs(pca_unscaled$rotation[, 1]), names.arg = c("Income", "Age"),
  col = c("coral", "steelblue"), main = "PC1 Loadings: UNSCALED",
  ylab = "Absolute Loading")
mtext("Income dominates (larger range)", side = 1, line = 2.5, cex = 0.8, col = "red")

barplot(abs(pca_scaled$rotation[, 1]), names.arg = c("Income", "Age"),
  col = c("coral", "steelblue"), main = "PC1 Loadings: SCALED",
  ylab = "Absolute Loading")
mtext("Both contribute fairly", side = 1, line = 2.5, cex = 0.8, col = "darkgreen")

par(mfrow = c(1, 1))
```


---

# Part 2: Box-Cox Transformation

## 8. A Different Problem: Unequal Variance

Regression assumes **constant variance** — the spread of residuals should be
the same everywhere. But real data often violates this:

```{r}
#| label: fig-heteroscedasticity
#| fig-cap: "The funnel shape: variance increases with the predicted value"

set.seed(42)
x_bc <- runif(100, 1, 20)
y_bc <- 2 * x_bc + rnorm(100, 0, x_bc * 0.8)  # variance grows with x

df_bc <- data.frame(x = x_bc, y = y_bc)
fit_bc <- lm(y ~ x, data = df_bc)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(df_bc$x, df_bc$y, pch = 19, col = "gray40",
  main = "Data: Variance Grows with X",
  xlab = "X", ylab = "Y")
abline(fit_bc, col = "steelblue", lwd = 2)

plot(fitted(fit_bc), residuals(fit_bc), pch = 19, col = "gray40",
  main = "Residual Plot: Funnel Shape",
  xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2, col = "coral")
text(25, 15, "Variance increases →\n(heteroscedasticity)", col = "coral", cex = 0.9)

par(mfrow = c(1, 1))
```

This funnel shape means the model's predictions are **unreliable** — accurate
for small values, wildly off for large values.

**Box-Cox fixes this** by transforming the response variable.


## 9. The Box-Cox Formula

$$T(y) = \begin{cases} \frac{y^{\lambda} - 1}{\lambda} & \text{if } \lambda \neq 0 \\ \log(y) & \text{if } \lambda = 0 \end{cases}$$

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $y$ | Original response variable | The thing you're predicting |
| $T(y)$ | Transformed response | The stabilized version |
| $\lambda$ | Power parameter | Controls the shape of the transformation |

**Special cases of $\lambda$:**

| $\lambda$ | Transform | Effect |
|:---------:|:----------|:-------|
| 1 | $y$ (no change) | Data is fine as-is |
| 0.5 | $\sqrt{y}$ | Mild compression of large values |
| 0 | $\log(y)$ | Strong compression — common in practice |
| -1 | $1/y$ | Very aggressive compression |

Box-Cox **finds the optimal $\lambda$** automatically by trying different
values and picking the one that makes the residuals most normally distributed.

```{r}
#| label: fig-boxcox-lambda
#| fig-cap: "Box-Cox finds the optimal λ that stabilizes variance"

library(MASS)

# Need positive y for Box-Cox
y_pos <- y_bc - min(y_bc) + 1
df_pos <- data.frame(x = x_bc, y = y_pos)
fit_pos <- lm(y ~ x, data = df_pos)

# Box-Cox profile
bc <- boxcox(fit_pos, plotit = TRUE)

# Find optimal lambda
optimal_lambda <- bc$x[which.max(bc$y)]
cat(sprintf("\nOptimal λ = %.2f\n", optimal_lambda))
cat(sprintf("Closest standard transform: %s\n",
  ifelse(abs(optimal_lambda) < 0.2, "log(y)",
    ifelse(abs(optimal_lambda - 0.5) < 0.2, "√y",
      ifelse(abs(optimal_lambda - 1) < 0.2, "no transform needed",
        paste0("y^", round(optimal_lambda, 2)))))))
```

```{r}
#| label: fig-boxcox-effect
#| fig-cap: "Before and after Box-Cox: the funnel shape disappears"

# Apply transform
if (abs(optimal_lambda) < 0.01) {
  y_transformed <- log(y_pos)
} else {
  y_transformed <- (y_pos^optimal_lambda - 1) / optimal_lambda
}

df_trans <- data.frame(x = x_bc, y = y_transformed)
fit_trans <- lm(y ~ x, data = df_trans)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Before
plot(fitted(fit_pos), residuals(fit_pos), pch = 19, col = "coral",
  main = "BEFORE Box-Cox: Residuals",
  xlab = "Fitted", ylab = "Residual")
abline(h = 0, lty = 2)

hist(residuals(fit_pos), breaks = 15, col = "coral", border = "white",
  main = "BEFORE: Residual Distribution", xlab = "Residual")

# After
plot(fitted(fit_trans), residuals(fit_trans), pch = 19, col = "steelblue",
  main = "AFTER Box-Cox: Residuals",
  xlab = "Fitted", ylab = "Residual")
abline(h = 0, lty = 2)

hist(residuals(fit_trans), breaks = 15, col = "steelblue", border = "white",
  main = "AFTER: Residual Distribution", xlab = "Residual")

par(mfrow = c(1, 1))
```

The funnel shape is gone. Residuals are more uniform and more normal.


## 10. Detrending: Removing Time's Influence

When you want to study the relationship between X and Y, but **both** are
trending over time, the time trend creates a **spurious correlation**.

**Fix:** Remove the trend first, then analyze the detrended data.

$$y_{\text{detrended}} = y - (\hat{a}_0 + \hat{a}_1 \cdot \text{time})$$

```{r}
#| label: fig-detrending
#| fig-cap: "Both variables trend upward with time → spurious correlation. Detrending reveals the truth."

set.seed(42)
time_dt <- 1:60

# Two UNRELATED things that both trend upward
ice_cream <- 10 + 0.3 * time_dt + rnorm(60, 0, 2)
drownings <- 5 + 0.15 * time_dt + rnorm(60, 0, 1.5)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Raw correlation (spurious)
plot(ice_cream, drownings, pch = 19, col = "coral",
  main = paste0("Raw: r = ", round(cor(ice_cream, drownings), 2)),
  xlab = "Ice Cream Sales", ylab = "Drownings")
abline(lm(drownings ~ ice_cream), col = "coral", lwd = 2)

# Both trending
plot(time_dt, ice_cream, type = "l", col = "coral", lwd = 2,
  main = "Both variables trend with time",
  xlab = "Time", ylab = "Value", ylim = c(0, 35))
lines(time_dt, drownings, col = "steelblue", lwd = 2)
legend("topleft", legend = c("Ice cream", "Drownings"), bty = "n",
  col = c("coral", "steelblue"), lwd = 2, cex = 0.8)

# Detrend
ice_dt <- residuals(lm(ice_cream ~ time_dt))
drown_dt <- residuals(lm(drownings ~ time_dt))

# Detrended correlation (real)
plot(ice_dt, drown_dt, pch = 19, col = "steelblue",
  main = paste0("Detrended: r = ", round(cor(ice_dt, drown_dt), 2)),
  xlab = "Ice Cream (detrended)", ylab = "Drownings (detrended)")
abline(lm(drown_dt ~ ice_dt), col = "steelblue", lwd = 2)

# Detrended series
plot(time_dt, ice_dt, type = "l", col = "coral", lwd = 2,
  main = "After detrending: no trend",
  xlab = "Time", ylab = "Detrended Value")
lines(time_dt, drown_dt, col = "steelblue", lwd = 2)
abline(h = 0, lty = 2, col = "gray50")

par(mfrow = c(1, 1))
```

After removing the time trend, the spurious correlation disappears.


## 11. PCA + Regression: The Full Workflow

The typical use of PCA in this guide:

1. You have many correlated predictors
2. Run PCA to get uncorrelated components
3. Use cross-validation to decide how many components to keep
4. Feed components into regression (or SVM, KNN, etc.)
5. Validate the final model on test data

```{r}
#| label: fig-pca-regression
#| fig-cap: "PCA + regression: compress 10 features into 3 components, then regress"

# Use the 10-feature data from earlier
y_demo <- 2 * hidden1 - 1.5 * hidden2 + rnorm(n, 0, 0.5)

# Full regression (all 10 features)
fit_full <- lm(y_demo ~ features)

# PCA regression (3 components)
scores3 <- pca10$x[, 1:3]
fit_pca <- lm(y_demo ~ scores3)

cat("=== Full model (10 predictors) ===\n")
cat(sprintf("  R² = %.3f\n", summary(fit_full)$r.squared))
cat(sprintf("  Adj R² = %.3f\n\n", summary(fit_full)$adj.r.squared))

cat("=== PCA model (3 components) ===\n")
cat(sprintf("  R² = %.3f\n", summary(fit_pca)$r.squared))
cat(sprintf("  Adj R² = %.3f\n", summary(fit_pca)$adj.r.squared))
cat("\n10 predictors → 3 components with similar Adj R²")
cat("\nSimpler model, less overfitting risk, no multicollinearity")
```


## 12. Cheat Sheet: The Whole Story on One Page

```
PCA RECIPE
==========

1. PURPOSE: Compress many correlated features into fewer uncorrelated components

2. ALGORITHM:
   Standardize → Covariance matrix → Eigenvalues/vectors → Sort by variance

3. KEY CONCEPTS:
   Eigenvector = direction of spread (becomes a component)
   Eigenvalue = how much variance in that direction
   PC1 = most variance, PC2 = second most, etc.

4. HOW MANY COMPONENTS?
   Scree plot elbow OR cumulative variance ≥ 80-90%

5. CRITICAL TRAP:
   PCA ranks by VARIANCE in X, not by PREDICTIVE POWER for Y
   High-variance PC might be noise. Low-variance PC might be the signal.
   ALWAYS validate with cross-validation.

6. REQUIREMENTS:
   - ALWAYS standardize first (scale. = TRUE)
   - Continuous features
   - Some correlation among features (otherwise PCA can't compress)

BOX-COX RECIPE
==============

1. PURPOSE: Stabilize variance (fix heteroscedasticity)

2. FORMULA: T(y) = (y^λ - 1) / λ

3. SPECIAL CASES:
   λ = 1  → no transform     λ = 0   → log(y)
   λ = 0.5 → √y              λ = -1  → 1/y

4. HOW: R finds optimal λ automatically (boxcox function)

5. WHEN: Residual plot shows funnel shape (variance not constant)

DETRENDING
==========
Remove time trend before analyzing relationships:
  y_detrended = y - (a₀ + a₁ × time)
Prevents spurious correlations caused by shared time trends.
```


## 13. Check Your Understanding

::: {.callout-note}
### Test Yourself

Before moving on, try to answer these without scrolling up:

1. What problem does PCA solve? When would you use it?
2. What is an eigenvector in PCA terms? What does the eigenvalue tell you?
3. Why must you standardize before PCA?
4. You run PCA and keep only PC1 and PC2 (85% variance). A colleague says
   you're losing important information. Could they be right? Why?
5. What does Box-Cox fix? How do you detect the problem it solves?
6. $\lambda = 0$ in Box-Cox means you should apply which transformation?
7. You find a strong correlation between umbrella sales and car accidents.
   What should you do before concluding umbrellas cause accidents?
8. PCA gives you 5 components from 20 features. How do you decide which
   components to keep?
:::
