---
title: "Simulation & Markov: Probability Models for Real Systems"
subtitle: "Distributions, QQ plots, M/M/1 queues, and steady states"
engine: knitr
---

## The Big Picture

Most of this guide has been "throw data at a model." This module inverts that: when the *only* thing you know about a data point is its value, a well-matched probability distribution can solve the problem without any feature engineering.

A useful opening story is a startup that wanted to resell empty season-ticket seats as upgrades. A big feature list (ticket holder demographics, weather, team record, opponent, day of week) looked inevitable — until the team realized fans tend to arrive at about the same *relative* position in the crowd every game. The deviation is approximately normal. A single distribution replaced the whole feature pipeline.

This demo walks through the distributions, QQ plots, a simple queuing simulation, and Markov-chain steady state — the four tools the module uses to reason about systems.

::: {.callout-tip}
**Rule of thumb:** before engineering features, fit a distribution. If it fits, you may be done.
:::

---

## The Distributions You Should Recognize

### Bernoulli → Binomial → Geometric

The Bernoulli trial is a single weighted coin flip. Chain $n$ of them together and count successes → **binomial**. Ask how many until the first success → **geometric**.

```{r}
#| label: fig-discrete-dists
#| fig-cap: "Binomial PMFs for different p (top); Geometric PMF and survival function (bottom)."
#| fig-width: 9
#| fig-height: 6
#| echo: false

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Binomial p = 0.3
x <- 0:20
plot(x, dbinom(x, size = 20, prob = 0.3), type = "h", lwd = 3,
     col = "#2d6a4f", xlab = "x (successes)", ylab = "P(X = x)",
     main = "Binomial(n=20, p=0.3)")
points(x, dbinom(x, size = 20, prob = 0.3), pch = 19, col = "#2d6a4f")

# Binomial p = 0.7
plot(x, dbinom(x, size = 20, prob = 0.7), type = "h", lwd = 3,
     col = "#d62828", xlab = "x (successes)", ylab = "P(X = x)",
     main = "Binomial(n=20, p=0.7)")
points(x, dbinom(x, size = 20, prob = 0.7), pch = 19, col = "#d62828")

# Geometric PMF (p = 0.25)
k <- 1:20
plot(k, dgeom(k - 1, prob = 0.25), type = "h", lwd = 3,
     col = "#1f4e79", xlab = "Trials until first success",
     ylab = "P(X = k)", main = "Geometric(p=0.25) PMF")
points(k, dgeom(k - 1, prob = 0.25), pch = 19, col = "#1f4e79")

# Geometric survival P(X > k)
plot(k, 1 - pgeom(k - 1, prob = 0.25), type = "l", lwd = 2,
     col = "#6c00a3", xlab = "k", ylab = "P(X > k)",
     main = "Geometric survival (memoryless)")
grid()
```

The **binomial** converges to the normal distribution as $n$ grows, which is why the normal is so useful — it's the large-$n$ limit of so many counting processes.

The **geometric** is your "how many trials until X" tool. It doubles as a *diagnostic*: if you want to know whether at-bats are really i.i.d. Bernoulli trials, check whether at-bats-until-broken-bat fits a geometric distribution. If it does, you don't need to model weather, pitcher, pitch type, batter — the independence assumption is already good enough.

::: {.callout-warning}
**Watch out:** in the geometric distribution, $p$ is the probability of the thing you're *waiting for*. Sometimes that's a good outcome (job offer), sometimes bad (defective unit). Define it explicitly every time — you'll see this trap on practice checks.
:::

### Poisson ↔ Exponential (the same distribution in disguise)

If customers arrive at an average rate $\lambda$ per minute and arrivals are independent, the **count in a fixed window** is Poisson and the **inter-arrival time** is exponential with mean $1/\lambda$. One distribution in time units, the other in count units.

```{r}
#| label: fig-poisson-exp
#| fig-cap: "Poisson PMF and the matching exponential density."
#| fig-width: 9
#| fig-height: 4
#| echo: false

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

lambda <- 4
x <- 0:15
plot(x, dpois(x, lambda), type = "h", lwd = 3, col = "#2d6a4f",
     xlab = "Arrivals per minute", ylab = "P(X = x)",
     main = bquote("Poisson(" * lambda == .(lambda) * ")"))
points(x, dpois(x, lambda), pch = 19, col = "#2d6a4f")

t <- seq(0, 2, length.out = 400)
plot(t, dexp(t, rate = lambda), type = "l", lwd = 2, col = "#d62828",
     xlab = "Inter-arrival time (min)", ylab = "density",
     main = bquote("Exponential(rate =" ~ .(lambda) * ")"))
grid()
```

### Weibull — failure rate that varies with age

The **Weibull** is like the geometric, but continuous in time. Its shape parameter $k$ controls whether the failure rate shrinks, stays flat, or grows.

```{r}
#| label: fig-weibull
#| fig-cap: "Weibull density for k < 1 (weak parts fail first), k = 1 (exponential), k > 1 (wear-out)."
#| fig-width: 7
#| fig-height: 4
#| echo: false

t <- seq(0.001, 3, length.out = 500)
plot(t, dweibull(t, shape = 0.5, scale = 1), type = "l", lwd = 2,
     col = "#2d6a4f", xlab = "time", ylab = "density",
     main = "Weibull density for three shape parameters k",
     ylim = c(0, 2))
lines(t, dweibull(t, shape = 1.0, scale = 1), lwd = 2, col = "#1f4e79")
lines(t, dweibull(t, shape = 2.0, scale = 1), lwd = 2, col = "#d62828")
legend("topright",
       legend = c("k = 0.5 (decreasing hazard)",
                  "k = 1   (constant = exponential)",
                  "k = 2   (increasing hazard = wear-out)"),
       col = c("#2d6a4f", "#1f4e79", "#d62828"), lwd = 2, bty = "n")
grid()
```

::: {.callout-note}
**Check your understanding:** Tires are more likely to fail the more miles they've been driven. Is this memoryless? Which distribution fits?

*Answer: not memoryless (depends on history), so not exponential. Weibull with $k > 1$ is the right tool.*
:::

---

## QQ Plots — Sanity Check for Distribution Fit

A statistical test collapses the fit to a single number. A QQ plot shows you *where* a distribution diverges from the data — which matters enormously if the tails are the point of your project.

```{r}
#| label: fig-qq
#| fig-cap: "QQ plots: (left) simulated exponential data vs theoretical exponential — nearly straight. (right) the same data vs theoretical normal — clearly bad, especially in the tails."
#| fig-width: 9
#| fig-height: 4
#| echo: false

set.seed(42)
exp_data <- rexp(500, rate = 1)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# QQ vs exponential
qq_exp <- qexp(ppoints(length(exp_data)))
plot(sort(qq_exp), sort(exp_data), pch = 19, cex = 0.6, col = "#2d6a4f",
     xlab = "Theoretical exponential quantiles",
     ylab = "Sample quantiles",
     main = "QQ vs Exponential (good fit)")
abline(0, 1, col = "gray60", lty = 2, lwd = 2)
grid()

# QQ vs normal
qqnorm(exp_data, pch = 19, cex = 0.6, col = "#d62828",
       main = "QQ vs Normal (bad fit)")
qqline(exp_data, col = "gray60", lty = 2, lwd = 2)
grid()
```

::: {.callout-tip}
If the data and the theoretical distribution match, points fall on the 45° line. *Where* they drift off tells you the story — upper-tail heavier than normal, lower-tail compressed, etc.
:::

---

## Why Averages Aren't Enough — A Queuing Simulation

A classic queueing example: wafers arrive at a semiconductor machine every 20 seconds on average; processing takes 19 seconds on average. Under these averages alone, the queue is always empty and throughput is effortless. Add Poisson arrivals and exponential service — same means — and the average queue explodes to ~19 wafers.

Let's simulate it.

```{r}
#| label: simulate-mm1
#| echo: true

set.seed(1)

simulate_mm1 <- function(n_customers = 5000, lambda = 1 / 20, mu = 1 / 19) {
  inter_arrivals <- rexp(n_customers, rate = lambda)
  service_times  <- rexp(n_customers, rate = mu)

  arrival_time <- cumsum(inter_arrivals)
  start_time   <- numeric(n_customers)
  finish_time  <- numeric(n_customers)

  start_time[1]  <- arrival_time[1]
  finish_time[1] <- start_time[1] + service_times[1]

  for (i in 2:n_customers) {
    start_time[i]  <- max(arrival_time[i], finish_time[i - 1])
    finish_time[i] <- start_time[i] + service_times[i]
  }

  wait_time <- start_time - arrival_time
  data.frame(arrival_time, start_time, finish_time, wait_time)
}

sim <- simulate_mm1()
cat("Mean wait time (sec):", round(mean(sim$wait_time), 2), "\n")
cat("Max wait time  (sec):", round(max(sim$wait_time), 2), "\n")
cat("Utilization (expected):", round((1 / 19) / (1 / 20), 3), "\n")
```

```{r}
#| label: fig-mm1
#| fig-cap: "Wait-time distribution for an M/M/1 queue with mean inter-arrival 20s and mean service 19s. Most waits are short but the tail is long — the *average* already disagrees with the 'no queue' story that averaging inputs would tell."
#| fig-width: 8
#| fig-height: 4
#| echo: false

hist(sim$wait_time, breaks = 60, col = "#c7e4d2", border = "#2d6a4f",
     main = "M/M/1 wait-time distribution",
     xlab = "Wait time (seconds)")
abline(v = mean(sim$wait_time), col = "#d62828", lwd = 2, lty = 2)
legend("topright", paste("mean =", round(mean(sim$wait_time), 1), "s"),
       col = "#d62828", lwd = 2, lty = 2, bty = "n")
```

::: {.callout-warning}
**Common pitfall:** "If arrivals average every 20 seconds and service takes 19 seconds, the queue is small." *This is wrong.* Variability in inter-arrival and service times — even with those means — creates a substantial queue. Simulation exists precisely because averages can't predict queue behavior once randomness enters.
:::

---

## Memorylessness — The Single Property That Makes the Math Tractable

The exponential distribution has one peculiar property: given that you've already waited $s$ units, the distribution of remaining time is *still* exponential with the same rate. Past history is irrelevant.

```{r}
#| label: memoryless-check
#| echo: true

set.seed(7)
samples <- rexp(50000, rate = 1)

# Unconditional
p_exceed_2 <- mean(samples > 2)

# Conditional: already waited 3, ask for 2 more
survived_3 <- samples[samples > 3]
p_exceed_3_plus_2 <- mean(survived_3 > 5)

cat("P(X > 2) unconditional:       ", round(p_exceed_2, 4), "\n")
cat("P(X > 5 | X > 3) conditional: ", round(p_exceed_3_plus_2, 4), "\n")
```

The two numbers are (approximately) the same. That's what memorylessness means — and it's why M/M/1 queues admit closed-form solutions. Tire failure, by contrast, *is* history-dependent (worn tires fail faster), so tires cannot be modeled as exponential — Weibull with $k > 1$ is the right tool.

---

## Markov Chains — Steady State via Eigen-Decomposition

A Markov chain has a finite set of states and a transition matrix $P$ where $P_{ij}$ is the probability of moving from state $i$ to state $j$ in one step. Starting with an initial distribution $\pi$, the distribution after $k$ steps is $\pi P^k$. The **steady-state** distribution $\pi^*$ is the one that doesn't change: $\pi^* P = \pi^*$ with $\sum \pi_i^* = 1$.

Mathematically, $\pi^*$ is the left eigenvector of $P$ with eigenvalue 1.

```{r}
#| label: markov-weather
#| echo: true

# Classic weather example: Sunny / Cloudy / Rainy
states <- c("Sunny", "Cloudy", "Rainy")
P <- matrix(c(
  0.70, 0.20, 0.10,   # from Sunny
  0.30, 0.40, 0.30,   # from Cloudy
  0.20, 0.45, 0.35    # from Rainy
), nrow = 3, byrow = TRUE)
rownames(P) <- colnames(P) <- states
P

# Eigen decomposition of P^T — the left eigenvector of P is the right eigenvector of P^T
eig <- eigen(t(P))

# Find the eigenvector corresponding to eigenvalue ~ 1
idx <- which.min(abs(eig$values - 1))
pi_star_raw <- Re(eig$vectors[, idx])
pi_star <- pi_star_raw / sum(pi_star_raw)
names(pi_star) <- states
round(pi_star, 4)
```

```{r}
#| label: markov-verify
#| echo: true

# Verify by simulation: start in Sunny, take many steps, record visit fractions
set.seed(42)
n_steps <- 100000
state <- 1L  # start in Sunny
visits <- integer(3)

for (step in seq_len(n_steps)) {
  visits[state] <- visits[state] + 1L
  state <- sample.int(3, size = 1, prob = P[state, ])
}

empirical <- visits / n_steps
names(empirical) <- states
round(empirical, 4)
```

The eigen-decomposition and the empirical random walk agree — that's the Markov-chain steady state.

::: {.callout-note}
**Check your understanding:** Real weather is *not* actually a good Markov example — real weather is highly history-dependent. Why bother with Markov chains at all?

*Answer: applications like PageRank (ranking web pages by random-walk steady state), LRMC college basketball rankings, urban sprawl, and long-run disease modeling. In these cases the system isn't literally memoryless short-term, but over long horizons the memorylessness assumption washes out and the steady state still gives useful rankings.*
:::

---

## PageRank — One Matrix Multiply

PageRank treats the web as a graph: each page is a state, and the transition probability from page $i$ to page $j$ is proportional to whether $i$ links to $j$. The steady-state distribution tells you how often a random walker lands on each page — Google's original ranking signal.

```{r}
#| label: pagerank
#| echo: true

# Tiny web of 5 pages
link_matrix <- matrix(0, nrow = 5, ncol = 5)
# Page 1 links to 2, 3
link_matrix[1, c(2, 3)] <- 1
# Page 2 links to 3
link_matrix[2, 3] <- 1
# Page 3 links to 1, 4
link_matrix[3, c(1, 4)] <- 1
# Page 4 links to 3, 5
link_matrix[4, c(3, 5)] <- 1
# Page 5 links to 3
link_matrix[5, 3] <- 1

# Row-normalize to get transition matrix
P_web <- link_matrix / rowSums(link_matrix)

# Steady-state via eigen
eig_web <- eigen(t(P_web))
rank_raw <- Re(eig_web$vectors[, which.min(abs(eig_web$values - 1))])
page_rank <- rank_raw / sum(rank_raw)
names(page_rank) <- paste0("Page", 1:5)
sort(round(page_rank, 4), decreasing = TRUE)
```

Page 3 dominates — it receives links from almost every other page, so a random walker gets funneled there.

---

## Cheat Sheet

| Situation | Distribution / tool |
|---|---|
| Single yes/no trial | Bernoulli |
| Count of successes in $n$ trials | Binomial |
| Trials until first success | Geometric |
| Arrivals in a fixed window | Poisson |
| Time between arrivals (memoryless) | Exponential |
| Time to failure (age-dependent) | Weibull ($k \ne 1$) |
| Visual distribution fit | QQ plot |
| Queue with closed-form math | M/M/1 (needs exponential) |
| Queue with arbitrary complexity | Simulation |
| Long-run rankings over a transition graph | Markov chain steady state |

::: {.callout-warning}
**Common pitfalls recap:**

1. **Averages ≠ queues.** Variability makes lines even when $\lambda < \mu$.
2. **Geometric $p$ direction.** Define $p$ as the thing you're waiting for, not always "success."
3. **Exponential only when memoryless.** If history matters, use Weibull.
4. **Weibull k = 1 reduces to exponential.** Software reporting "best fit Weibull $k = 1.002$" is probably just fitting noise.
5. **Markov chains need irreducibility.** A steady state may not exist if states can't reach each other.
:::
