---
title: "SVM From Scratch: Unpacking the Formula"
subtitle: "Start from zero, end with intuition"
engine: knitr
---

## 1. The Problem: Drawing a Line Between Two Groups

Imagine you have a bunch of data points that belong to two categories — say,
"approve" vs "deny" for a loan application. Each point has two measurements
(features). Can we draw a line that separates them?

```{r}
#| label: fig-two-groups
#| fig-cap: "Two groups of points. Can you draw a line to separate them?"

library(ggplot2)
set.seed(42)

# Two clusters
df <- data.frame(
  x1 = c(rnorm(30, 2, 0.6), rnorm(30, 5, 0.6)),
  x2 = c(rnorm(30, 2, 0.6), rnorm(30, 5, 0.6)),
  group = factor(rep(c("Deny", "Approve"), each = 30))
)

ggplot(df, aes(x1, x2, color = group, shape = group)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("steelblue", "coral")) +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1 (e.g., income)", y = "Feature 2 (e.g., credit score)")
```

You could probably draw a line through the gap by eye. But **which** line?
There are infinitely many lines that separate these two groups.


## 2. Not Just Any Line — The Best Line

Here's the key insight: some separating lines are better than others.
A line that barely grazes the closest points is fragile. A line that sits
right in the middle of the gap is robust.

```{r}
#| label: fig-many-lines
#| fig-cap: "Many lines separate the data, but which one is best?"

ggplot(df, aes(x1, x2, color = group, shape = group)) +
  geom_point(size = 3) +
  # Bad line — too close to blue
  geom_abline(intercept = -0.5, slope = 1, linetype = "dashed", color = "gray60") +
  # Bad line — too close to red
  geom_abline(intercept = 1.5, slope = 1, linetype = "dashed", color = "gray60") +
  # Good line — right in the middle
  geom_abline(intercept = 0.5, slope = 1, linewidth = 1.2, color = "black") +
  scale_color_manual(values = c("steelblue", "coral")) +
  annotate("text", x = 1.5, y = 5.5, label = "Too close to red", color = "gray40", size = 3.5) +
  annotate("text", x = 4.5, y = 1.5, label = "Too close to blue", color = "gray40", size = 3.5) +
  annotate("text", x = 5.5, y = 4.2, label = "Best line", fontface = "bold", size = 4) +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2")
```

The **best** line is the one with the **maximum margin** — the widest possible
gap between the line and the nearest points on either side.

That's literally what SVM does: **find the line (or hyperplane) with the
maximum margin**.


## 3. What's a Hyperplane?

Don't let the word scare you. It's just the generalization of a "line" to
higher dimensions.

| Dimensions | Separator      | Math               |
|:----------:|:--------------:|:------------------:|
| 2D         | A line         | $w_1 x_1 + w_2 x_2 + b = 0$ |
| 3D         | A flat plane   | $w_1 x_1 + w_2 x_2 + w_3 x_3 + b = 0$ |
| nD         | A hyperplane   | $\mathbf{w} \cdot \mathbf{x} + b = 0$ |

The equation $\mathbf{w} \cdot \mathbf{x} + b = 0$ defines the decision
boundary. Let's unpack every symbol.


## 4. Unpacking the Formula Piece by Piece

### The Decision Boundary

$$\mathbf{w} \cdot \mathbf{x} + b = 0$$

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $\mathbf{x}$ | A data point | One observation — like (income, credit_score) |
| $\mathbf{w}$ | Weight vector | Controls the **tilt** of the line |
| $b$          | Bias (intercept) | Shifts the line **up or down** |
| $\mathbf{w} \cdot \mathbf{x}$ | Dot product | $w_1 x_1 + w_2 x_2 + \dots$ — a weighted sum |

**How it classifies a new point**: plug in $\mathbf{x}$, compute the value:

- If $\mathbf{w} \cdot \mathbf{x} + b > 0$ → class +1 (e.g., Approve)
- If $\mathbf{w} \cdot \mathbf{x} + b < 0$ → class -1 (e.g., Deny)
- If $\mathbf{w} \cdot \mathbf{x} + b = 0$ → exactly on the boundary

```{r}
#| label: fig-sides-of-line
#| fig-cap: "Points on each side of the decision boundary get different signs"

# Show which side of the line each point falls on
w <- c(1, 1)  # weight vector
b <- -7       # bias

df$decision_value <- w[1] * df$x1 + w[2] * df$x2 + b
df$side <- ifelse(df$decision_value > 0, "Positive (+1)", "Negative (-1)")

ggplot(df, aes(x1, x2, color = side)) +
  geom_point(size = 3) +
  geom_abline(intercept = -b / w[2], slope = -w[1] / w[2], linewidth = 1) +
  scale_color_manual(values = c("steelblue", "coral")) +
  annotate("text", x = 2, y = 5.5,
    label = "w · x + b < 0\n(Negative side)", color = "steelblue", size = 4) +
  annotate("text", x = 5, y = 1.5,
    label = "w · x + b > 0\n(Positive side)", color = "coral", size = 4) +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2", color = "Side of boundary")
```

### The Weight Vector Controls Direction

$\mathbf{w}$ is **perpendicular to the decision boundary**. It points from the
negative class toward the positive class.

- Changing $\mathbf{w}$ **rotates** the line
- Changing $b$ **slides** it parallel to itself

```{r}
#| label: fig-w-rotates
#| fig-cap: "Different weight vectors rotate the decision boundary"

ggplot(df, aes(x1, x2, color = group, shape = group)) +
  geom_point(size = 3, alpha = 0.4) +
  # w = (1, 1): 45-degree line
  geom_abline(intercept = 7, slope = -1, linewidth = 1, color = "black") +
  # w = (2, 1): steeper
  geom_abline(intercept = 10, slope = -2, linewidth = 1, color = "purple", linetype = "dashed") +
  # w = (1, 2): shallower
  geom_abline(intercept = 5, slope = -0.5, linewidth = 1, color = "darkgreen", linetype = "dashed") +
  scale_color_manual(values = c("steelblue", "coral")) +
  annotate("text", x = 1, y = 6.5, label = "w = (1,1)", size = 3.5) +
  annotate("text", x = 3.5, y = 6.5, label = "w = (2,1)", color = "purple", size = 3.5) +
  annotate("text", x = 1, y = 4.2, label = "w = (1,2)", color = "darkgreen", size = 3.5) +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2")
```


## 5. The Margin

The **margin** is the distance between the decision boundary and the nearest
data point on either side.

$$\text{margin} = \frac{2}{\|\mathbf{w}\|}$$

where $\|\mathbf{w}\| = \sqrt{w_1^2 + w_2^2 + \dots}$ is the length of the
weight vector.

**Key insight**: to make the margin **bigger**, we need $\|\mathbf{w}\|$ to be
**smaller**. So SVM tries to **minimize** $\|\mathbf{w}\|$ while still
correctly classifying everything.

```{r}
#| label: fig-margin
#| fig-cap: "The margin is the gap between the boundary and the closest points (support vectors)"

library(e1071)

svm_fit <- svm(group ~ x1 + x2, data = df, kernel = "linear", cost = 1, scale = FALSE)

# Extract the weight vector and intercept from the SVM model
w_svm <- t(svm_fit$coefs) %*% svm_fit$SV
b_svm <- -svm_fit$rho

slope <- -w_svm[1] / w_svm[2]
intercept <- -b_svm / w_svm[2]
margin_offset <- 1 / w_svm[2]

sv_indices <- svm_fit$index

ggplot(df, aes(x1, x2, color = group, shape = group)) +
  geom_point(size = 3) +
  # Decision boundary
  geom_abline(intercept = intercept, slope = slope, linewidth = 1.2) +
  # Margin lines
  geom_abline(intercept = intercept + margin_offset, slope = slope,
              linetype = "dashed", color = "gray50") +
  geom_abline(intercept = intercept - margin_offset, slope = slope,
              linetype = "dashed", color = "gray50") +
  # Circle support vectors
  geom_point(data = df[sv_indices, ], aes(x1, x2),
             color = "black", shape = 1, size = 6, stroke = 1.5) +
  scale_color_manual(values = c("steelblue", "coral")) +
  annotate("text", x = 1.2, y = 6, label = "MARGIN", fontface = "bold", size = 5) +
  annotate("text", x = 1.2, y = 5.5,
           label = paste0("Support vectors: ", length(sv_indices)),
           size = 3.5) +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2")
```

The circled points are **support vectors** — the closest points to the
boundary. They're the only points that matter. Move any other point and the
boundary stays the same. Move a support vector and the boundary shifts.

That's why it's called **Support** Vector Machine — the boundary is
"supported" by these critical points.


## 6. The Optimization Problem (The Full Formula)

Putting it all together, SVM solves:

$$\min_{\mathbf{w}, b} \quad \frac{1}{2}\|\mathbf{w}\|^2$$

subject to:

$$y_i(\mathbf{w} \cdot \mathbf{x}_i + b) \geq 1 \quad \text{for all } i$$

Let's unpack both parts:

### The objective: $\frac{1}{2}\|\mathbf{w}\|^2$

- We want to **maximize** the margin ($\frac{2}{\|\mathbf{w}\|}$)
- Maximizing $\frac{2}{\|\mathbf{w}\|}$ is the same as minimizing $\|\mathbf{w}\|$
- We minimize $\frac{1}{2}\|\mathbf{w}\|^2$ instead (same answer, easier calculus — squared removes the square root, the $\frac{1}{2}$ simplifies the derivative)

### The constraint: $y_i(\mathbf{w} \cdot \mathbf{x}_i + b) \geq 1$

- $y_i$ is the label of point $i$: either +1 or -1
- $\mathbf{w} \cdot \mathbf{x}_i + b$ is the "raw score" — how far point $i$ is from the boundary
- The product $y_i \times (\text{raw score})$ is positive when the point is on the **correct side**
- The $\geq 1$ means every point must be **at least 1 unit** away from the boundary (on the correct side)

```{r}
#| label: fig-constraint-check
#| fig-cap: "The constraint value for each point — all must be ≥ 1"

# Compute y_i * (w · x_i + b) for each point
df$y <- ifelse(df$group == "Approve", 1, -1)
df$raw_score <- as.numeric(w_svm[1] * df$x1 + w_svm[2] * df$x2 + b_svm)
df$constraint <- df$y * df$raw_score
df$is_sv <- seq_len(nrow(df)) %in% sv_indices

ggplot(df, aes(x = seq_along(constraint), y = constraint,
              color = group, shape = is_sv)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("steelblue", "coral")) +
  scale_shape_manual(values = c(16, 17), labels = c("Regular", "Support Vector")) +
  annotate("text", x = 5, y = 1.15, label = "Constraint boundary (≥ 1)",
           color = "red", size = 3.5) +
  theme_minimal(base_size = 14) +
  labs(x = "Point index", y = expression(y[i] * (w %.% x[i] + b)),
       shape = "Type", color = "Class")
```

Notice: support vectors sit **right at the constraint boundary** (value ≈ 1).
They're the tightest points. Everything else has comfortable margin.


## 7. Soft Margins: What If the Data Overlaps?

Real data is messy. Groups overlap. A perfect separating line might not exist.
**Soft margin** SVM allows some points to be on the wrong side — for a penalty.

$$\min_{\mathbf{w}, b} \quad \frac{1}{2}\|\mathbf{w}\|^2 + C \sum_{i=1}^{n} \xi_i$$

New pieces:

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $\xi_i$ (xi) | Slack variable for point $i$ | How much point $i$ violates the margin |
| $C$    | Cost parameter | How much we **penalize** violations |

- $\xi_i = 0$ → point is on the correct side, outside the margin (happy)
- $0 < \xi_i < 1$ → point is on the correct side, but inside the margin (mild violation)
- $\xi_i \geq 1$ → point is on the **wrong side** (misclassified)

```{r}
#| label: fig-cost-comparison
#| fig-cap: "Small C = wide margin (some errors OK). Large C = narrow margin (few errors tolerated)."

# Overlapping data
set.seed(99)
df2 <- data.frame(
  x1 = c(rnorm(50, 2, 1.2), rnorm(50, 4, 1.2)),
  x2 = c(rnorm(50, 2, 1.2), rnorm(50, 4, 1.2)),
  group = factor(rep(c("A", "B"), each = 50))
)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

for (C in c(0.01, 1, 100)) {
  fit <- svm(group ~ x1 + x2, data = df2, kernel = "linear", cost = C, scale = FALSE)

  grid2 <- expand.grid(
    x1 = seq(-1, 7, length.out = 150),
    x2 = seq(-1, 7, length.out = 150)
  )
  grid2$pred <- predict(fit, grid2)

  plot(df2$x1, df2$x2,
    col = ifelse(df2$group == "A", "steelblue", "coral"),
    pch = 19, cex = 1.2,
    main = paste("C =", C),
    xlab = "Feature 1", ylab = "Feature 2"
  )

  points(grid2$x1, grid2$x2,
    col = ifelse(grid2$pred == "A",
      adjustcolor("steelblue", 0.15),
      adjustcolor("coral", 0.15)),
    pch = 15, cex = 0.3
  )

  points(df2$x1[fit$index], df2$x2[fit$index],
    pch = 1, cex = 2.5, lwd = 2)

  legend("topleft", bty = "n", cex = 0.9,
    legend = c(
      paste("SVs:", length(fit$index)),
      ifelse(C < 0.1, "Wide margin", ifelse(C > 10, "Narrow margin", "Balanced"))
    )
  )
}
par(mfrow = c(1, 1))
```

**The C parameter is the most important tuning knob in SVM:**

- **Small C** (like 0.01): "I don't care much about errors" → wide margin, many support vectors, simpler model (risk of **underfitting**)
- **Large C** (like 100): "Every error is expensive" → narrow margin, few support vectors, complex model (risk of **overfitting**)
- You choose C via **cross-validation** (try different values, pick the one that performs best on held-out data)

::: {.callout-warning}
## Two C Conventions — Know Which One You're Using

**R's `ksvm()` / `e1071` / sklearn** use C as the **cost of misclassification**: Large C = narrow margin = overfit risk. This is what the demo above shows.

**Some formulations** write: $\min \sum \max(0, \ldots) + C \sum a_j^2$ where C multiplies the **regularization term** (like λ). Here: Large C = wider margin = underfit risk. This is the **opposite** convention.

**In practice:** Read the formula. If C is on the regularization term (penalty on coefficients), then large C = wider margin. If C is on the error term (cost of misclassification), then large C = narrower margin. The notation will usually tell you — don't assume one convention.
:::


## 8. The Kernel Trick: When a Line Won't Cut It

What if no straight line can separate the data?

```{r}
#| label: fig-nonlinear-data
#| fig-cap: "This data can't be separated by any straight line"

set.seed(7)
n <- 200
angle <- runif(n, 0, 2 * pi)
r_inner <- rnorm(n / 2, 1.5, 0.3)
r_outer <- rnorm(n / 2, 4, 0.5)

df3 <- data.frame(
  x1 = c(r_inner * cos(angle[1:(n/2)]), r_outer * cos(angle[(n/2+1):n])),
  x2 = c(r_inner * sin(angle[1:(n/2)]), r_outer * sin(angle[(n/2+1):n])),
  group = factor(rep(c("Inner", "Outer"), each = n / 2))
)

ggplot(df3, aes(x1, x2, color = group)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("coral", "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(x = "Feature 1", y = "Feature 2") +
  coord_equal()
```

The **kernel trick** maps data into a higher dimension where a linear separator
exists, **without actually computing the transformation**. It replaces the dot
product $\mathbf{x}_i \cdot \mathbf{x}_j$ with a kernel function
$K(\mathbf{x}_i, \mathbf{x}_j)$.

Common kernels:

| Kernel | Formula | When to use |
|:------:|:--------|:------------|
| Linear | $K = \mathbf{x}_i \cdot \mathbf{x}_j$ | Data is (roughly) linearly separable |
| Polynomial | $K = (\mathbf{x}_i \cdot \mathbf{x}_j + 1)^d$ | Interaction effects between features |
| **RBF (Gaussian)** | $K = e^{-\gamma\|\mathbf{x}_i - \mathbf{x}_j\|^2}$ | **Default choice** — handles most nonlinear patterns |

```{r}
#| label: fig-kernel-comparison
#| fig-cap: "RBF kernel draws a curved boundary where linear can't"

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

grid3 <- expand.grid(
  x1 = seq(-6, 6, length.out = 150),
  x2 = seq(-6, 6, length.out = 150)
)

# Linear kernel — fails
fit_lin <- svm(group ~ x1 + x2, data = df3, kernel = "linear", cost = 1)
grid3$pred_lin <- predict(fit_lin, grid3)

plot(df3$x1, df3$x2,
  col = ifelse(df3$group == "Inner", "coral", "steelblue"),
  pch = 19, main = "Linear Kernel (fails)", xlab = "x1", ylab = "x2")
points(grid3$x1, grid3$x2,
  col = ifelse(grid3$pred_lin == "Inner",
    adjustcolor("coral", 0.15), adjustcolor("steelblue", 0.15)),
  pch = 15, cex = 0.3)

# RBF kernel — works
fit_rbf <- svm(group ~ x1 + x2, data = df3, kernel = "radial", cost = 1, gamma = 0.5)
grid3$pred_rbf <- predict(fit_rbf, grid3)

plot(df3$x1, df3$x2,
  col = ifelse(df3$group == "Inner", "coral", "steelblue"),
  pch = 19, main = "RBF Kernel (works)", xlab = "x1", ylab = "x2")
points(grid3$x1, grid3$x2,
  col = ifelse(grid3$pred_rbf == "Inner",
    adjustcolor("coral", 0.15), adjustcolor("steelblue", 0.15)),
  pch = 15, cex = 0.3)

par(mfrow = c(1, 1))
```


## 9. Cheat Sheet: The Whole Story on One Page

```
THE SVM RECIPE
==============

1. GOAL: Find the boundary with the widest margin

2. FORMULA (soft margin):

   min  ½||w||²  +  C × Σξᵢ
        --------     --------
        "keep the    "penalize
         margin       errors"
         wide"

   subject to: yᵢ(w·xᵢ + b) ≥ 1 - ξᵢ

3. KEY PARAMETERS:
   C     → error tolerance  (tune via cross-validation)
   kernel → shape of boundary (linear, RBF, polynomial)
   γ     → RBF "reach"      (high γ = local, low γ = global)

4. SUPPORT VECTORS = points on the margin edge
   - Only these points define the boundary
   - More SVs = simpler/wider margin
   - Fewer SVs = tighter/narrower margin

5. ALWAYS SCALE YOUR FEATURES FIRST
   SVM uses distances — unscaled features with large ranges dominate
```


## 10. Check Your Understanding

::: {.callout-note}
### Test Yourself

Before moving on, try to answer these without scrolling up:

1. What does the weight vector $\mathbf{w}$ control geometrically?
2. Why do we minimize $\|\mathbf{w}\|^2$ instead of maximizing the margin directly?
3. What happens when you increase C? What about decreasing it?
4. Why are they called "support" vectors?
5. When would you use an RBF kernel instead of a linear one?
6. Why must you scale features before using SVM?
:::
