---
title: "Time Series From Scratch: Predicting What Comes Next"
subtitle: "Exponential smoothing, ARIMA, and GARCH — three tools, one goal"
engine: knitr
---

## 1. The Problem: What Will Tomorrow's Value Be?

You have data ordered in time — daily sales, monthly temperatures, hourly
stock prices. The past has a pattern. Can you **extend** that pattern into the
future?

This is **forecasting**: using the history of a value to predict where it's
going next.

```{r}
#| label: fig-time-series-intro
#| fig-cap: "Monthly airline passengers — a clear trend and seasonal pattern. What comes next?"

library(ggplot2)

# Classic airline passengers dataset
ap <- AirPassengers
df_ap <- data.frame(
  time = as.numeric(time(ap)),
  passengers = as.numeric(ap)
)

ggplot(df_ap, aes(time, passengers)) +
  geom_line(linewidth = 0.8, color = "steelblue") +
  geom_point(size = 0.8, color = "steelblue") +
  annotate("rect", xmin = 1960, xmax = 1961.5, ymin = 0, ymax = 700,
           fill = "coral", alpha = 0.15) +
  annotate("text", x = 1960.7, y = 650, label = "Predict\nthis?",
           color = "coral", fontface = "bold", size = 4) +
  theme_minimal(base_size = 14) +
  labs(x = "Year", y = "Passengers (thousands)",
       title = "Can we forecast the future from the past?")
```

Three patterns appear in time series data:

| Pattern | What it looks like | Example |
|:--------|:------------------|:--------|
| **Level** | Flat, bouncing around a constant | Daily temperature in a stable climate |
| **Trend** | Going steadily up or down | Company revenue growing year over year |
| **Seasonality** | Repeating cycle at fixed intervals | Retail sales spiking every December |

Different forecasting methods handle different combinations of these.


## 2. Exponential Smoothing: The Weighted Average Approach

### Simple Exponential Smoothing (Level Only)

The simplest idea: your forecast is a **weighted average** of the current
observation and your previous forecast.

$$S_t = \alpha \cdot x_t + (1 - \alpha) \cdot S_{t-1}$$

$$\text{Forecast: } \hat{x}_{t+1} = S_t$$

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $S_t$ | Smoothed value at time $t$ | Your best estimate of the current level |
| $x_t$ | Actual observation at time $t$ | What really happened |
| $S_{t-1}$ | Previous smoothed value | What you predicted last time |
| $\alpha$ | Smoothing parameter (0 to 1) | How much to trust the new data vs the old estimate |
| $\hat{x}_{t+1}$ | Forecast for next period | Your prediction |

**The key insight:** $\alpha$ controls how much **memory** the model has.

- **High $\alpha$ (near 1):** Mostly trusts the latest observation → reactive, follows changes quickly, but noisy
- **Low $\alpha$ (near 0):** Mostly trusts the history → smooth, stable, but slow to react

```{r}
#| label: fig-ses-alpha
#| fig-cap: "High alpha follows every wiggle. Low alpha smooths through the noise."

set.seed(42)

# Simulated data with a level shift
n <- 80
x <- c(rnorm(40, 50, 3), rnorm(40, 55, 3))
time <- 1:n

# Compute SES for different alphas
ses <- function(x, alpha) {
  s <- numeric(length(x))
  s[1] <- x[1]
  for (t in 2:length(x)) {
    s[t] <- alpha * x[t] + (1 - alpha) * s[t - 1]
  }
  s
}

alphas <- c(0.1, 0.5, 0.9)
colors <- c("steelblue", "forestgreen", "coral")

plot(time, x, pch = 19, cex = 0.6, col = "gray50",
  main = "Simple Exponential Smoothing: Effect of α",
  xlab = "Time", ylab = "Value")

for (i in seq_along(alphas)) {
  lines(time, ses(x, alphas[i]), col = colors[i], lwd = 2)
}

legend("topleft", bty = "n",
  legend = paste0("α = ", alphas),
  col = colors, lwd = 2, cex = 0.9)
```

### Why "Exponential"?

Unrolling the formula shows that older observations get **exponentially
decreasing** weights:

$$S_t = \alpha \cdot x_t + \alpha(1-\alpha) \cdot x_{t-1} + \alpha(1-\alpha)^2 \cdot x_{t-2} + \dots$$

```{r}
#| label: fig-exponential-weights
#| fig-cap: "Past observations receive exponentially decaying weights"

alpha <- 0.3
lags <- 0:15
weights <- alpha * (1 - alpha)^lags

barplot(weights, names.arg = lags,
  col = "steelblue", border = NA,
  xlab = "How many steps in the past",
  ylab = "Weight on that observation",
  main = paste0("Exponential decay of weights (α = ", alpha, ")"))
```

Recent data matters most. Distant past fades away. That's the "exponential"
in exponential smoothing.


## 3. Double Exponential Smoothing (Holt's Method): Level + Trend

Simple ES can't handle a trend — it always lags behind. Holt's method adds
a **trend component**.

$$S_t = \alpha \cdot x_t + (1 - \alpha)(S_{t-1} + T_{t-1})$$
$$T_t = \beta(S_t - S_{t-1}) + (1 - \beta) \cdot T_{t-1}$$
$$\text{Forecast: } \hat{x}_{t+h} = S_t + h \cdot T_t$$

Two new pieces:

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $T_t$ | Trend estimate at time $t$ | How much the level is changing per period |
| $\beta$ | Trend smoothing parameter (0 to 1) | How quickly to update the trend estimate |
| $h$ | Forecast horizon | How many steps ahead to predict |

```{r}
#| label: fig-holt
#| fig-cap: "Simple ES lags behind the trend. Holt's method tracks it."

set.seed(42)
n <- 60
x_trend <- 20 + 0.5 * (1:n) + rnorm(n, 0, 2)

# Simple ES
s_simple <- ses(x_trend, 0.3)

# Holt's method (manual)
alpha_h <- 0.3
beta_h <- 0.2
S <- numeric(n)
Tr <- numeric(n)
S[1] <- x_trend[1]
Tr[1] <- 0

for (t in 2:n) {
  S[t] <- alpha_h * x_trend[t] + (1 - alpha_h) * (S[t - 1] + Tr[t - 1])
  Tr[t] <- beta_h * (S[t] - S[t - 1]) + (1 - beta_h) * Tr[t - 1]
}

forecast_holt <- S + Tr  # one-step-ahead forecast

plot(1:n, x_trend, pch = 19, cex = 0.6, col = "gray50",
  main = "Trend data: Simple ES vs Holt's",
  xlab = "Time", ylab = "Value")
lines(1:n, s_simple, col = "coral", lwd = 2)
lines(1:n, forecast_holt, col = "steelblue", lwd = 2)
legend("topleft", bty = "n",
  legend = c("Simple ES (lags behind)", "Holt's (tracks trend)"),
  col = c("coral", "steelblue"), lwd = 2, cex = 0.9)
```


## 4. Triple Exponential Smoothing (Holt-Winters): Level + Trend + Seasonality

When data has a **repeating cycle** (holiday sales, summer peaks), you need a
third component.

$$S_t = \alpha \cdot \frac{x_t}{C_{t-L}} + (1 - \alpha)(S_{t-1} + T_{t-1})$$
$$T_t = \beta(S_t - S_{t-1}) + (1 - \beta) \cdot T_{t-1}$$
$$C_t = \gamma \cdot \frac{x_t}{S_t} + (1 - \gamma) \cdot C_{t-L}$$
$$\text{Forecast: } \hat{x}_{t+h} = (S_t + h \cdot T_t) \cdot C_{t+h-L}$$

New pieces:

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $C_t$ | Seasonal factor at time $t$ | How much this point in the cycle typically deviates |
| $\gamma$ | Seasonal smoothing parameter | How quickly to update the seasonal pattern |
| $L$ | Season length | Length of one full cycle (12 for monthly, 7 for daily-weekly) |

```{r}
#| label: fig-holt-winters
#| fig-cap: "Holt-Winters captures trend AND seasonality"

# Use AirPassengers — has both trend and seasonality
fit_hw <- HoltWinters(AirPassengers)

# Forecast 24 months ahead
pred_hw <- predict(fit_hw, n.ahead = 24, prediction.interval = TRUE)

par(mar = c(4, 4, 3, 1))
plot(fit_hw, pred_hw,
  main = "Holt-Winters: Airline Passengers",
  xlab = "Year", ylab = "Passengers (thousands)",
  col = "steelblue", col.predicted = "coral", lwd = 2)
legend("topleft", bty = "n",
  legend = c("Observed", "Fitted", "Forecast"),
  col = c("steelblue", "black", "coral"), lwd = 2, cex = 0.9)
```

### Which Exponential Smoothing to Use?

```
Does the data have a trend?
  |
  +-- No ──── Does it have seasonality?
  |             |
  |             +-- No ──── Simple ES (one parameter: α)
  |             +-- Yes ─── Seasonal ES
  |
  +-- Yes ─── Does it have seasonality?
                |
                +-- No ──── Holt (two parameters: α, β)
                +-- Yes ─── Holt-Winters (three parameters: α, β, γ)
```


## 5. All Parameters at a Glance

```{r}
#| label: fig-parameter-summary
#| fig-cap: "What each smoothing parameter controls"

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

set.seed(42)
n2 <- 50

# Alpha demo: level data with a jump
x_level <- c(rnorm(25, 50, 2), rnorm(25, 55, 2))
plot(1:n2, x_level, pch = 19, cex = 0.5, col = "gray50",
  main = "α controls level responsiveness",
  xlab = "Time", ylab = "Value")
lines(1:n2, ses(x_level, 0.1), col = "steelblue", lwd = 2)
lines(1:n2, ses(x_level, 0.9), col = "coral", lwd = 2)
legend("topleft", bty = "n", cex = 0.8,
  legend = c("α=0.1 (smooth)", "α=0.9 (reactive)"),
  col = c("steelblue", "coral"), lwd = 2)

# Beta demo: trending data
x_tr <- 10 + 0.8 * (1:n2) + rnorm(n2, 0, 2)
# Low beta
S1 <- Tr1 <- S2 <- Tr2 <- numeric(n2)
S1[1] <- S2[1] <- x_tr[1]; Tr1[1] <- Tr2[1] <- 0
for (t in 2:n2) {
  S1[t] <- 0.3 * x_tr[t] + 0.7 * (S1[t-1] + Tr1[t-1])
  Tr1[t] <- 0.1 * (S1[t] - S1[t-1]) + 0.9 * Tr1[t-1]
  S2[t] <- 0.3 * x_tr[t] + 0.7 * (S2[t-1] + Tr2[t-1])
  Tr2[t] <- 0.8 * (S2[t] - S2[t-1]) + 0.2 * Tr2[t-1]
}
plot(1:n2, x_tr, pch = 19, cex = 0.5, col = "gray50",
  main = "β controls trend responsiveness",
  xlab = "Time", ylab = "Value")
lines(1:n2, S1 + Tr1, col = "steelblue", lwd = 2)
lines(1:n2, S2 + Tr2, col = "coral", lwd = 2)
legend("topleft", bty = "n", cex = 0.8,
  legend = c("β=0.1 (stable trend)", "β=0.8 (reactive trend)"),
  col = c("steelblue", "coral"), lwd = 2)

# Gamma demo: seasonal data
x_seas <- 50 + rep(c(0, 5, 10, 5, 0, -5, -10, -5), 6) + rnorm(48, 0, 1.5)
ts_seas <- ts(x_seas, frequency = 8)
fit_lo <- HoltWinters(ts_seas, gamma = 0.1)
fit_hi <- HoltWinters(ts_seas, gamma = 0.9)
plot(1:48, x_seas, pch = 19, cex = 0.5, col = "gray50",
  main = "γ controls seasonal responsiveness",
  xlab = "Time", ylab = "Value")
lines(9:48, fit_lo$fitted[, "xhat"], col = "steelblue", lwd = 2)
lines(9:48, fit_hi$fitted[, "xhat"], col = "coral", lwd = 2)
legend("topleft", bty = "n", cex = 0.8,
  legend = c("γ=0.1 (stable season)", "γ=0.9 (reactive season)"),
  col = c("steelblue", "coral"), lwd = 2)

par(mfrow = c(1, 1))
```

**How to choose α, β, γ?** Cross-validation (Module 3) — try different values,
pick the ones that minimize forecast error on held-out data. R's `HoltWinters()`
does this automatically.


## 6. ARIMA: A Different Approach

Exponential smoothing is intuitive but limited in the patterns it can express.
**ARIMA** is more flexible — it combines three ideas into one framework.

### ARIMA(p, d, q) — What the Letters Mean

| Letter | Name | What it does |
|:------:|:-----|:-------------|
| **p** | AutoRegressive order | Use the last $p$ values to predict the next |
| **d** | Differencing order | How many times to difference the data to remove trend |
| **q** | Moving Average order | Use the last $q$ forecast errors to correct predictions |

### AR: Autoregressive (the "p" part)

Tomorrow's value depends on today's value (and maybe yesterday's):

$$x_t = c + \phi_1 x_{t-1} + \phi_2 x_{t-2} + \dots + \phi_p x_{t-p} + \epsilon_t$$

Like regression, but instead of predicting from other variables, you predict
from **your own past values**.

```{r}
#| label: fig-ar-example
#| fig-cap: "AR(1): each value depends on the previous one, plus noise"

set.seed(42)
n3 <- 100

# AR(1) process
ar_data <- numeric(n3)
ar_data[1] <- 0
phi <- 0.8

for (t in 2:n3) {
  ar_data[t] <- phi * ar_data[t - 1] + rnorm(1, 0, 1)
}

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(1:n3, ar_data, type = "l", col = "steelblue", lwd = 1.5,
  main = paste0("AR(1) with φ = ", phi),
  xlab = "Time", ylab = "Value")

# Show the relationship: x_t vs x_{t-1}
plot(ar_data[1:(n3-1)], ar_data[2:n3],
  pch = 19, cex = 0.8, col = "steelblue",
  main = "x_t vs x_{t-1}",
  xlab = expression(x[t-1]), ylab = expression(x[t]))
abline(lm(ar_data[2:n3] ~ ar_data[1:(n3-1)]), col = "coral", lwd = 2)
legend("topleft", bty = "n", cex = 0.8,
  legend = paste0("Slope ≈ φ = ", round(cor(ar_data[1:(n3-1)], ar_data[2:n3]), 2)))

par(mfrow = c(1, 1))
```

### I: Integrated / Differencing (the "d" part)

ARIMA requires **stationary** data (no trend, constant variance). If your data
has a trend, **differencing** removes it:

$$x_t' = x_t - x_{t-1}$$

```{r}
#| label: fig-differencing
#| fig-cap: "Differencing removes trend, making the data stationary"

# Trending data
x_trended <- 20 + 0.3 * (1:n3) + cumsum(rnorm(n3, 0, 1))
x_diffed <- diff(x_trended)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(1:n3, x_trended, type = "l", col = "steelblue", lwd = 1.5,
  main = "Original (has trend)",
  xlab = "Time", ylab = "Value")

plot(1:(n3-1), x_diffed, type = "l", col = "steelblue", lwd = 1.5,
  main = "After differencing (d=1)",
  xlab = "Time", ylab = "Differenced Value")
abline(h = 0, lty = 2, col = "gray50")

par(mfrow = c(1, 1))
```

- **d = 0**: Data is already stationary (no trend)
- **d = 1**: Difference once (removes linear trend)
- **d = 2**: Difference twice (removes quadratic trend) — rarely needed

### MA: Moving Average (the "q" part)

Tomorrow's value depends on today's **forecast error**:

$$x_t = c + \epsilon_t + \theta_1 \epsilon_{t-1} + \theta_2 \epsilon_{t-2} + \dots + \theta_q \epsilon_{t-q}$$

This corrects systematic patterns in the errors. If you keep over-predicting,
the MA term adjusts.

### Putting It Together

ARIMA(1,1,1) means: difference once, then model with 1 AR term and 1 MA term.

```{r}
#| label: fig-arima-fit
#| fig-cap: "ARIMA automatically selects p, d, q and forecasts ahead"

# Fit ARIMA to airline passengers
fit_arima <- arima(AirPassengers, order = c(1, 1, 1),
                   seasonal = list(order = c(1, 1, 1), period = 12))

# Forecast
pred_arima <- predict(fit_arima, n.ahead = 24)

par(mar = c(4, 4, 3, 1))
ts.plot(AirPassengers, pred_arima$pred,
  col = c("steelblue", "coral"), lwd = c(1.5, 2),
  main = "ARIMA Forecast: Airline Passengers",
  ylab = "Passengers (thousands)")

# Confidence interval
upper <- pred_arima$pred + 1.96 * pred_arima$se
lower <- pred_arima$pred - 1.96 * pred_arima$se
time_pred <- time(pred_arima$pred)
polygon(c(time_pred, rev(time_pred)), c(upper, rev(lower)),
  col = adjustcolor("coral", 0.2), border = NA)

legend("topleft", bty = "n",
  legend = c("Observed", "Forecast", "95% CI"),
  col = c("steelblue", "coral", adjustcolor("coral", 0.3)),
  lwd = c(2, 2, 8), cex = 0.9)
```


## 7. ARIMA vs Exponential Smoothing: When to Use Which

| | Exponential Smoothing | ARIMA |
|:--|:---------------------|:------|
| **Simplicity** | Simpler, fewer parameters | More complex, flexible |
| **Data needed** | Works with shorter series | Needs ~40+ observations |
| **Noisy/volatile data** | More robust | Can struggle with outliers |
| **Complex patterns** | Limited to level/trend/season | Can model many patterns |
| **Interpretability** | Parameters have clear meaning | p, d, q less intuitive |
| **Special case** | Simple ES | = ARIMA(0,1,1) |

::: {.callout-tip}
### Practical Rule of Thumb

- Short, noisy, volatile data → **exponential smoothing**
- Long, stable data with many observations → **ARIMA**
- When in doubt → try both, pick the one with lower cross-validated error
:::


## 8. GARCH: Forecasting How MUCH Things Will Vary

Everything above forecasts the **value** — what will the temperature be
tomorrow? GARCH answers a different question: **how volatile will it be?**

Think of stock markets: some weeks are calm, some are wild. The actual price
might be unpredictable, but the **amount of volatility** has patterns.

$$\sigma_t^2 = \omega + \alpha_1 \epsilon_{t-1}^2 + \beta_1 \sigma_{t-1}^2$$

| Symbol | What it is | Plain English |
|:------:|:-----------|:-------------|
| $\sigma_t^2$ | Predicted variance at time $t$ | How much volatility we expect |
| $\omega$ | Baseline variance | The minimum level of volatility |
| $\epsilon_{t-1}^2$ | Previous squared error | How big yesterday's surprise was |
| $\alpha_1$ | Shock sensitivity | How much a single surprise increases future volatility |
| $\beta_1$ | Persistence | How long volatility stays elevated after a shock |
| $\sigma_{t-1}^2$ | Previous variance | Yesterday's volatility forecast |

**Key insight:** $\alpha$ captures **shocks** (a big surprise today → more
volatility tomorrow). $\beta$ captures **persistence** (volatility tends to
stay high once it's high).

```{r}
#| label: fig-garch-concept
#| fig-cap: "Volatility clusters: calm periods and turbulent periods alternate"

set.seed(42)

# Simulate GARCH-like data
n4 <- 300
returns <- numeric(n4)
vol <- numeric(n4)
omega <- 0.1
alpha_g <- 0.15
beta_g <- 0.8

vol[1] <- 1
returns[1] <- rnorm(1, 0, sqrt(vol[1]))

for (t in 2:n4) {
  vol[t] <- omega + alpha_g * returns[t-1]^2 + beta_g * vol[t-1]
  returns[t] <- rnorm(1, 0, sqrt(vol[t]))
}

par(mfrow = c(2, 1), mar = c(3, 4, 3, 1))

plot(1:n4, returns, type = "l", col = "steelblue", lwd = 0.8,
  main = "Simulated Returns (notice: calm and turbulent clusters)",
  ylab = "Return", xlab = "")

plot(1:n4, sqrt(vol), type = "l", col = "coral", lwd = 1.5,
  main = "True Volatility (what GARCH estimates)",
  ylab = "Volatility (σ)", xlab = "Time")

par(mfrow = c(1, 1))
```

### When to Use GARCH

A common practice check asks whether you know the difference:

| Question asks about... | Use |
|:-----------------------|:----|
| "Forecast the **value**" | Exponential smoothing or ARIMA |
| "Forecast **volatility/risk/variance**" | GARCH |
| "Predict how much values will **fluctuate**" | GARCH |
| "Build **confidence intervals** that change over time" | GARCH |
| "**Portfolio risk**" or "**Value at Risk**" | GARCH |


## 9. The Complete Decision Tree

```
What are you forecasting?
  |
  +-- The VALUE itself ──── Is there a trend? Seasonality?
  |                           |
  |                           +-- Level only ────────── Simple ES
  |                           +-- Level + trend ─────── Double ES (Holt)
  |                           +-- Level + trend + season ── Triple ES (Holt-Winters)
  |                           +-- Complex pattern, lots of data ── ARIMA
  |
  +-- The VOLATILITY/VARIANCE ── GARCH
  |
  +-- Has something CHANGED? ── CUSUM (Module 6)
```


## 10. Cheat Sheet: The Whole Story on One Page

```
TIME SERIES FORECASTING RECIPE
================================

EXPONENTIAL SMOOTHING
---------------------
Simple:   Sₜ = α·xₜ + (1-α)·Sₜ₋₁                    [level only]
Holt:     + Tₜ = β·(Sₜ - Sₜ₋₁) + (1-β)·Tₜ₋₁         [+ trend]
H-W:      + Cₜ = γ·(xₜ/Sₜ) + (1-γ)·Cₜ₋ₗ              [+ seasonality]

Parameters:
  α → level responsiveness    (high = reactive, low = smooth)
  β → trend responsiveness
  γ → seasonal responsiveness
  All chosen via cross-validation

"Exponential" = past data gets exponentially decaying weights

ARIMA(p, d, q)
--------------
p = autoregressive terms  (predict from past values)
d = differencing           (remove trend)
q = moving average terms   (correct from past errors)

ARIMA(0,1,1) = simple exponential smoothing
Use when: stable data, 40+ observations, complex patterns

GARCH(p, q)
-----------
σₜ² = ω + α·εₜ₋₁² + β·σₜ₋₁²

Forecasts VARIANCE, not value.
Use when: "volatility", "risk", "how much will values vary"

PRACTICAL DECISION:
  "Forecast the value" → ES or ARIMA
  "Forecast the variance/volatility" → GARCH
  "Detect a change" → CUSUM
```


## 11. Check Your Understanding

::: {.callout-note}
### Test Yourself

Before moving on, try to answer these without scrolling up:

1. What does $\alpha$ control in exponential smoothing? What happens when
   you increase it?
2. Why is it called "exponential" smoothing?
3. When do you need Holt's method instead of simple ES?
4. What does the "d" in ARIMA stand for, and why is it needed?
5. What is the key difference between ARIMA and exponential smoothing?
   When would you prefer each?
6. GARCH forecasts _______, not _______. (Fill in the blanks.)
7. A manager asks you to "predict next quarter's sales." Which method?
   Now they ask "predict how volatile next quarter's sales will be."
   Which method?
8. What is ARIMA(0,1,1) equivalent to?
9. Data has 20 observations, lots of noise, and a seasonal pattern.
   Exponential smoothing or ARIMA? Why?
:::
